Acquisition of auditory discrimination mediated by different processes through two distinct circuits linked to the lateral striatum

Version 1

Abstract

Abstract

The striatum, the central hub of cortico-basal ganglia loops, contains functionally heterogeneous subregions distinguished by the topographic patterns of structural connectivity. These subregions mediate various processes of procedural learning. However, it remains unclear when and how striatal subregions engage in the acquisition of sensory stimulus-based decision-making. A neuroimaging of regional brain activity shows that the anterior dorsolateral striatum (aDLS) and posterior ventrolateral striatum (pVLS) are activated in a different temporal pattern during the acquisition phase of auditory discrimination. Chronic and transient pharmacologic manipulations show that the aDLS promotes the behavioral strategy driven by the stimulus-response association while suppressing that by the response-outcome association, and that the pVLS contributes to forming and maintaining the stimulus-response strategy. Electrophysiological recording indicates that subpopulations of aDLS neurons predominantly represent the outcome of specific behaviors at the initial period of discrimination learning, and that pVLS subpopulations encode the beginning and ending of each behavior according to the progress of learning. In addition, other subpopulations of striatal neurons indicate sustained activation after obtaining reward with distinct patterns reflecting the stimulus-response associations. Our findings demonstrate that aDLS and pVLS neurons integrate the new learning of auditory discrimination in spatiotemporally and functionally different manners.

Introduction

The neural network connecting the cerebral cortex, basal ganglia, and thalamus forms functionally heterogeneous loops distinguished by the topographic patterns of connectivity1,2, and is implicated in various learning processes by the integration of cognitive, sensorimotor, and reward information3,4. The striatum is a key node in the cortico-basal ganglia loops, and its segregated subregions are considered to mediate procedural learning, such as motor skills and instrumental behavior, through parallel processing via different loops5–7. Previous studies in humans and non-human primates have reported that the neural activity within the striatum shifts from the associative subregions (caudate and anterior putamen) to the sensorimotor subregions (posterior putamen) as motor learning progresses8–11. In rodents, an instrumental behavior initiates as goal-directed actions and then transits to stimulus-response habits12,13, requiring the dorsomedial striatum (DMS; a homologue to the caudate in primates) and the dorsolateral striatum (DLS; a homologue to the putamen in primates), respectively14,15. Based on these results, a model of procedural learning has been proposed, in which the functional dominance changes from the associative to sensorimotor subregions in the striatum during the learning phases16.

In contrast, deviations from the prior model are evident in other behavioral tasks on decision-making based on external sensory cues. For example, the neural activity in the caudate in humans is associated with learning of a visuomotor association task and the execution of learned behavior17,18. The tail of the caudate in non-human primates contributes to the automatic, habitual behavior in decision-making task using complex visual stimuli19. In rodents, the DLS is involved in the acquisiton and execution of the discrimination task with visual and auditory cues, wheras the DMS madiates the performance of learned behavior20,21. Moreover, other striatal subregions, such as the posterior striatum and ventrolateral striatum in rodents are also engaged in the auditory discrimination22–24. Based on these results, a hypothesis has arisen that the acquisition of sensory cue-based decision-making requires wide-range spatiotemporal processing within the striatum. However, the neural mechanism by which the functional circuits linked to striatal subregions change throughout learning processes is not yet fully understood.

To investigate a large-scale striatal reorganization in accordance with the acquisition of external cue-dependent decisions, we conducted a two-alternative auditory discrimination task25. Using a small-animal neuroimaging technique, we found that the anterior dorsolateral striatum (aDLS) and posterior ventrolateral striatum (pVLS) are activated to the highest level at the middle and late stages during the acquisition phase of auditory discrimination, respectively. Then, chronic pharmacologic manipulations confirmed that the aDLS and pVLS are necessary for discrimination learning, and the transient manipulations also ascertained the function of these subregions in the learning at the corresponding stages. The resultant effects of the manipulations on the behavioral strategy showed that the aDLS promotes the strategy based on the stimulus-response association, while suppressing that based on response-outcome associations. Additionally, the pVLS engages in the formation and maintenance of the stimulus-response association strategy. Electrophysiological recording of striatal neurons indicated that subpopulations of aDLS neurons mainly represent the outcome of specific behaviors at the initial period of discrimination learning, whereas pVLS subpopulations encode the beginning and ending of each behavior in association with the progress of the learning. In addition, other subpopulations of aDLS and pVLS neurons show sustained activation after obtaining reward with distinct patterns of the combination between the stimulus and response. These findings reveal that the aDLS and pVLS regulate the acquisition of auditory discrimination through distinct spatiotemporal and functional manners, challenging the prior model which proposed a transition from the associative to sensorimotor striatal subregions during learning.

Results

Leaning processes of auditory discrimination

We employed a two-alternative auditory discrimination task, in which animals are required to make new associations between a tone instruction cue (2 or 10 kHz) and a response (pressing the left or right lever)25, setting the interval between the stimulus presentation and lever insertion to 3 s (Figure 1A). The success rate gradually increased, reaching a plateau around Day 13, which persisted through Day 24 (Figure 1B). When the success rates across Days 2, 6, 10, 13 and 24 were compared, significant differences were observed between Days 2 and 6, Days 6 and 10, as well as Days 10 and 13, showing no significant difference between Days 13 and 24 (Figure 1C). In contrast, the response time and omission rate did not vary among the training days (Figures 1D and 1E), and the variation of response bias decreased along with the progress of auditory discrimination (Figure 1F). Based on changes observed in the success rate, we categorized the learning processes into two phases: the acquisition phase, which lasted until Day 13; and the learned phase, which began after Day 13.

Behavioral performance of two-alternative auditory discrimination learning

(A) Schematic diagram of the auditory discrimination task. Each trial started by presentation of a tone instruction cue with the high (10 kHz) or low (2 kHz) frequency. Three seconds later, the room light was illuminated, and two retractable levers were inserted at the same time. The rats were required to press the right and left levers in response to the high and low tones, respectively. (B) Learning curve of the auditory discrimination in intact rats (n = 14). Inset indicates the cumulative curve of number achieved the success rate of more than 90%. (C) Success rate at Days 2, 6, 10, 13, and 24 (one-way repeated ANOVA, F[1.897,24.659] = 106.706, p = 1.1 × 10-12, post hoc Bonferroni test, Day 2 vs. Day 6, p = 0.001, Day 6 vs. Day 10, p = 3.6 × 10-4, Day 10 vs. Day 13, p = 0.040, and Day 13 vs. Day 24, p = 0.080). (D) Response time (one-way repeated ANOVA, F[2.203,28.642] = 2.386, p = 0.105). (E) Omission rate (F[4,52] = 0.699, p = 0.596). (F) Response bias (one-way repeated ANOVA, F[1.129,14.674] = 0.250, p = 0.653). Data are indicated as the mean ± s.e.m., and individual data are overlaid. *p < 0.05, **p < 0.01, and ***p < 0.001.

Distinct brain activity patterns in striatal subregions during learning processes

To investigate dynamic changes in regional brain activity in the entire striatum during auditory discrimination in the same animals, we conducted the positron emission tomography (PET) for small animals with 2-deoxy-2-[18F]fluoro-D-glucose (18F-FDG), which measures cerebral glucose metabolism correlated to brain activity26–28 (Figure 2A). We performed 18F-FDG-PET scanning on Day 4 of the single lever press task without sound-based decision as a control condition and during a series of the trainings including the acquisition (Days 2, 6, and 10) and learned (Day 24) phases (Figure 2B). The average number of lever presses in the single lever press task exceeded 80 times (81.6 ± 1.3, mean ± s.e.m.) on Day 2 and maintained that level until Day 4 (Figure 2C). In the following auditory discrimination task, the success rate gradually increased along with the training days (Figure 2D), showing a significant difference among sessions in which we performed the 18F-FDG-PET scanning (Figure 2E). The response time and omission rate were consistent among the days, and the variation of response bias became smaller as the task progresses (Figure 2E). These performances were similar to those in the intact rats (compared to Figures 1C–1F), suggesting that the procedure for 18F-FDG-PET scans does not affect the acquisition of discrimination.

Dissociable brain activity patterns during auditory discrimination learning among three striatal subregions

(A) Schematic illustration of an awake rat received intravenously 18F-FDG injection through an indwelling catheter attached to the tail. The rat was conducted for the behavioral experiment and then used for microPET imaging. (B) Schedule for the behavioral training and 18F-FDG-PET scan. After 18F-FDG injection, rats (n = 14 rats) were subjected to the behavioral experiment (30 min) and returned to the home cage (15 min). Ten min before the PET scan started, the rats were anesthetized and then the scan was started (30 min). (C and D) Learning curves of the single lever press task (C) and auditory discrimination task (D). Arrowheads indicate the PET scan days. (E) Behavioral performance on the scan days during the auditory discrimination. Success rate (one-way repeated ANOVA, F[3,39] = 125.012, p = 4.8 × 10-20, post hoc Bonferroni test; Day 2 vs. Day 6, p = 5.0 × 10-7; Day 6 vs. Day 10, p = 0.001; Day 10 vs. Day 24, p = 0.003), response time (one-way repeated ANOVA, F[3,39] = 0.156, p = 0.926), omission rate (one-way repeated ANOVA, F[3,39] = 1.559, p = 0.215), and response bias (one-way repeated ANOVA, F[1.779,23.130] = 17.734, p = 3.6 × 10-5) are shown. (F) Representative images of coronal sections that compared the brain activity in the single lever press task with the activity on Day 2, 6, 10, or 24 in the discrimination task. Left panels show schematic illustrations of striatal subregions. (G) Horizontal (top) and coronal (bottom) images of striatal activation areas shown on Day 10 vs. single lever in (F). (H) Representative images of coronal section that compared the brain activity on Day 2 with that on Day 6, 10, or 24 in the discrimination task. Color bars indicate the T-values, and a value of 3.8 was used as the threshold corresponding to the uncorrected threshold (p < 0.001). (I) Schematic pictures showing voxel of interests for the aDLS, pVLS, and DMS (green, left hemisphere; and purple, right hemisphere). (J-L) Regional 18F-FDG uptakes in the aDLS (J, one-way repeated ANOVA, left aDLS, F[4,52] = 10.322, p = 3.0 × 10-6, post hoc Bonferroni test; single lever vs. Day 6, p = 0.016; Day 2 vs. Day 6, p = 6.5 × 10-5; Day 2 vs. Day 10, p = 0.019; Day 2 vs. Day 24, p = 0.017; Day 6 vs. Day 24, p = 0.041; right aDLS, F[4,52] = 7.462, p = 7.9 × 10-5, post hoc Bonferroni test, Day 2 vs. Day 6, p = 2.9 × 10-4; Day 2 vs. Day 10, p = 0.036; Day 6 vs. Day 24, p = 0.005), pVLS (K, one-way repeated ANOVA, left pVLS, F[2.368,30.784] = 4.152, p = 0.020, post hoc Bonferroni test; single lever vs. Day 10, p = 1.8 × 10-5; right pVLS, F[4,52] = 5.995, p = 4.8 × 10-4, post hoc Bonferroni test, single lever vs. Day 10, p = 0.001; Day 2 vs. Day 10, p = 0.011; Day 6 vs. Day 10, p = 0.032), and DMS (L, one-way repeated ANOVA, left DMS, F[4,52] = 12.836, p = 2.4 × 10-7, post hoc Bonferroni test; Day 2 vs. Day 24, p = 1.8 × 10-5; Day 6 vs. Day 24, p = 4.8 × 10-4; Day 10 vs. Day 24, p = 5.5 × 10-5; right DMS, F[4,52] = 10.717, p = 2.0 × 10-6, post hoc Bonferroni test; single lever vs. Day 24, p = 0.036; Day 2 vs. Day 24, p = 0.011; Day 6 vs. Day 24, p = 2.6 × 10-4; Day 10 vs. Day 24, p = 0.011). Arrowheads indicate the day with the most activations throughout the learning process. Data are indicated as the mean ± s.e.m., and individual data are overlaid. The anteroposterior coordinates from bregma (mm) are shown (F–I). Scale bar; 2 mm (I). *p < 0.05, **p < 0.01, and ***p < 0.001.

The regional brain activity related to the behavioral task was analyzed with a voxel-based statistical parametric analysis26, in which the 18F-FDG uptake on either day of the discrimination task was compared with that on the single lever task. The task-related brain activity in the whole brain is summarized in Table S1. In the present study, we focused on the task-related activity in striatal subregions. We found significant increases of the activity in the unilateral aDLS and the bilateral pVLS on Days 6 and 10 during the acquisition phase (Figures 2F and 2G). However, no significant change was observed in the DMS during the acquisition phase except a unilateral decrease on Day 24 (Figure 2F). Next, to evaluate the activity related to the progress of learning, 18F-FDG uptake on Days 6, 10, and 24 was compared with that on Day 2. The learning-dependent brain activity was significantly increased in the bilateral aDLS and the unilateral pVLS on Days 6 and 10, respectively (Figure 2H), whereas the activity decreased in the bilateral DMS on Day 24 (Figure 2H). This decrease was also detected when 18F-FDG uptakes were compared between Days 10 and 24 (Figure 2H), suggesting the down-regulation of the DMS activity in the learned phase.

To quantitatively validate changes in the brain activity in striatal subregions, we analyzed the amount of 18F-FDG uptake in the voxel of interest of these subregions (Figure 2I). The 18F-FDG uptake in the bilateral aDLS reached a peak on Day 6, and it decreased during the subsequent days (Figure 2J), whereas the uptake in the bilateral pVLS gradually increased along with the progress of learning, showing the maximal value on Day 10 (Figure 2K). The uptake in the DMS did not alter during the acquisition phase, and it decreased on Day 24 (Figure 2L). These results suggest that the aDLS and pVLS contribute to learning processes in a different temporal pattern during the acquisition phase, whereas the DMS may be engaged to the execution of the behavior in the learned phase.

Excitotoxic lesion of the aDLS and pVLS, but not the DMS, disrupts the acquisition of auditory discrimination

To study whether the changed activity of striatal subregions is involved in learning processes, rats received a bilateral injection of solution containing ibotenic acid (IBO; 8 mg/mL, 0.3 μL/site) or phosphate buffered saline (PBS) into the subregions, and we conducted for the behavioral study. The range of IBO lesion in the subregions was analyzed by immunostaining with a neuronal marker NeuN after the behavioral tests (Figures 3A–3C). The number of single lever presses increased similarly along with the training in both the PBS- and IBO-injected groups into the aDLS, pVLS, or DMS along with the training (Figures 3D–3F). In the discrimination learning, increases of the success rate were impaired in both IBO-injected aDLS and pVLS groups, compared to the corresponding PBS-injected groups (Figures 3G and 3H). The response time in the aDLS-lesioned group did not alter during the discrimination training, whereas that in the pVLS-lesioned group was continuously prolonged (Figure. S1A and S1B), and the omission rate and response bias were similar between the lesioned and the corresponding control groups (Figures S1D/S1G for aDLS lesion and Figures S1E/S1H for pVLS lesion). In contrast, the success rate in the discrimination task was comparable between the PBS- and IBO-injected groups into the DMS (Figure 3I). The DMS lesion lengthened the response time through the learning (Figure S1C), whereas it did not alter the omission rate and response bias (Figures S1F and S1I). These results indicate that the aDLS and pVLS are required for the acquisition of auditory discrimination, whereas the DMS contributes to the execution but not learning of the discriminative behavior, being consistent with previous reports29,30.

Impacts of excitotoxic lesion of striatal subregions on the acquisition of auditory discrimination

Rats were given intracranial injection of PBS or IBO solution into the aDLS (n = 8 for each injection), pVLS (n = 8 for each injection), or DMS (n = 7 for PBS injection, and n = 6 for IBO injection) before start of the single lever press task. (A-C) Representative images of NeuN immunostaining and individual schematic illustrations showing lesioned area in the aDLS (A), pVLS (B), or DMS (C). Dotted lines in the images indicate the range of the lesioned area. ac, anterior commissure; cc, corpus callosum; and LV, lateral ventricle. (D-F) Learning curves during the single lever press task in the groups injected into the aDLS (D, two-way repeated ANOVA, group, F[1,14] = 1.275, p = 0.278, day, F[1.512,21.171] = 26.281, p = 7.0 × 10-6, group × day, F[1.512,21.171] = 0.834, p = 0.418), pVLS (E, two-way repeated ANOVA, group, F[1,14] = 2.542, p = 0.133, day, F[1.547,21.658] = 40.183, p = 2.2 × 10-7, group × day, F[1.547,21.658] = 0.953, p = 0.380), or DMS (F, two-way repeated ANOVA, group, F[1,11] = 0.025, p = 0.876, day, F[1.205,13.252] = 14.818, p = 0.001, group × day, F[1.205,13.252] = 0.610, p = 0.478). (G-I) Learning curves during the auditory discrimination task in the groups injected into the aDLS (G, two-way repeated ANOVA, group, F[1,14] = 11.578, p = 0.004, day, F[2.805,39.274] = 60.733, p = 1.8 × 10-14, group × day, F[2.805,39.274] = 3.863, p = 0.018), pVLS (H, two-way repeated ANOVA, group, F[1,14] = 7.221, p = 0.018, day, F[3.646,51.041] = 51.794, p = 9.2 × 10-17, group × day, F[3.646,51.041] = 1.612, p = 0.190), or DMS (I, two-way repeated ANOVA, group, F[1,11] = 0.469, p = 0.507, day, F[2.677,29.452] = 48.860, p = 3.9 × 10-11, group × day, F[2.677,29.452] = 1.094, p = 0.362). (J-L) Changes in the proportion of the WSW or LSL strategy through the discrimination learning in the groups treated into the aDLS (J, two-way repeated ANOVA, group, F[1,14] = 13.451, p = 0.003, day, F[3.048,42.673] = 54.777, p = 9.1 × 10-15, group × day, F[3.048,42.673] = 3.288, p = 0.029 for the WSW; group, F[1,14] = 5.039, p = 0.041, day, F[3.866,54.119] = 14.578, p = 5.2 × 10-8, group × day, F[3.866,54.119] = 2.471, p = 0.057 for the LSL), pVLS (K, two-way repeated ANOVA, group, F[1,14] = 9.251, p = 0.009, day, F[3.206,44.881] = 47.168, p = 2.7 × 10-14, group × day, F[3.206,44.881] = 1.862, p = 0.146 for the WSW; group, F[1,14] = 0.762, p = 0.397, day, F[4.773,66.815] = 24.018, p = 1.9 × 10-13, group × day, F[4.773,66.815] = 1.338, p = 0.260 for the LSL), or DMS (L, two-way repeated ANOVA, group, F[1,11] = 0.703, p = 0.420, day, F[3.052,33.568] = 48.424, p = 2.3 × 10-12, group × day, F[3.052,33.568] = 0.846, p = 0.480 for the WSW; group, F[1,11] = 0.002, p = 0.965, day, F[3.442,37.866] = 13.787, p = 1.0 × 10-6, group × day, F[3.442, 37.866] = 0.581, p = 0.654 for the LSL). Data are indicated as the mean ± s.e.m., and individual data are overlaid (except for panels J–L). The anteroposterior coordinates from bregma (mm) are shown (A–C). Scale bars; 2 mm (A–C). *p < 0.05 and **p < 0.01.

Previous studies have reported that the dorsal striatum is involved in both behavioral strategies based on the response-outcome association and the stimulus-response association during stimulus-response learning12,13. To assess the impact of striatal lesions on the behavioral strategies, we analyzed the proportion of response attributed to two representative strategies in all responses of each session. One is the “win-shift-win (WSW)” strategy, in which, after a correct response in the previous trial, rats press the opposite lever in the current trial in response to a shift of the instruction cue, resulting in the correct response. The WSW is considered to reflect the behavioral strategy based on the stimulus-response association31,32. Another strategy is the “lose-shift-lose (LSL)” strategy, in which, after an error response in the previous trial rats press the opposite lever in the current trial despite a shift of the instruction cue, leading to another error response. The LSL is considered to appear as a consequence of the behavioral strategy based on the response-outcome association31. In the PBS-injected group into either of three striatal subregions, the proportion of WSW and LSL strategies increased and decreased, respectively, during the acquisition phase (Figures 3J–3L). In the IBO-lesioned group into the aDLS, the increase in the WSW proportion was markedly impaired as compared to the corresponding control group, and the decrease in the LSL proportion was delayed compared to the same group (Figure 3J). In the pVLS-lesioned group, the increasing WSW proportion was also impaired as compared to the corresponding controls, although the change in the LSL proportion was comparable to the control response (Figure 3K). In contrast, in the DMS lesion group both WSW and LSL proportions did not show significant differences from the corresponding values in their control group (Figure 3L). These results indicate that the aDLS promotes the WSW strategy through stimulus-response association and suppresses the LSL strategy rooted in response-outcome association, and that the pVLS selectively facilitates the WSW strategy acquisition. Thus, these two subregions not only play roles in the acquisition of auditory discrimination but also control distinct behavioral strategies during the learning processes.

After the completion of discrimination training, we tested locomotion and feeding behavior of rats with aDLS and pVLS lesions. There was no difference in locomotor activity in the open field and the amount of food intake in both lesioned groups (Figures S1J-S1M), suggesting that learning deficits following excitotoxic lesions of the striatal subregions cannot be attributable to abnormalities in these general motor behaviors.

Inhibiting aDLS and pVLS activity impairs discriminative behavior at different stages of the acquisition phase

Our 18F-FDG-PET imaging study indicated that the brain activity in the aDLS and pVLS was increased in a different temporal pattern during the acquisition phase of auditory discrimination. To examine whether neuronal activity in striatal subregions is linked with the processes at different stages, we temporarily inhibited the activity in each subregion by the treatment with a GABAA receptor agonist muscimol (MUS). In a pilot study, we tested the effect of the bilateral injection of several doses of MUS (0.1, 0.2, and 0.5 μg/μL) into the aDLS or pVLS on the single lever press behavior and confirmed that the administration of the lowest dose of MUS did not affect the frequency of single lever presses and the quantity of food intake (Figures S2A–S2D). In addition, we divided the acquisition phase into three stages based on the success rate that included the early (< 60%), middle (60–80%), and late (80–100%) stages. To achieve the stage-specific inactivation of the aDLS or pVLS neurons, rats were given bilateral injections of MUS or saline (SAL) at three different timings during a series of the training. The first injection was conducted on Day 2 to target the early stage. The second and third injections were conducted on the day after the success rate had reached 60% and 80% for the first time through the training, respectively, to target the middle or late stage. After the injections, the rats were used for the behavioral test. Placement locations of the injection cannula into the striatal subregions were assessed by cresyl violet staining after the behavioral tests (Figures 4A and 4B). The success rates on the day before the injection (day N-1) at the early, middle, and late stages were similar between the two groups that received the bilateral injections of MUS or SAL (Figures S2E and S2F). Rats were then administered injections into the aDLS on the next day (Day N), and the success rate in the MUS-injected groups significantly decreased at the middle stage compared to the corresponding SAL-injected groups, but not at the early and late stages (Figure 4C, left). In contrast, the success rate in the MUS-injected groups into the pVLS significantly decreased at the late stage, but not at the early and middle stages, compared with the respective SAL-injected groups (Figure 4D, left). To normalize the variation of learning speed in the individual animals, we calculated the value by subtracting the success rate on Day N-1 from that on Day N in the individual animals. There were significant reductions in the subtracted rate at the middle stage in the MUS-injected group into the aDLS (Figure 4C, right) and at the late stage in the MUS-injected group into the pVLS (Figure 4D, right). These results indicate that the aDLS and pVLS mainly function at the middle and late stages, respectively, in the acquisition phase of auditory discrimination, supporting the results obtained from our imaging study.

Influence of transient inhibition of striatal subregions at different timings on the performance of auditory discrimination

Rats received intracranial injection of SAL or MUS solution into the aDLS (n = 8 for each injection) or pVLS (n = 6 for each injection). (A and B) Representative images of cresyl violet staining and schematic illustrations showing placement sites of the tip of guide cannula in the aDLS (A) or pVLS (B). Dotted lines in the images indicate the position of the cannula placement. Ac, anterior commissure; cc, corpus callosum; and LV, lateral ventricle. (C and B) Effects of transient striatal inhibition on the performance. For aDLS inhibition (C), success rate (left, two-way repeated ANOVA, stage, F[2,28] = 28.708, p = 1.7 × 10-7, group, F[1,14] = 8.840, p = 0.010, stage × group, F[2,28] = 0.757, p = 0.478; unpaired Student’s t-test, early, t[14] = 0.558, p = 0.586; middle, t[14] = 2.764, p = 0.015; Welch’s t-test, late, t[12.010] = 1.617, p = 0.132, p = 0.016 after a Bonferroni correction method) and subtracted success rate on day N-1 from that on day N (right, two-way repeated ANOVA, stage, F[2,28] = 1.231, p = 0.307, group, F[1,14] = 10.797, p = 0.005, stage × group, F[2,28] = 2.360, p = 0.113; unpaired Student’s t-test, early, t[14] = 0.285, p = 0.780, late, t[14] = 1.271, p = 0.225, Welch’s t-test, middle, t[9.603] = 3.331, p = 0.008, p = 0.016 after a Bonferroni correction method). For pVLS inhibition (D), success rate (left, two-way repeated ANOVA, stage, F[2,20] = 17.642, p = 3.8 × 10-5, group, F[1,10] = 43.942, p = 5.9 × 10-5, stage × group, F[2,20] = 4.729, p = 0.012; simple main effect, early, p = 0.055, middle, p = 0.109, late, p = 4.2 × 10-5) and subtracted success rate on day N-1 from that on day N (right, two-way repeated ANOVA, stage, F[2,20] = 12.105, p = 3.6 × 10-4, group, F[1,10] = 16.310, p = 0.002, stage × group, F[2,20] = 8.500, p = 0.002; simple main effect, early, p = 0.595, middle, p = 0.085, late, p = 3.0 × 10-6). (E and F) Subtracted proportion of the WSW or LSL strategy in the groups injected into the aDLS at the middle stage (E, unpaired Student’s t-test, t[14] = 2.038, p = 0.061 for the WSW, and t[14] = 2.714, p = 0.017 for the LSL) and the late stage (F, unpaired Student’s t-test, t[14] = 0.898, p = 0.384 for the WSW, and t[14] = 0.226, p = 0.824 for the LSL). (G and H) Subtracted proportion of the WSW or LSL strategy in the groups injected into the pVLS at the middle stage (G, unpaired Student’s t-test, t(10) = 1.924, p = 0.083 for the WSW, and Welch’s t-test, t[6.095] = 1.364, p = 0.221 for the LSL) and the late stage (H, unpaired Student’s t-test, t[10] = 3.629, p = 0.005 for the WSW, and t[10] = 1.577, p = 0.146 for the LSL). Data are indicated as the mean ± s.e.m., and individual data are overlaid. The anteroposterior coordinates from bregma (mm) are shown (A and B). Scale bars; 2 mm (A and B). *p < 0.05, **p < 0.01, and ***p < 0.001.

Next, we analyzed the influence of temporal striatal inactivation on the behavioral strategy at the middle and late stages of the acquisition phase. We calculated the proportion of WSW and LSL strategies, and subtracted the values of WSW or LSL proportion on Day N-1 from those on Day N. The proportion of both strategies on Day N-1 were comparable between the MUS- and SAL-injected groups (Figures S2G and S2H). In the aDLS, the subtracted proportion of LSL strategy at the middle stage significantly increased in the MUS-injected group as compared to the SAL-injected group, although the subtracted proportion of WSW strategy tended to decrease in the MUS-injected group as compared to the control group (Figure 4E), and the subtracted proportion of WSW or LSL strategy at the late stage was similar between both groups (Figure 4F). In the pVLS, the subtracted proportions of the WSW and LSL strategies at the middle stage were similar between the MUS- and SAL-injected groups (Figure 4G), and the subtracted WSW proportion at the late stage significantly decreased in the MUS-injected group as compared to the SAL-injected group, although the subtracted LSL proportion was comparable between the two groups (Figure 4H). Our results show that transient inhibition of aDLS activity at the middle stage affects the LSL strategy and has a moderate effect on the WSW strategy, supporting that the aDLS mediates the acquisition process of auditory discrimination at the middle stage at least partly through the suppression of the behavior based on the response-outcome association. Although the impact of inhibition on the WSW strategy was modest, excitotoxic lesion of the aDLS influenced the WSW strategy in addition to the LSL strategy (see Figure 3J), suggesting that the aDLS acts to enhance the auditory discrimination by promoting the stimulus-response association, together with the suppression of the response-outcome association. The discrepancy in the extent of changes in the WSW strategy between the two treatments may be generated from the difference in the timing or sustainability of drug inhibition of the striatal subregion. In addition, our results show that pVLS inactivation at the late stage selectively impairs the WSW strategy, supporting contribution of the pVLS to the process via progression of the stimulus-response association.

Neuronal activity in the aDLS and pVLS during the acquisition phase

To examine the firing activity in striatal subregions during the acquisition phase of auditory discrimination, we simultaneously recorded neuronal activity in the aDLS and pVLS at the unilateral (left) side by using a multi-unit recording procedure (Figure 5A). Using immunostaining for tyrosine hydroxylase (TH) after the recordings, we verified the location of electrodes within the striatal subregions (Figure 5B). After the single lever press task, we conducted the auditory discrimination task for the recordings. In correct trials, the delay period (0.5 ± 0.2 s) was inserted between a lever press and a sound indicating the reward to distinguish neuronal activity related to the two events (Figure 5C). The behavioral data in individual rats were divided into three grouped sessions corresponding to the early, middle, and late stages, based on a sigmoid function analysis33 of the success rate in the task. The success rates in the grouped sessions were 52.7 ± 0.8 %, 61.6 ± 0.7 %, and 85.3 ± 1.7 % (mean ± s.e.m.) for the early, middle, and late stages, respectively, and the rate was significantly increased throughout the three stages (Figure S3A). However, the response time, omission rate, and response bias were similar among the stages (Figures S3B-S3D). Thus, these stages corresponded to the timing window defined in the acquisition phase of the transient MUS inhibition experiment of neuronal activity.

Multi-unit recording of neurons in striatal subregions and firing activity related to the behavioral outcome of RS-HR neurons in the aDLS

(A) Schematic illustration of a freely moving rat used for simultaneous multi-unit recordings in the aDLS and pVLS. Six rats were used for the following analysis. (B) Representative images showing positions of electrode tips (arrows) in the aDLS and pVLS (left rows). The recording sites estimated by electrode tracks and electrical marks in individual rats are shown (center and right rows). ac, anterior commissure; LV, lateral ventricle. (C) Sequence of some events in correct and error trials. The delay periods were pseudorandomly added between a correct lever press and the reward sound (0.5 ± 0.2 s). The room light turned off the extended time (4 s) after the correct response or immediately after the error response. (D) Mean firing rate (top rows) and auROC values (bottom rows) of RS-HR neurons in the aDLS during the period when the RS is presented (green shadows, 500 ± 200 ms after the lever press) at the early (n = 43 neurons), middle (n = 37 neurons), and late (n = 44 neurons) stages. Time bins with significant differences between the mean firing rate in the HR or LR trial and any of the rates in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. (E) Averaged firing rate during the RS period in the HR and LR trials at the early, middle, and late stages (Kruskal-Wallis test: HR trial, χ2 = 3.028, p = 0.220; and LR trial, χ2 = 1.249, p = 0.536). (F) Cumulative probability of the proportion of RS-HR neuron number at the three stages (Kruskal-Wallis test, χ2 = 45.378, p = 1.4 × 10-10; post hoc Tukey-Kramer test, p = 7.1 × 10-5 for early vs. middle, p = 1.0 × 10-9 for early vs. late, and p = 0.040 for middle vs. late). Data are indicated as the mean ± s.e.m. (D) or the median and quartiles with the maximal and minimal values (E). The anteroposterior coordinates from bregma (mm) are shown (B). Scale bars; 2 mm (B). *p < 0.05 and ***p < 0.001.

We identified 1,062 and 423 well-isolated neurons from the aDLS and pVLS, respectively, across the three stages (Figure S3E) and focused on the major population of striatal neurons, putative medium spiny neurons (MSNs; Figures S3F and S3G). The total numbers of MSNs at the early, middle, and late stages were 295, 364, and 360 neurons for the aDLS, and 87, 110, and 190 neurons for pVLS, respectively. Based on the combinatorial pattern of the tone instruction cue and lever press in our discrimination task, we categorized the electrophysiological data into four trial types: (1) high-frequency tone/right lever press (HR) and (2) low-frequency tone/left lever press (LL) as correct responses; and (3) low-frequency tone/right lever press (LR) and (4) high-frequency tone/left lever press (HL) as error responses (see Figure 1A). We identified HR or LL type neurons showing significant changes in firing rate related to the instruction cue onset (CO), choice response (CR), reward sound (RS), or first licking (FL) as compared to the baseline firing rate in each of two trials (Mann-Whitney U test, p < 0.01). These neurons were further divided into two groups based on increased or decreased activity (Figures S4A and S4B). In the following analyses, we focused on the HR and LL type neurons with increased event-related activity to explore the firing patterns of neurons underlying behavioral processes of auditory discrimination.

Subpopulations of aDLS neurons, but not of pVLS neurons, show firing activity related to the behavioral outcome

Previous studies have reported that neurons in the dorsal striatum show the outcome-related activity in the instrumental conditioning tasks24,34. We thus investigated whether aDLS and pVLS neurons exhibit a response related to the behavioral outcome during the acquisition of auditory discrimination. In the aDLS, the firing activity of RS-related HR type (RS-HR) neurons showed a higher extent in the increase during a period, when the RS is presented in the correct responses, in the HR trial as compared with the LR, LL, or HL trial throughout the three stages (Figure 5D, top rows). The rate of these neurons also moderately increased during the same period in the LR trial compared to the LL or HL trial across the three stages, although the extent of the increase was lower than that in the HR trial (Figure 5D, top rows). The area under the receiver operating characteristic (auROC) values revealed a significant difference of the HR trial from the LR trial during the RS period, whereas the values of HL and LL trials indicated a significant difference in the opposite direction from the LR trial during the same period (Figure 5D, bottom rows). The averaged firing rate during the RS period in the HR or LR trial appeared to modestly increase at the middle stage, but no significant difference existed among the stages (Figure 5E). To directly compare the number of neurons showing event-related activity among different stages, we calculated the number of RS-HR neurons by matching the trial number at each stage (Figure 5F). The proportion decreased progressively from the early to middle stages, and to the late stage. In addition, the firing rate of RS-related LL type (RS-LL) neurons exhibited a higher increase in the LL trial and a modest increase in the HL trial while the averaged firing rate during the RS period was similar among the three stages. Furthermore, the proportion of RS-LL neurons reduced progressively through the stages (Figures S5A–S5C). These results indicate that the aDLS contained neurons related to the outcome of choice, showing a greater increase in the firing activity for the correct responses (HR and LL trials) than the error responses (LR and HL trials), and that the proportion of these subpopulations was gradually reduced through the progress of learning. In contrast, during the RS period at all stages, the firing rate and auROC values of RS-HR and RS-LL neurons in the pVLS revealed no significant differences across the four kinds of trial (Figures S5D and S5E), indicating that the firing activity of pVLS neurons did not appear to be associated with the behavioral outcome.

Subpopulations of aDLS and pVLS neurons show sustained firing activity after the reward in different manners

The dorsal striatum is known to demonstrate a response of neuronal activity after the positive outcome in the instrumental behaviors35,36. We thus examined whether aDLS and pVLS neurons show the response after obtaining the reward during the acquisition phase of auditory discrimination. In the aDLS, the firing rate of FL-related HR type (FL-HR) neurons indicated a sustained elevation after the FL in the HR trial as compared with any of the other three trials across the early, middle, and late stages (Figure 6A, top rows). The auROC values revealed a significant difference of the LR, LL, or HL trial from the HR trial across the three stages (Figure 6A, middle rows). In addition, we calculated the averaged firing rate and increased firing period, defined as the total number of time bins above 3 s.d. of the baseline firing rate for 5 s after the FL in the HR trial. The averaged firing rate was comparable between the early and middle stages, but the rate was significantly lower at the late stage compared to either of the other two stages (Figure 6B). The increased firing period was also comparable between the early and middle stages, showing a significant reduction at the late stage (Figure 6C). The proportions of FL-HR neurons were similar between the early and late stages, although they showed a slight decrease at the middle stage (Figure 6D). There were few neurons with significant correlations between the firing rate and licking performance (Figure 6E). In addition, the firing rate of the FL-related LL type (FL-LL) neurons exhibited the largest increase in the LL trial at the three stages, showing a tendency to reduce the average firing rate and increased firing period after the FL across the stages (Figures S6A-S6C). The proportion of FL-LL neurons during the learning processes displayed similar changes as observed in the FL-HR neurons (Figure S6D). These results indicate that the aDLS contained neurons showing long-lasting, increased activity after the reward in the specific correct combination of the stimulus and response (HR or LL trial), and that the extent of firing activity and length of firing period in the response appeared to be gradually reduced with no correlation between the proportion of neurons and the progress of the learning stages.

Sustained activity after the reward of FL-HR neurons in the aDLS and pVLS

(A) Firing rate (top rows), auROC (middle rows), and licking rate (bottom rows) after the reward of FL-HR neurons in the aDLS at the early (n = 27 neurons), middle (n = 31 neurons), and late (n = 49 neurons) stages. Time bins with significant differences between the mean firing rate in the HR trial and any of the rate in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. Timings of the RS and FL are shown. (B) Averaged firing rate for 5 s after the FL in the HR trial (Kruskal-Wallis test, χ = 16.396, p = 2.8 × 10, post hoc Tukey-Kramer test, early vs. middle, p = 0.726, early vs. late, p = 7.2 × 10-4, middle vs. late, p = 0.009). (C) Averaged total number of time bins above 3 S.D. of the baseline firing of aDLS neurons for 5 s after the FL in the HR trial (Kruskal-Wallis test, χ2 = 12.803, p = 0.002, post hoc Tukey-Kramer test, early vs. middle, p = 0.173, early vs. late, p = 0.001, middle vs. late, p = 0.219). (D) Cumulative probability of the proportion of FL-HR neuron number (Kruskal-Wallis test, χ2 = 39.280, p = 3.0[×[10-9, post hoc Tukey-Kramer test, early vs. middle, p = 1.3 × 10-7, early vs. late, p = 0.995, middle vs. late, p = 2.3 × 10-7). (E) Distribution of the correlation coefficient between the numbers of spikes of FL-HR neurons and licking. Closed column indicates the number of neurons showing significant correlations between the two parameters (5 out of 107 neurons). (F) Firing rate (top rows), auROC (middle rows), and licking rate (bottom rows) after the reward of FL-HR neurons in the pVLS at the early (n = 17 neurons), middle (n = 11 neurons), and late (n = 64 neurons) stages. Time bins with significant differences between the rate in the HR trial and any of the rate in other trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. The timings of the RS and FL are shown. (G) Averaged firing rate of pVLS neurons after the FL in the HR trial (Kruskal-Wallis test, χ2 = 7.652, p = 0.022; post hoc Tukey-Kramer test, early vs. middle, p = 0.975, early vs. late, p = 0.076, and middle vs. late, p = 0.096). (H) Averaged total number of time bins above 3 S.D. of the baseline firing after the FL in the HR trial (Kruskal-Wallis test, χ2 = 0.023, p = 0.989). (I) Cumulative probability of the proportion of FL-HR neuron number (Kruskal-Wallis test, χ2 = 47.922, p = 3.9 × 10-11; post hoc Tukey-Kramer test, early vs. middle, p = 0.011, early vs. late, p = 1.7 × 10-4, and middle vs. late, p = 9.7 × 10-10). (J) Distribution of the correlation coefficient between the number of spikes in the pVLS and the licking. Closed column indicates the number of neurons showing significant correlations between the two parameters (5 out of 92 neurons). Data are indicated as the mean ± s.e.m. (A and F) or the median and quartiles with the maximal and minimal values (B, C, G, and H). **p < 0.01 and ***p < 0.001.

In the pVLS, FL-HR neurons showed a sustained elevation in the firing rate after receiving a reward in the HR trial compared to that in the LR or HL trial throughout the three stages (Figure 6F, top rows). A similar elevation in the firing rate was also observed in the LL trial throughout the stages (Figure 6F, top rows). The auROC values revealed a significant difference of the LR or HL trial, but not the LL trial, from the HR trial throughout the stages (Figure 6F, middle rows). The average firing rate did not show significant changes among the three stages (Figure 6G), and the increased firing period was also similar across the stages (Figure 6H). The proportion of FL-HR neurons increased at the late stage compared to the early or middle stage (Figure 6I). The firing rate showed almost no significant correlation to the licking performance (Figure 6J). In addition, FL-LL neurons exhibited similar firing patterns and changes in the proportion of neurons across the stages (Figures S6F–S6J). Therefore, the pVLS included neurons representing the long-lasting increased activity after the reward in the two kinds of correct responses (HR and LL trials) with the increased number of these neurons at the late stage of learning.

Neurons in the pVLS, but not in the aDLS, exhibit firing patterns related to the beginning and ending of the behavior

The dorsal striatum plays a key role in the acquisition and execution of behaviors based on the stimulus-response association, and the association is accompanied by a transient activation of neuronal firing immediately after the stimulus presentation or action in trials, which is considered to represent the beginning and ending of a learned behavior37–39. We first examined whether the firing activity of CO-related HR type (CO-HR) or CO-related LL type (CO-LL) neurons corresponds to the beginning of the behavior in the auditory discrimination task. In the aDLS, the firing rate of CO-HR or CO-LL neurons showed a significant difference at some time points in each trial compared to those in any of the other three trials, with a significant difference of auROC values between the trials (Figures S7A and S7C). However, the transiently increased activity did not emerge at the time points corresponding to the beginning of the behavior. In contrast, CO-HR neurons in the pVLS showed a transient increase in the firing rate immediately after the CO, and the activity appeared at the middle and late stages (Figure 7A, top). The transient activation was observed not only in the HR trial but also in the other three trials, being consistent with the results of a previous report40. The increased firing rate showed a slight but notable difference between the HR and LL trials with a significant difference in the auROC values between the two trials (Figure 7A, low), suggesting that different combinations between the stimulus and response may affect the level of firing activity. Differences in the averaged firing rate between the two kinds of trial was observed at the middle and late stages (Figure 7B). The proportion of CO-HR neurons increased at the middle and late stages compared to the early stage (Figure 7C). In addition, the firing rate of CO-LL neurons exhibited patterns similar to those observed in the CO-HR neurons (Figure S7E). The data indicate that subpopulations of pVLS neurons show firing activity related to the beginning of the behavior, with a rising number of neurons throughout discrimination learning.

Transient activity related to the beginning and ending of a behavior of CO-HR and CR-HR neurons in the pVLS

(A) Firing rate (top rows) and auROC values (bottom rows) of CO-HR neurons in the pVLS at the early (n = 4 neurons), middle (n = 21 neurons), and late (n = 24 neurons) stages. Time bins with significant differences between the mean firing rate in the HR trial and either of the rate in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. (B) Averaged firing rate of CO-HR neurons in HR and LL trials (0-600 ms after the CO) (Wilcoxon signed rank test; early, p = 0.875, middle, p = 0.006, and late, p = 0.004). (C) Cumulative probability of the proportion of CO-HR neuron number (Kruskal-Wallis test, χ2 = 51.114, p = 8.0 × 10-12; post hoc Tukey-Kramer test, early vs. middle, p = 9.6 × 10-10, early vs. late, p = 0.002, and middle vs. late, p = 6.4 × 10-4). (D) Firing rate (top rows) and auROC values (bottom rows) of CR-HR neurons in the pVLS at the early (n = 11 neurons), middle (n = 22 neurons), and late (n = 41 neurons) stages. Time bins with significant differences between the mean rate in the HR trial and any of the rate in other trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. The timing of the RS is presented. (E) Averaged firing rate of CR-HR neurons in HR and LL trials (0-600 ms after the CR) (Wilcoxon signed rank test; early, p = 0.278, middle, p = 0.758, and late, p = 0.279). (F) Cumulative probability of the proportion of CR-HR neuron number (Kruskal-Wallis test, χ2 = 13.786, p = 0.001; post hoc Tukey-Kramer test, early vs. middle, p = 0.003, early vs. late, p = 0.005, and middle vs. late, p = 0.976). Data are indicated as the mean ± s.e.m. (A and B) or the median and quartiles with the maximal and minimal values (B and E). **p < 0.01 and ***p < 0.001.

Next, we examined the firing activity of CR-related HR type (CR-HR) or CR-related LL type (CR-LL) neurons, corresponding to the ending of a behavior in auditory discrimination. In the aDLS, the firing rate of these neurons did not exhibit the transient increase related to the ending of the behavior, instead, there was a modest increase before the lever press (Figures S7B and S7D). In contrast, CR-HR neurons in the pVLS were transiently activated immediately after the lever press in the four kinds of trial at the middle and late stages, and the rate appeared to gradually decline toward the RS presentation (Figure 7D). The averaged firing rate was similar between the HR and LL trials though the stages (Figure 7E). The proportion of the CR-HR neurons increased at the middle and late stages compared to the early stage (Figure 7F). In addition, the rate of CR-LL neurons displayed similar firing patterns to those observed in the CR-HR neurons (Figure S7F). Thus, subpopulations of pVLS neurons also represent the activity related to the ending of the behavior in auditory discrimination, increasing the neuron number throughout the learning stages in a similar pattern to neurons relevant to the beginning of the behavior in the task.

Discussion

Our findings demonstrate that the aDLS and pVLS are required for the acquisition of auditory discrimination at different stages of learning, and that the DMS is not necessary for the acquisition of discrimination, although it appears to contribute to the execution of actions, in particular to the regulation of response time. These results not only support some recent studies reporting no requirement of the DMS in the procedural learning20,41,42, but also reveal important roles of two different subregions in the lateral striatum in the external cue-dependent decision-making. We propose that aDLS and pVLS neurons act to integrate the new learning of auditory discrimination in spatiotemporally and functionally distinct manners, showing a remarkable difference from the prior learning model that proposes the functional dominance changes from the DMS to DLS subregions during procedural learning12–16.

Our 18F-FDG-PET imaging study demonstrated that the brain activity in the aDLS reaches a peak at the middle stage of the acquisition phase of auditory discrimination (Figure 2I). The results of excitotoxic lesion experiment of striatal subregions suggest that the aDLS functions to enhance the acquisition of discrimination through the promotion of behavior based on the stimulus-response association and the suppression of behavior based on the response-outcome association (Figure 3J). Transient inhibition experiment of neuronal activity confirmed this aDLS function at the middle stage of the learning, although the effect of drug inhibition was moderate for the promotion of the stimulus-response behavior (Figures 4E and 4F). In addition, we found subpopulations of aDLS neurons (RS-related neurons) which represent the behavioral outcome and differentiate the correct (HR or LL) and incorrect (LR or HL) choices (Figures 5D and S5A). The distinct responses of these neurons may explain the mechanism by which the aDLS regulates the stimulus-response and response-outcome associations through distinct neural circuits leading to a differential control of the two associations. Alternatively, the responses mainly promote the behavior based on the stimulus-response association, thereby secondarily inhibiting the response-outcome association. Indeed, some previous studies have suggested that an action based on either the stimulus-response association or response-outcome association is generated as a result of competition between different striatal subregions43. We also found other subpopulations of aDLS neurons (FL-related neurons) showing long-lasting increased activity after obtaining reward in response to a specific combination between the stimulus and response (HR or LL trial) (Figures 6A and S6A). This activity may function as the feedback signal to promote the stimulus and response association, resulting in the enhancement of discrimination learning. The proportion of neurons related to the behavioral outcome showed a gradual reduction through the progress of auditory discrimination (Figures 5E and S5C), and the extent of the firing activity and the length of the firing period of neurons showing sustained activation after reward were also decreased at the late stage of learning (Figures 6B and 6C). The learning-dependent changes in the properties of these subpopulations in the aDLS may explain the mechanism underlying the earlier acquisition process of the auditory discrimination.

The 18F-FDG-PET imaging indicated that the pVLS activity was gradually elevated through the acquisition phase of auditory discrimination, showing the maximal value at the late stage (Figure 2J). The lesion experiment suggested that the pVLS drives the formation of discrimination learning via the selective progression of the stimulus-response association (Figure 3K), and the transient inhibition experiment ascertained the pVLS function at the late stage of learning (Figures 4G and 4H). Previous studies have shown that the stimulus-response association is encoded in the DLS or VLS as the transient neuronal activity representing the beginning and ending of a learned behavior37-39,44. A recent study reported that the VLS is causally related to learned performance based on the auditory stimulus, while the DLS shows a weaker relationship to the behavior22. In the present study, we found subpopulations of pVLS neurons that represent the beginning and ending of a behavior (CO- and CR-related neurons, respectively) (Figures 7A, 7D, S7E, and S7F), showing an increased number of neurons along with the progress of discrimination learning (Figures 7C and 7F). These data suggest that pVLS neurons contribute to the formation of the stimulus-response association through the learning processes. In addition, we found another subpopulation of pVLS neurons (FL-related neurons) that are continuously activated after the reward outcome in response to two kinds of correct choices (HR and LL trials) (Figures 6F and S6F), and the proportion of neurons was increased at the late stage (Figures 6I and S6I). This activity may act as a feedback signal to consolidate the stimulus and response association, leading to the maintenance of associative memory at the late stage during the acquisition phase of learning.

In the present study, we identified neurons with four kinds of event-related activity in the aDLS and pVLS, and classified them into the HR and LL type neurons in which their firing activity showed a significant increase in the HR and LL trials, respectively, as compared to the baseline (Figures 5–7 and S5-S7). Parts of these two types of neurons with the respective event-related activity overlapped, and the numbers of overlapping neurons varied between striatal subregions (Figures S8A and S8B). In the pVLS, the overlapping neurons for the same event-related activities were 35.1%, 45.2%, and 67.0% in the CO, CR, and FL, respectively. In contrast, the aDLS had lower overlaps of 12.6% and 22.7% in the RS and FL, respectively. Thus, distinct neurons in the aDLS tend to have properties of a single trial type while the same neurons in the pVLS seem to possess properties of both types. One possible explanation for the diversity in the proportion of overlapping neurons between the two subregions may be the difference of input patterns into striatal neurons or plastic changes in functional connectivity during learning processes. The detailed mechanism by which aDLS and pVLS neurons react to the events in a different manner should be investigated in the future.

Anatomically, striatal subregions are known to form heterogeneous cortico-basal ganglia loops in mice2,45,46. We examined the neural circuits linked to the aDLS and pVLS in rats by using bidirectional tracers. The results showed that two striatal subregions receive innervations from different subregions in the cerebral cortex, intralaminar thalamus, and substantia nigra pars compacta, and send projections to different subregions in the basal ganglia nuclei, including the globus pallidus, entopeduncular nucleus, and substantia nigra pars reticulata (Figure S9), confirming that the aDLS and pVLS form the heterogenous cortico-basal ganglia loops in the rat brain. These data suggest that the aDLS and pVLS mediate the acquisition of auditory discrimination via the parallel loops, and cooperate to achieve neural processing related to the dual regulation of the stimulus-response and response-outcome associations. Further studies are needed to elucidate the mechanism by which the two kinds of loops are processed separately and integrate the new learning of discrimination.

Materials and Methods

Animals

Animal care and handling procedures were carried out in accordance with the guidelines established by RIKEN Center for Biosystems Dynamics Research, Fukushima Medical University, and Osaka City University. All procedures were approved by their Institutional Animal Care and Use Committees. All efforts were made to minimize the number of animals used and their suffering throughout the course of the experiments. Male Long-Evans rats (8–13 weeks old, Institute for Animal Reproduction (Ibaraki, Japan)) were used for the present study. The rats were maintained at 22 ± 2 °C and 60% humidity in a 12-h light/12-h dark cycle, and food and water were continuously available unless stated otherwise.

Surgery

The rats underwent a stereotaxic surgery under isoflurane anesthesia (1–5%). For excitotoxic lesion of striatal neurons, IBO was dissolved in PBS (0.8 mg/mL, Wako Pure Chemical), and IBO solution (0.3 μL per site) was injected bilaterally with the coordinates (mm) from bregma and dura for the aDLS (AP +1.6, ML ±3.3, DV −3.9), pVLS (AP −1.6, ML ±4.6, and DV −5.6), or DMS (AP −1.6, ML ±1.4, and DV −4.8) according to the atlas of rat brain47. Injection was performed through a glass pipette (diameter: 60 μm) at a constant velocity of 0.1 mL/min with the microinfusion pump (EICOM). For transient inactivation of striatal neurons, two guide cannulae (length: 10 mm, outer diameter: 26 gauge) were bilaterally placed on the skull and fixed using four screws and dental cement. The chip of the cannula was placed at 1 mm above the target coordinates. A solution containing MUS (Sigma-Aldrich) in saline (0.1 μg/μL, 0.25 μL per site) was injected into the target coordinates through the internal cannula (outer diameter: 30 gauge) at a constant velocity of 0.25 μL/min. For anterograde and retrograde tracing, a solution containing cholera toxin subunit B (CTb; 1.0 mg/mL) conjugated to Alexa 488 or Alexa 555 (Invitrogen) as a bidirectional tracer was injected into the aDLS and pVLS (0.3 μL/site), respectively. The coordinates for the injection into striatal subregions were the same as described above. For the electrophysiological experiments, surgery was performed as described previously48. A 64-channel silicon probe (Buzsaki64spl; NeuroNexus) attaching to a 3D-printed microdrive with a movable screw was chronically implanted with the coordinates (mm) from the bregma and dura into the left aDLS (AP +1.6, ML −3.3, and DV −2.5) and pVLS (AP −1.6, ML −4.6, and DV −3.6). A faraday cage composed of copper mesh was secured on the skull with dental cement to reduce electrical noise and avoid damage to the implants. The rats were allowed to recover for 1 week after the surgeries.

Behavioral analysis

Behavioral tasks were conducted in operant conditioning chambers (30.5 × 24.1 × 29.2 cm; ENV-007CT; MED Associates) equipped with two retractable levers on both sides of a reward port in the front panel (ENV-203M-45, ENV-652AM, and ENV-201A; MED Associates). A multi-tone generator (ENV-223; MED Associates) and room light were mounted on the top of the center and corner panels on the rear side, respectively. For the electrophysical experiment, a 3D printed reward port was used to detect the licking and to prevent contact with the head attachment and front panel. The behavioral experiment in intact rats was performed using grain pellets (F0021, Bioserv) as a food reward. To prevent spillover of radioactivity into the temporal cortex as a consequence of excessive 18F-FDG accumulation into the muscles by their activity to eat the pellets49,50, the PET imaging experiments were carried out using 0.02% saccharine solution (B0131, Tokyo Chemical Industry) as a liquid reward. The electrophysiological experiment was also carried out using the saccharine solution. The pharmacological experiment was performed using the pellets.

The rats were subjected to dietary restrictions of 12 g/day (CE-2, CLEA Japan) and water access was limited to 6 to 9 h in accordance with the reward used in the experiments for 3 days before the beginning of the training schedule, and the body weights of the rats were maintained at ∼85% of their normal weights. The training schedule consisted of four continuous steps. In the first step (magazine training), the rats were habituated to the operant conditioning chambers, and the reward was presented with the click every 20 s. Each daily session was started by illumination of the room light, and continued for 30 min. The sessions were conducted for 3 consecutive days. In the second step (shaping), only one lever on the left or right side was inserted. When the rats pressed the lever, they earned a reward accompanied by a click. Each daily session was started by illumination of the light, and lasted 30 min or until the criteria, in which the rats received 90 rewards. When the number of lever presses on either side reached the criteria, the lever was inserted in the opposite side on the next day. When the rats achieved the criteria on both sides for 2 or 3 days, they moved to the following step. In the third step (single lever press task), each trial was started by illumination of the light and insertion of one lever on either the left or right side in a pseudorandom manner with equal frequencies on both sides for every set of four trials. When the rats pressed the lever, the reward was presented with the sound, and the lever was immediately retracted. The room light turned off 4 s after the lever press. When they did not press the lever within 10 s after the start, the trial was terminated when the light turned off. The trial was repeated every 20 s. Each daily session continued for 30 min, and the sessions were conducted for 4 consecutive days. The number of lever presses in each session was counted. In the final step (two-alternative auditory discrimination task), each trial was started by the presentation of a high (10 kHz) or low (2 kHz) frequency tone as an instruction cue in a pseudorandom manner. After three seconds, the room light was illuminated, and two retractable levers were inserted at the same time. The rats were required to press the left or right levers in response to the low or high tone cues, respectively. When the rats pressed the correct lever, the reward was presented with the sound, and the levers were immediately retracted. Correct response allowed the rats to access the reward: the retractable spout delivering water was inserted for 1 s for the PET imaging experiment; a pellet was delivered by a food dispenser for the pharmacological experiment; and three drops of water were delivered through a solenoid valve for the electrophysiological experiment. The room light turned off 4 s after the correct response. On the other hand, error response resulted in the retraction of levers and the turning off of the room light. No response within 10 s after the lever insertion was recorded as an omission response. The success rate in each session was calculated by dividing the number of correct response trials with the number of lever press trials minus the number of omission trials. The trial was repeated every 20 s. Each daily session continued for 60 min, and the sessions were conducted for 14 or 24 days according to the experiments.

To assess the behavioral strategies in the auditory discrimination task, a trial-by-trial analysis was conducted; the WSW strategy was defined as a correct response, in which, after a correct response during the previous trial, the rats press the opposite lever during the current trial in response to the shift of the tone instruction cue; the LSL strategy was defined as an error response, in which, after an error response during the previous trial, the rats press the opposite lever during the current trial despite the shift of the instruction cue. The proportion of both strategies was calculated by dividing the number of each strategy with the number of lever press trials minus the number of omission trials.

For the transient inactivation experiment, MUS injections were bilaterally performed 20 min before start of the session. For habituation, all rats were injected with saline bilaterally into the corresponding striatal subregions on the last day (Day 4) of the single lever press task. During the auditory discrimination task, the first injection was conducted on Day 2 of the task. The second and third injections were conducted the next day, after the success rate had exceeded 60% and 80%, respectively. Each session continued for 15 min.

For the electrophysiological experiments, a delay period (0.5 ± 0.2 s) was added in the auditory discrimination task between the lever press and reward sound to distinguish the motor- and reward-related responses of neurons. When the rats pressed the lever, the lever was immediately retracted. After the delay period, the reward was presented with a sound. The room light turned off 4 s after the lever press. The trial was repeated every 20 s, and each daily session continued for 60 min.

PET imaging analysis

An 18F-FDG-PET scan was performed with a microPET Focus220 (Siemens Medical Solutions, Knoxville) designed for high-resolution imaging of small animals (spatial resolution of 1.4 mm in full width at half maximum at the center of the field of view) as described previously26. Serial 18F-FDG-PET scans were performed on the last day of the single lever press task and on Days 2, 6, 10, and 24 of the auditory discrimination task. Under the awake condition, a solution of 18F-FDG (ca. 74 MBq/0.4 mL) was intravenously injected just before the beginning of each behavioral session through an indwelling catheter attached to the tail. After the 30-min session, the rats were returned to their home cage, and then a 30-min static PET scan was performed under anesthesia with 2.0– 2.5% isoflurane 55-min after the 18F-FDG injection. During the PET scan, the body temperature of the anesthetized rats was maintained at 37°C using a small animal warmer (BWT-100A, Bio Research Center). The acquired emission data were sorted into a single sinogram and reconstructed by a standard 2D filtered back projection (FBP) with a ramp filter and cutoff frequency at 0.5 cycles per pixel or by a statistical maximum a posteriori probability algorithm (MAP), 12 iterations with point spread function effect.

Voxel-based analysis of PET images was conducted according to the method described previously26. Briefly, individual MAP-reconstructed FDG (MAP-FDG) images were aligned to a FDG template image of Long Evans rats using PMOD software package (version 3.2, PMOD Technologies), and then the aligned images were transformed into the space of a T2 weighted MRI templates for the rats (https://www.nitrc.org/projects/tpm_rat/). The transformation parameters obtained from individual MAP-FDG images were used to match the individual FBP-reconstructed FDG (FBP-FDG) images with the MRI templates. The voxel size of the template was magnified by 10 to a size similar to that of a human brain. All FBP-FDG images were resampled with a voxel size of 1.2 × 1.2 × 1.2 mm and spatially smoothed through an isotropic Gaussian kernel (6-mm FWHM). Voxel-based statistical analysis was assessed by using the Statistical Parametric Mapping software (version 8; https://www.fil.ion.ucl.ac.uk/spm/). The voxel of interest (VOI) was determined as voxels with more than 85% of the maximum T-value within activated areas obtained by voxel-based statistical comparison of FBP-FDG images. The average regional 18F-FDG uptake was calculated on the basis of the values of the VOI in FBP-FDG images.

Electrophysiological recording

Extracellular multi-unit recording was conducted at a sampling rate of 20[kHz from the striatum of freely moving rats by using a 512-ch acquisition board (Open Ephys) via 64-ch recording headstages (C3325; Intan Technologies). The tips of the probe were lowered by turning the screw (> 140[µm/day) until it reached the target region (the aDLS and pVLS) during the single lever press task. The tips were sometimes lowered by the same procedure (140 to 280 µm/day) after the daily recording session to improve cell yield.

Spike sorting was performed automatically by the Kilosort software (https://github.com/cortex-lab/KiloSort). Clear noise clusters were removed in the first manual sorting using the phy GUI (https://github.com/cortex-lab/phy). The remaining clusters were then re-clustered by using the Klusta Kwik (https://klusta.readthedocs.io/en/latest/), and the clusters were adjusted in the second manual sorting by using the phy GUI again. The sorted clusters satisfying all the following criteria were used for further analysis: L-ratio < 0.05, isolation distance > 15, and ISI index < 0.251,52. Well-isolated single neurons (n = 1,485; aDLS, n = 1,062; pVLS, n = 423) were classified as putative medium spiny neurons (MSNs), fast spiking interneurons (FSIs), tonically active neurons (TANs), and unclassified interneurons (UIs). Putative MSNs and TANs were separated from FSIs and UIs by trough-to-peak duration of mean spike waveform (> 0.6 ms) according to previous studies, with slight modifications53,54. Putative FSIs were then separated from UIs by utilizing the proportion of time associated with long inter-spike intervals. This was achieved by summing the inter-spike intervals longer than 2 s and subsequently dividing the resulting sum by the total recording time (PropISIs > 2 s)55. Neurons with values of PropISIs > 2 s less than 0.4 were classified as putative FSIs, while neurons with PropISIs > 2 s greater than 0.4 were classified as putative UIs. The putative MSNs and TANs were separated by measuring the post-spike suppression55. We measured the length of suppressed time during which the firing rate of a neuron was suppressed following a spike by counting the number of 1-ms bins in its autocorrelation function until the rate of the neuron was equal to or greater than its averaged rate over the 600-ms to 900-ms autocorrelation bins. Putative TANs were separated by post-spike suppression of less than 40 ms, and the remaining neurons were classified as putative MSNs. Because the numbers of FSIs, TANs, and UIs were low (79 out of 1,485, 5.3%), they were excluded from further analysis.

To divide the three stages during the acquisition phase, the sigmoidal function was fitted using the following four-parameter logistic equation:

Where α is the inflection point, β is a variable for slope, and λ and γ are parameters related to the lower asymptote (γ) and upper asymptote (λ) of the learning curve. The function was used to evaluate the day when the success rate reached 20% (t20%) or 90% (t90%) of the maximum rate change defined by the difference between the γ and λ. Three stages were defined according to the following criteria: early stage (up to 3 days before the middle stage); middle stage (3 days including before and after the day of t20%); and late stage (3 days including the day of t90% and before). One rat learned the discrimination faster, and the success rate reached t90% 2 days after the day of t20%. One day before the day of t90% was assigned to the late stage. For the following analysis, spike data in each striatal subregion were separated into the early, middle, and late stages (n = 295, 364 and 360, respectively, for the aDLS; and n = 87, 110 and 190, respectively, for the pVLS).

To determine the neurons showing significant changes in firing rate related to the task events, electrophysiological data were grouped into four trial types based on the instruction cue (high or low tone) and the choice (right or left lever); HR, high tone and right lever press; HL, high tone and left lever press; LR, low tone and right lever press; and LL, low tone and left lever press. Neurons with three or more trial data were analyzed. We identified the HR or LL type neurons, detecting significant increases or decreases of the firing rate by applying the Wilcoxon rank-sum test (p < 0.01) to trial-by-trial spike number between the baseline period (BL, −1,500 to −500 ms or −1,100 to −500 ms before presentation of the instruction cue) and any of four peri-event periods (CO, 0 to +1,000 ms after the onset of the instruction cue; CR, −700 to +300 ms around the time of lever press; RS, −300 to 300 ms around the time of sound onset instructing the reward delivery; and FL, 0 to +1,000 ms after the time of first licking after the CR). Any spikes around the events were averaged in 50-ms bins. The event-related activity was shown as averaged firing rates aligned based on the timing of an event for each trial. The auROC values were calculated by using the spike counts between the HR or LR trials and any of the other three trials. Bins showing significant differences in the auROC data were determined by applying the Wilcoxon rank-sum test (p < 0.05) to the distribution of 200 ms bin-by-bin auROC values and 0.5.

To compare the number of neurons showing event-related activity among different stages, the number of HR or LL type neurons were calculated by matching the trial number at each stage. Twenty trials were randomly selected for a neuron, and neurons with less than 20 trials were removed from the analysis. The neurons with event-related activity were identified as described above. The proportion (percentage) of neurons relative to the total number of neurons was calculated at each individual stage. The same analysis was repeated 30 times and the distribution of the proportion of neurons with event-related activity at the stage was obtained. The cumulative probability against the proportion was calculated using cumulative distribution function (cdfplot, MATLAB code from Mathworks).

Licking responses were detected with an infrared photobeam sensor placed in the 3D printed reward port, and generated timestamps for the onset of each spout contact according to the analog voltage signal. The licking responses corresponding to four kinds of trial in the FL-related neurons were shown as averaged rates (50-ms bins) aligned based on the timing of the CR. To calculate the correlation coefficient between neuronal activity and licking responses, the spikes and licking for 5 s after the FL were counted every trial for each FL-related neuron. Trials in which neurons never responded were removed from both the spike and licking data. The correlation coefficients between the two remaining data were calculated for each neuron (corr, MATLAB code from Mathworks), and neurons showing significant differences were determined by applying the Pearson correlation coefficient (p < 0.01).

Histology

After the behavioral experiments, the rats were anesthetized with pentobarbital (50 mg/kg body weight) and isoflurane (3%), then transcardially perfused 4% paraformaldehyde in phosphate-buffer (PB). The brains were post-fixed overnight at 4 [and cryoprotect with 10, 20, and 30% sucrose in PB. Brain tissues were cut into 30-µm sections by using a cryostat (Leica). For the lesion experiment, cresyl violet staining or immunostaining were performed to define the lesioned areas in the striatum. The sections were stained with a mouse monoclonal anti-NeuN IgG (1:1000 dilution; Millipore, Cat# MAB377, RRID:AB_2298772), and then with a goat anti-mouse IgG conjugated with Alexa Fluor 488 (1:400 dilution; Thermo Fisher Scientific, Cat# A-11029, RRID:AB_2534088). The sections were counterstained with 0.3 μM 4’,6-diamidino-2-phenylindole (DAPI; Sigma-Aldrich). For transient inactivation experiment, sections were stained by cresyl violet to visualize the positions of guide cannulae. For the electrophysiological experiments, small lesions were made by applying an electrolytic current (3 μA, 10 s) through top and bottom channels of each shank by using the stimulus isolator (A365, World Precision Instrument). Sections through the striatum were stained with a mouse monoclonal anti-TH IgG (1:1,000 dilution; Millipore, Cat# MAB318, RRID:AB_2201528), and then with a goat anti-mouse IgG conjugated with Alexa Fluor 488 (1:400 dilution; Thermo Fisher Scientific, Cat# A-11029, RRID:AB_2534088). The recording sites were verified by the staining with DAPI and red fluorescent Nissl stain solution (1:200 dilution; Thermo Fisher Scientific, Cat# N21482, RRID:AB_2620170). Fluorescence and immunostained signals were visualized under fluorescence microscopes (BZ-X800; Keyence) and confocal microscope (A1R; Nikon).

Statistical analysis

Statistical analysis was performed using MATLAB (MathWorks) and SPSS Statistics (Ver. 27, IBM). Comparisons of data between two groups were performed using the Student’s t-test (paired), the Wilcoxon signed rank test (paired), and the Wilcoxon rank-sum test (unpaired). Comparisons of data among three groups were constructed using a one-way repeated measures ANOVA, two-way repeated measures ANOVA, or the Kruskal-Wallis test (without assuming normal distribution and homogeneity of variances). All statistical tests were two-sided. The Bonferroni post-hoc test and Mann–Whitney U-test with Bonferroni adjustment were carried out as needed to compare means. For boxplots, the data points below the lower (25th percentile) −1.5 interquartile range (IQR) and above the upper (75th percentile) quartiles +1.5 IQR were regarded as outliers, and they are not included in the boxplots.

Supporting information

eLife Assessment

This study provides a potentially valuable understanding of the contribution of different striatal subregions, the anterior Dorsal Lateral Striatum (aDLS) and the posterior Ventrolateral Striatum (pVLS), to auditory discrimination learning. The combined methods used to probe this are compelling, yet the data presented are incomplete to support the conclusions. There is insufficient data visualization of learning vs. performance, and missing details about timing of manipulations and microPET imaging.

Reviewer #1 (Public Review):

Summary:

In this study, Setogawa et al. employ an auditory discrimination task in freely moving rats, coupled with small animal imaging, electrophysiological recordings, and pharmacological inhibition/lesioning experiments to better understand the role of two striatal subregions: the anterior Dorsal Lateral Striatum (aDLS) and the posterior Ventrolateral Striatum (pVLS), during auditory discrimination learning. Attempting to better understand the contribution of different striatal subregions to sensory discrimination learning strikes me as a highly relevant and timely question, and the data presented in this study are certainly of major interest to the field. The authors have set up a robust behavioral task and systematically tackled the question about a striatal role in learning with multiple observational and manipulative techniques. Additionally, the structured approach the authors take by using neuroimaging to inform their pharmacological manipulation experiments and electrophysiological recordings is a strength.

However, the results as they are currently presented are not easy to follow and could use some restructuring, especially the electrophysiology. Also, the main conclusion that the authors draw from the data, that aDLS and pVLS contribute to different phases of discrimination learning and influence the animal's response strategy in different ways, is not strongly supported by the data and deserves some additional caveats and limitations of the study in the discussion.

Strengths:

See above. In addition, the electrophysiology data is a major strength.

Weaknesses:

(1) The authors have rigorously used PET neuroimaging, which is an interesting non-invasive method to track brain activity during behavioral states. However, in the case of a freely moving behavior where the scans are performed ~30 minutes after the behavioral task, it is unclear what conclusions can be drawn about task-specific brain activity. The study hinges on the neuroimaging findings that both areas of the lateral striatum (aDLS and pVLS) show increased activity during acquisition, but the DMS shows a reduction in activity during the late stages of behavior, and some of these findings are later validated with complementary experiments. However, the limitations of this technique can be further elaborated on in the discussion and the conclusions.

a) In commenting on the unilateral shifts in brain striatal activity during behavior, the authors use the single lever task as a control, where many variables affecting neuronal activity might be different than in the discriminatory task. The study might be better served using Day 2 measurements as a control against which to compare activity of all other sessions since the task structures are similar.

b) From the plots in J, K, and L, it seems that shifts in activity in the different substructures are not unilateral but consistently bilateral, in contrast to what is mentioned in the text. Possibly the text reflects comparisons to the single lever task, and here again, I would emphasize comparing within the same task.

(2) In Figure 2, the authors present compelling data that chronic excitotoxic lesions with ibotenic acid in the aDLS, pVLS, and DMS produce differential effects on discrimination learning. However, the significant reduction in success rate of performance happens as early as Day 6 in both IBO groups in both aDLS and pVLS mice. This would seem to agree with conclusions drawn about the role of aDLS in the middle stages of learning in Figure 2, but not the pVLS, which only shows an increased activity during the late stages of the behavior.

(3) In Figure 4, the authors show interesting data with transient inactivation of subregions of the striatum with muscimol, validating their findings that the aDLS mediates the middle and the pVLS the late stages of learning, and the function of each area serves different strategies. However, the inference that aDLS inactivation suppresses the WSW strategy "moderately" is not reflected in the formal statistical value p=0.06. While there still may be a subtle effect, the authors would need to revise their conclusions appropriately to reflect the data. In addition, the authors could try a direct comparison between the success rate during muscimol inhibition in the mid-learning session between the aDLS and pVLS-treated groups in Figure 4C (middle) and 4D (middle). If this comparison is not significant, the authors should be careful to claim that inhibition of these two areas differentially affects behavior.

(4) The authors have used in-vivo electrophysiological techniques to systematically investigate the roles of the aDLS and the pVLS in discriminatory learning, and have done a thorough analysis of responses with each phase of behavior over the course of learning. This is a commendable and extremely informative dataset and is a strength of the study. However, the result could be better organized following the sequence of events of the behavioral task to give the reader an easier structure to follow. Ideally, this would involve an individual figure to compare the responses in both areas to Cue, Lever Press, Reward Sound, and First Lick (in this order).

(5) An important conceptual point presented in the study is that the aDLS neurons, with learning, show a reduction in firing rates and responsiveness to the first lick as well as the behavioral outcome, and don't play a role in other task-related events such as cue onset. However, the neuroimaging data in Figure 2 seems to suggest a transient enhancement of aDLS activity in the mid-stage of discriminatory learning, that is not reflected in the electrophysiology data. Is there an explanation for this difference?

(6) A significant finding of the study is that CO-HR and CO-LL responses are strikingly obvious in the pVLS, but not in the aDLS, in line with the literature that the posterior (sensory) striatum processes sound. This study also shows that responses to the high-frequency tone indicating a correct right-lever choice increase with learning in contrast to the low-frequency tone responses. To further address whether this difference arises from the task contingency, and not from the frequency representation of the pVLS, an important control would be to switch the cue-response association in a separate group of mice, such that high-frequency tones require a left lever press and vice versa. This would also help tease apart task-evoked responses in the aDLS, as I am given to understand all the recording sites were in the left striatum.

Reviewer #2 (Public Review):

The study by Setogawa et al. aims to understand the role that different striatal subregions belonging to parallel brain circuits have in associative learning and discrimination learning (S-O-R and S-R tasks). Strengths of the study are the use of multiple methodologies to measure and manipulate brain activity in rats, from microPET imaging to excitotoxic lesions and multielectrode recordings across anterior dorsolateral (aDLS), posterior ventral lateral (pVLS)and dorsomedial (DMS) striatum.

The main conclusions are that the aDLS promotes stimulus-response association and suppresses response-outcome associations. The pVLS is engaged in the formation and maintenance of the stimulus-response association. There is a lot of work done and some interesting findings however, the manuscript can be improved by clarifying the presentation and reasoning. The inclusion of important controls will enhance the rigor of the data interpretation and conclusions.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public Review):

Summary:

In this study, Setogawa et al. employ an auditory discrimination task in freely moving rats, coupled with small animal imaging, electrophysiological recordings, and pharmacological inhibition/lesioning experiments to better understand the role of two striatal subregions: the anterior Dorsal Lateral Striatum (aDLS) and the posterior Ventrolateral Striatum (pVLS), during auditory discrimination learning. Attempting to better understand the contribution of different striatal subregions to sensory discrimination learning strikes me as a highly relevant and timely question, and the data presented in this study are certainly of major interest to the field. The authors have set up a robust behavioral task and systematically tackled the question about a striatal role in learning with multiple observational and manipulative techniques. Additionally, the structured approach the authors take by using neuroimaging to inform their pharmacological manipulation experiments and electrophysiological recordings is a strength.

However, the results as they are currently presented are not easy to follow and could use some restructuring, especially the electrophysiology. Also, the main conclusion that the authors draw from the data, that aDLS and pVLS contribute to different phases of discrimination learning and influence the animal's response strategy in different ways, is not strongly supported by the data and deserves some additional caveats and limitations of the study in the discussion.

We appreciate the reviewer’s valuable feedback, which has been beneficial for improvement of our manuscript. In response to the reviewer’s comments, we have revised multiple parts of the manuscript, including explanations of electrophysiological data. We have also provided additional data to support our main conclusion and addressed caveats and limitations related to the data in the Discussion section. For more details, please refer to the responses to each comment.

Comment 1: The authors have rigorously used PET neuroimaging, which is an interesting noninvasive method to track brain activity during behavioral states. However, in the case of a freely moving behavior where the scans are performed ~30 minutes after the behavioral task, it is unclear what conclusions can be drawn about task-specific brain activity. The study hinges on the neuroimaging findings that both areas of the lateral striatum (aDLS and pVLS) show increased activity during acquisition, but the DMS shows a reduction in activity during the late stages of behavior, and some of these findings are later validated with complementary experiments. However, the limitations of this technique can be further elaborated on in the discussion and the conclusions.

As described in our response to the following two comments (a, b) from the reviewer, in the PET imaging study we first analyzed task-related activity by comparing 18F-FDG uptake on different days of the auditory discrimination task with that on Day 4 of the single lever press task as a control. Next, we analyzed learning-dependent activity by comparing the uptake on different days of the discrimination task with that on Day 2 of the same task. Based on the results of both analyses, we concluded that the activity in the striatal subregions changes during the progress of discrimination learning. The behavioral significance of striatal subregions was tested by excitotoxic lesion and pharmacological blockade experiments. The explanation of imaging data analysis may have been insufficient to fully communicate dynamic changes in the activity of striatal subregions. Therefore, we have clarified our voxel-based statistical parametric analysis method to better explain the dynamic activity changes in the striatal subregions. Please refer to the following responses to comments 1 (a, b).

Comment 1 (a): In commenting on the unilateral shifts in brain striatal activity during behavior, the authors use the single lever task as a control, where many variables affecting neuronal activity might be different than in the discriminatory task. The study might be better served using Day 2 measurements as a control against which to compare activity of all other sessions since the task structures are similar.

We initially analyzed task-related activity by comparing 18F-FDG uptake on one of Days 2, 6, 10, or 24 of auditory discrimination task with that on Day 4 of the single lever press task. This task was used as a control that does not require a decision process based on the auditory stimulus. We observed significant increases in the activity of the unilateral aDLS on Day 6 and in that of the bilateral pVLS on Day 10 of the discrimination task. We also observed a significant decrease in the unilateral DMS on Day 24 (see Figures 2F and 2G). Next, as suggested, we compared the uptake on one of Days 6, 10, or 24 with that on Day 2 as a control to evaluate learning-dependent activity. The activity showed significant increases in the bilateral aDLS on Day 6 and in the unilateral pVLS on Day 10, and a significant decrease in the bilateral DMS on Day 24 (see Figures 2H).

The reviewer has suggested a discrepancy in the activity of the unilateral or bilateral striatal subregions under certain conditions between the image data (shown in Figures 2F–H) and plot data (Figures 2J–L). This discrepancy is also suggested in the following Comment 1 (b). For example, in the image data the brain activity was increased in the unilateral (left) aDLS on Day 6 of the discrimination task as compared to Day 4 of the single lever task (Figure 2F). In the plot data, 18F-FDG uptake reached a peak on Day 6 in both the left and right sides of the aDLS (Figure 2J), and the uptake in the left aDLS on Day 6 significantly increased relative to the value of the single lever press, whereas the value in the right aDLS on Day 6 tended to increase relative to that of the single lever press with no significant difference. The plot data showing the unilaterality in the aDLS activation relative to the single lever press are consistent with the image data. On the other hand, the 18F-FDG uptake in the aDLS on Day 6 compared to the value on Day 2 was significantly increased in both sides. Similar observations were made in the activity in the pVLS on Day 10 compared to that on Day 2, as well as in the DMS activity on Day 24 relative to that of the single lever press.

Our analysis of both task-related and learning-dependent activities revealed dynamic changes in striatal subregions during discrimination learning. We investigated the brain regions in which 18F-FDG uptake significantly increased or decreased during the learning processes, applying a statistical significance threshold (p < 0.001, uncorrected) and an extent threshold, by using a voxel-based statistical parametric analysis. In the image data, the voxels showing significant differences between two conditions are visualized on the brain template. The plot data show the amount of 18FFDG uptake in the voxels, which was detected by the voxel-based analysis. The insufficient explanation of the data analysis of PET imaging in the initial manuscript may have led to a misunderstanding regarding the activity in the unilateral or bilateral striatal subregions. Therefore, we have revised the explanation for voxel-based statistical parametric analysis, adding a more detailed description of the thresholds in the text (page 7, lines 143–145) and Methods (page 27, lines 672–675).

Comment 1 (b): From the plots in J, K, and L, it seems that shifts in activity in the different substructures are not unilateral but consistently bilateral, in contrast to what is mentioned in the text. Possibly the text reflects comparisons to the single lever task, and here again, I would emphasize comparing within the same task.

Please see our response to the first comment (a) regarding our explanation of the consistency in the activity of the unilateral or bilateral striatal subregions between the image and plot data. We have also revised the explanation in the corresponding sections of the manuscript, as described above.

Comment 2: In Figure 2, the authors present compelling data that chronic excitotoxic lesions with ibotenic acid in the aDLS, pVLS, and DMS produce differential effects on discrimination learning. However, the significant reduction in success rate of performance happens as early as Day 6 in both IBO groups in both aDLS and pVLS mice. This would seem to agree with conclusions drawn about the role of aDLS in the middle stages of learning in Figure 2, but not the pVLS, which only shows an increased activity during the late stages of the behavior.

Figure 3 shows the behavioral effects of ibotenic acid injections into striatal subregions in rats. For the aDLS injection, we performed two-way repeated ANOVA, which revealed a significant main effect of group or day and a significant interaction of group × day, and added the simple main effects between the treatments to the figure (Figure 3G). We observed significant differences in the success rate mainly at the middle stage of learning. In contrast, for the pVLS injection there was no significant interaction for group × day, although the main effects of group or day was significant by two-way repeated ANOVA (Figure 3H). Consequently, it was unclear as to when exactly the significant reduction occurred. These results indicate that the aDLS and pVLS are necessary for the acquisition of auditory discrimination, and that the aDLS is mainly required for the middle stage. Similar results were observed in the win-shift-win strategy in the aDLS and pVLS (Figures 3J and 3K).

Next, we performed temporal inhibition of neuronal activity in striatal subregions by muscimol treatment in order to examine whether the activity in the subregions is linked with learning processes at different stages. In this experiment, muscimol was injected into the aDLS or pVLS at the middle or late stage, and the resultant effects on the success rate were investigated. The success rate in the muscimol-injected groups into the aDLS significantly decreased at the middle stage, but not at the early and late stages (Figure 4C). In contrast, the rate in the muscimol groups into the pVLS significantly decreased at the late stage, but not at the early or middle stages (Figure 4D). The results indicate that the aDLS and pVLS are mainly involved in the processes at the middle and late stages, respectively, and support the PET imaging data showing the activation of two striatal subregions at the various stages.

We have now provided the results of simple main effects analysis for the aDLS lesion (Figures 3G and 3J) and revised the description of the Results section (page 8, lines 174–178, page 8, lines 186–188, and page 9, line 205-206) and Figure legend (page 44, lines 1000‒1003, and page 44, lines 1010–1013). We have also added the results of simple main effects analysis in Figure 3J.

Comment 3: In Figure 4, the authors show interesting data with transient inactivation of subregions of the striatum with muscimol, validating their findings that the aDLS mediates the middle and the pVLS the late stages of learning, and the function of each area serves different strategies. However, the inference that aDLS inactivation suppresses the WSW strategy "moderately" is not reflected in the formal statistical value p=0.06. While there still may be a subtle effect, the authors would need to revise their conclusions appropriately to reflect the data. In addition, the authors could try a direct comparison between the success rate during muscimol inhibition in the mid-learning session between the aDLS and pVLS-treated groups in Figure 4C (middle) and 4D (middle). If this comparison is not significant, the authors should be careful to claim that inhibition of these two areas differentially affects behavior.

In Figure 4E, aDLS inhibition showed a tendency to reduce slightly win-shift-win strategy at the middle stage (t[14] = 2.038, p = 0.061, unpaired Student’s t-test). In accordance with the reviewer’s comment, we changed the word “moderate” to “subtle” (page 12, line 272).

In the temporal inhibition of the striatal subregions, the aDLS and pVLS experiments (panels C and D, respectively) were conducted separately. Since it is difficult to directly compare the data obtained from different experiments, we did not carry out a direct comparison of the success rate between the aDLS and pVLS injections.

Comment 4: The authors have used in vivo electrophysiological techniques to systematically investigate the roles of the aDLS and the pVLS in discriminatory learning, and have done a thorough analysis of responses with each phase of behavior over the course of learning. This is a commendable and extremely informative dataset and is a strength of the study. However, the result could be better organized following the sequence of events of the behavioral task to give the reader an easier structure to follow. Ideally, this would involve an individual figure to compare the responses in both areas to Cue, Lever Press, Reward Sound, and First Lick (in this order).

We first showed changes in the proportion of event-related neurons during the acquisition phase (Figure S5). Next, we conducted a detailed analysis of the characteristics of aDLS and pVLS neuronal activity. Specifically, we found several types of event-related neurons, including: (1) reward sound-related neurons representing behavioral outcomes in the aDLS; (2) first licking-related neurons showing sustained activity after the reward in the aDLS and pVLS; and (3) cue-onset and cue-response neurons associated with the beginning and ending of a behavior in the pVLS.

Descriptions of the characteristics of event-related neurons according to the sequence of events in a trial, as the reviewer has suggested, is another way to provide an easy structure for understandings on the electrophysiological data. However, we focused on the characteristics of aDLS neurons at the middle stage and pVLS neurons at the late stage of discrimination learning. Therefore, we explained the electrophysiological data based on the order of learning stages rather than the sequence of events in the trial, as described above.

Comment 5: An important conceptual point presented in the study is that the aDLS neurons, with learning, show a reduction in firing rates and responsiveness to the first lick as well as the behavioral outcome, and don't play a role in other task-related events such as cue onset. However, the neuroimaging data in Figure 2 seems to suggest a transient enhancement of aDLS activity in the mid-stage of discriminatory learning, that is not reflected in the electrophysiology data. Is there an explanation for this difference?

In the 18F-FDG PET imaging study, the brain activity in the aDLS reached a peak at the middle stage of the acquisition phase of auditory discrimination (Figure 2J). In the multi-unit electrophysiological recording experiment, the firing activity of the aDLS neuron subpopulations related to the behavioral outcome showed no significant differences among the three stages (Figure 5E), while the proportion of these subpopulations were gradually reduced through the progress of learning stages (Figure 5F). The extent of the firing activity and length of the firing period of other subpopulations showing sustained activation after the reward appeared to show a learning-dependent decrease (Figures 6B and 6C), although the proportion of these subpopulations indicated no correlation with the progress of the learning (Figure 6D). Patterns of the temporal changes in brain activity in striatal subregions across the learning stages did not match completely the time variation in the property or proportion of specific event-related neurons. In our electrophysiological analysis, we identified well-isolated neurons from the striatal subregions during the auditory discrimination task, focusing on putative medium spiny neurons (Figures S4E–S4G). Based on the combinatorial pattern of the tone instruction cue (high tone/H or low tone /L), and lever press (right/R or left/L), we categorized the electrophysiological data into the four trials, including the HR, LL. LR, and HL. We identified HR or LL type neurons showing significant changes in the firing rate related to specific events, such as cue onset, choice response, reward sound, and first licking compared to the baseline firing rate. These neurons were further divided into two groups with increased or decreased activity relative to the baseline firing (Figures S5A and S5B). In the present study, we focused on event-related neurons with increased activity. Because of the analysis limited to neuronal subpopulations related to specific events with the increased activity, it is difficult to fully explain dynamic shifts in the brain activity of striatal subregions dependent on the progress of learning by the time variation of firing activity of individual event-related neurons. The activity of other subpopulations in the striatum may be involved in the shift in brain activity during the learning processes. In addition, recent studies have reported that the activity of glial cells influences the uptake of 18FFDG (Zimmer et al., Nat Neurosci., 2017) and that these cells regulate spike timingdependent plasticity (Valtcheva and Venance, Nat Commun, 2016). Changes in glial cellular activity, through the control of synaptic plasticity, may partly contribute to the pattern formation of learning-dependent shifts in brain activity.

To explain the difference in the time course between the brain activity and the firing activity of specific event-related neurons, we have added the aforementioned information to the Limitations section (pages 21 to 22, lines 512–539).

Comment 6: A significant finding of the study is that CO-HR and CO-LL responses are strikingly obvious in the pVLS, but not in the aDLS, in line with the literature that the posterior (sensory) striatum processes sound. This study also shows that responses to the highfrequency tone indicating a correct right-lever choice increase with learning in contrast to the low-frequency tone responses. To further address whether this difference arises from the task contingency, and not from the frequency representation of the pVLS, an important control would be to switch the cue-response association in a separate group of mice, such that high-frequency tones require a left lever press and vice versa. This would also help tease apart task-evoked responses in the aDLS, as I am given to understand all the recording sites were in the left striatum.

We did not conduct an experiment switching cue-response association in the auditory discrimination task. However, the transient activity of cue onset-related neurons in the pVLS, as the reviewer has suggested, did not appear at the early stage of learning, but was observed in a learning-dependent manner (Figures 7A and S8E). In addition, the cue onset-HR activity showed a slight but notable difference between the HR and LL trials at the middle and late stages (Figure 7B), but there was no difference in activity in the HL and LR incorrect trials at the corresponding stages (Wilcoxon signed rank test; early, p = 0.375, middle, p = 0.931, and late, p = 0.668). These results suggest that the activity of cue onset-related neurons in the pVLS is associated with the stimulus and response association (task contingency) rather than the tone frequency.

Reviewer #1 (Recommendations For The Authors):

Minor comment 1: The readability and appeal of this study would be improved by explaining the various neuronal response types, and task-related events in slightly more detail in the results section, and minimizing the use of non-standard abbreviations wherever possible.

As suggested, we have replaced the abbreviations related to electrophysiological events (CO, CR, RS, and FL) with the original terms, and improved the explanation for neuronal response types and event-related neurons.

Minor comment 2: It would be helpful to label DLS and VLS recordings more clearly on the figures instead of only in the figure caption.

Thank you for pointing this out. The terms “aDLS” and “pVLS” have now been added to the panels showing firing pattern of neurons: “aDLS” in Figures 5D, 6A, S6A, S7A, S8A, S8B. S8C, and S8D; and “pVLS” in Figures 6F, 7A, 7D, S6D, S6E, S7F, S8E, and S8F.

Minor comment 3: The authors suggest that aDLS HR- and LL- neurons are more sensitive to the behavioral outcome than those in pVLS (Fig 5 and S5). However, their conclusions are based on sample sizes as low as n=3 for each response type.

We identified event-related neurons from single neurons detected in both the aDLS and pVLS using the same criteria. In the pVLS, we found a small number of neurons that increased their activity during the period when the reward sound is presented (Figures S6D and S6E) (6, 4, and 17 HR type neurons at the early, middle, and late stages, respectively; 3, 5, and 15 LL type neurons at the early, middle, and late stages, respectively). The number of LL type neurons at the early stage was particularly lower, as the reviewer has suggested. However, when we plotted the firing rates of these neurons around the event, their activity did not reflect behavioral outcome. In the aDLS, we detected a large number of reward sound-related neurons representing behavioral outcome (Figures 5 and S6A) (43, 37, and 44 HR type neurons at the early, middle, and late stages, respectively; 49, 62, and 59 LL type neurons at the early, middle, and late stages, respectively). These observations suggest that aDLS neurons are more sensitive to behavioral outcomes than pVLS neurons.

Minor comment 4: Typo in Figure 4C and D, right plots, y-axis label: "subtracted".

The typographic errors in Figures 4C–4H have now been corrected to “subtracted”.

Reviewer #2 (Public Reviews):

The study by Setogawa et al. aims to understand the role that different striatal subregions belonging to parallel brain circuits have in associative learning and discrimination learning (S-O-R and S-R tasks). Strengths of the study are the use of multiple methodologies to measure and manipulate brain activity in rats, from microPET imaging to excitotoxic lesions and multielectrode recordings across anterior dorsolateral (aDLS), posterior ventral lateral (pVLS)and dorsomedial (DMS) striatum. The main conclusions are that the aDLS promotes stimulus-response association and suppresses response-outcome associations. The pVLS is engaged in the formation and maintenance of the stimulus-response association. There is a lot of work done and some interesting findings however, the manuscript can be improved by clarifying the presentation and reasoning. The inclusion of important controls will enhance the rigor of the data interpretation and conclusions.

We appreciate the reviewer’s valuable feedback, which has been beneficial in our endeavor to improve our manuscript. In response to the comments, we have revised the description of the experimental methods and underlying rationale, as well as the Results section. We have also provided additional data for some of the experiments that support the conclusions. For more details, please refer to the responses to each comment, included below.

Reviewer #2 (Recommendations For The Authors):

Comment 1: Generally, the manuscript is hard to read because of the cumbersome sentence structure, overuse of poorly defined acronyms, and lack of clarity on the methods used.

According to the following comments (a)–(d), we have revised the corresponding text in the manuscript to clarify the sentence structure, definitions of terms, and methodology.

Comment 1 (a): For example, the single lever task used as a control for the auditory discrimination task could be introduced better, explaining the reasoning and the strategy for subtracting it from the images obtained during the discrimination phase at the start of the section.

We analyzed task-related activity by comparing 18F-FDG uptake on Days 2, 6, 10, or 24 of auditory discrimination task with that on Day 4 of the single lever press task. This task was used as a control that does not require a decision process based on the auditory stimulus. For clarification, we have provided a more detailed explanation of the flow of the single lever press task used in the PET experiment, including the rationale for employing this task as a control (page 6, lines 129–135). We have also revised the explanation of voxel-based statistical parametric analysis, adding a more detailed description of the thresholds (page 7, lines 143–145).

Comment 1 (b): Another example is that important methodological information is buried deep in the text and complicates the interpretation of the results.

We have revised the following sentences in the manuscript in order to provide clearer methodological information.

(1) As described above, explanations for the single lever task (page 6, lines 129–135) and voxel-based statistical parametric analysis were added (page 7, lines 143–145).

(2) Definition of the early, middle, and late stages were described in the initial behavioral experiment (page 6, lines 113–119).

(3) Abbreviations related to behavioral strategies (WSW and LSL) and electrophysiological events (CO, CR, RS, and FL) were replaced with the original terms.

Comment 1 (c): The specie being studied is not stated in the abstract, nor the introduction, and only in the middle of the result section. Please include the specie in the abstract and the first part of the result also for clarity.

We included the name of the species (rats) in the Abstract (page 3, line 47), at the end of the Introduction (page 5, lines 87–88) and at the beginning of the Results (page 5, line 109).

Comment 1 (d): The last part of the intro is copied/pasted from the abstract. Please revise.

The last part of the Introduction was revised accordingly (page 5, lines 97–104).

Comment 2: The glucose microPET imaging is carried out 30 mins after the rats performed the task and it is expected to capture activation during the task. Is this correct? This assumption has to be validated with an experiment, which is a control showing a validation of the microPET approach used, and this way can report activation of brain areas during the task completed 20-30 minutes before. For example, V1 or A1 would be a control that we would expect to be activated during the task.

Our PET experiment was conducted in accordance with previously established methods (Cui et al, Neuroimage, 2015), where rats received intravenous administration of 18FFDG solution just before the start of the behavioral session, which lasted for 30 min. The 18F-FDG uptake in the brain starts immediately and reaches the maximum level until 30 min after the administration, and the level is kept at least for 1 h (Mizuma et al., J Nucl Med, 2010). The rats were returned to their home cages, and a 30-min PET scan started 25 min after the session. The start time of the scan was chosen to allow for sufficient reduction of 18F radioactivity in arterial blood to increase the S/N ratio of the radioactivity (Mizuma et al., J Nucl Med, 2010). As shown in Table S1, we confirmed that the brain activity in the medial geniculate body (auditory thalamus) was increased on Days 6 and 10 in the acquisition phase, although the activity in the auditory cortex was not changed, which is consistent with the results of a previous study reporting that the auditory cortex does not show the causality for the pure-tone discrimination task (Gimenez et al., J Neurophysiol., 2015).

Comment 3: Why are Days 2, 6, 10, and 13 chosen and compared for the behavior? Why aren't these the same days chosen in the other part of the study? It is unclear why authors focused on these days and why the focus changed later.

We conducted daily training of the discrimination task. The success rate reached a plateau on Day 13 and was maintained until Day 24 (Figure 1B). Based on these results, we categorized the learning processes into the acquisition and learned phases, and then divided the acquisition phase into the early (< 60%), middle (60–80%), and late (> 80%) stages. In the PET experiment, we selected Days 2, 6, and 10 as the representatives of each stage during the acquisition phase. In addition, we also selected Day 24 for the learned phase. However, no scan was performed on Day 13 due to the transition between the two phases.

Comment 4: (A) Is the learning and acquisition of the single lever press and discrimination task completed by day 4? Or are rats still learning? The authors claimed no changes in DMS activity between single lever press & discrimination, and therefore DMS isn't involved in learning. But to make this claim we should have measures that the learning has already happened, which I am not sure have been provided. (B) On this same point, the DMS activity is elevated on Day 4 of a single lever press compared to the aDLS and pVLS. So is it possible that the activity in DMS was already elevated on Day 4 of single lever press training? Especially given that DMS is supposedly involved in goal-directed behavior?

(A) In the single lever press task, the number of lever presses plateaued on Day 2 (Figure 1C). In addition, we analyzed response time and its variability, which plateaued from Day 3 and Day 2, respectively (see Author response image 1). These results indicate that the learning in the task was completed by Day 4. In the auditory discrimination task, Day 4 corresponded to the transition period from the early-tomiddle stages of the acquisition phase, suggesting that learning was still progressing.

In the imaging analysis, we examined task-related activity by comparing 18F-FDG uptake on either day of the discrimination task with that on Day 4 of the single lever press task, and did not find any changes in the brain activity in the DMS. In addition, we investigated learning-related activity, and the DMS activity did not change during acquisition phase. These results suggest that the DMS is not involved in the acquisition phase of learning. Furthermore, comparisons between Days 10 and Day 24 showed a decrease in DMS activity during the learned phase, suggesting that DMS activity was downregulated during the learned phase. In addition, chronic lesion in the DMS indicated that the success rate in the discrimination task was comparable between the control and lesioned groups (Figure 3I), whereas the response time lengthened throughout the learning in the lesioned group compared to the controls (Figure S1C). These results support our notion that the DMS contributes to the execution, but not learning, of discriminative behavior (Figure 3I and S1C).

Performance of single lever press task conducted before auditory discrimination task.

(A) Number of lever presses. (B) Response time (Kruskal-Wallis test, χ2 = 38.063, p = 2.7 × 10-8, post hoc Tukey–Kramer test, p = 0.047 for Day 1 vs. Day 2; p = 2.3 × 10-7 for Day 1 vs. Day 3; and p = 4.0 × 10-6 for Day 1 vs. Day 4; p = 0.019 for Day 2 vs. Day 3; p = 0.082 for Day 2 vs. Day 4; p = 0.951 for Day 3 vs. Day 4). (C) Response time variability (Kruskal-Wallis test, χ2 = 28.929, p = 2.3 × -6, post hoc Tukey–Kramer test, p = 0.077 for Day 1 vs. Day 2; p = 5.7 × 10-6 for Day 1 vs. Day 3; and p = 1.3 × 10-4 for Day 1 vs. Day 4; p = 0.060 for Day 2 vs. Day 3; p = 0.253 for Day 2 vs. Day 4; p = 0.912 for Day 3 vs. Day 4). Data obtained from the task shown in Figure 2C are plotted as the median and quartiles with the maximal and minimal values. *p < 0.05, **p < 0.01, and ***p < 0.001.

(B) We compared 18F-FDG uptakes among striatal subregions on Day 4 of the single lever press task (334.8 ± 2.86, 299.0 ± 1.71, and 336.8 ± 2.18 for the aDLS, pVLS, and DMS, respectively; one-way ANOVA, F[2,41] = 104.767, p = 2.1 × 10-16). The uptake was comparable between the aDLS and DMS (post hoc Tukey-Kramer test, p = 0.058), but it was significantly lower in the pVLS compared to either of the other two subregions (post hoc Tukey-Kramer test, aDLS vs. pVLS, p = 5.1 × 10-9, post hoc Tukey-Kramer test, pVLS vs. DMS, p = 5.1 × 10-9). However, since we did not measure the brain activity in the single lever task outside of Day 4, it is unclear whether there was an increase in DMS activity during the acquisition of the task. Similarly, since we did not confirm the behavioral modes, which include goal-directed and habitual actions, it is difficult to conclude that the lever presses in the task were controlled by the goaldirected mode. However, our chronic lesion experiment suggests that the DMS is involved in the execution of discrimination behavior (Figure S1C). A clearer understanding of the DMS function in discrimination learning is an important challenge in the future.

Comment 5: It seems like the procedure of microPET imaging affects performance on the task. The anesthesia used maybe? Figures 2C and D show evidence that the behavior was negatively affected on the days on which microPET imaging was performed after the training. Can the author clarify/comment?

Isoflurane anesthesia may slightly reduce behavioral performance. We carried out anesthesia (median [interquartile range]: 6 [5–8] min) during the insertion of the catheter for FDG injection, and set a recovery period of at least 2 h until the beginning of the behavioral session, to minimize the impact of anesthesia. The performances in Figure 2E were similar to those in the intact rats (compared to Figures 1C–1F), suggesting that the procedure for PET scans does not affect the acquisition of discrimination.

We have added detailed information on the isoflurane anesthesia to the Methods section (page 26, lines 649–653).

Comment 6: More on clarity. Section 3 of the results (muscimol inactivation) refers a lot to "the behavioral strategies" without really clarifying what these are - are they referring to WSW / LSL (which also could use a better introduction) or goal-directed/habitual or stimulus-response/stimulus-outcome?

The dorsal striatum is involved in both behavioral strategies based on stimulus-response association and the response-outcome association during instrumental learning. To assess the impact of striatal lesions on the behavioral strategies, we analyzed the proportion of response attributed to two strategies in all responses of each session. One is the “win-shift-win” strategy, which is considered to reflect the behavioral strategy based on the stimulus-response association. In this strategy, after a correct response in the previous trial, the rats press the opposite lever in the current trial in response to a shift of the instruction cue, resulting in the correct response. Another strategy is the “lose-shift-lose” strategy, which is considered to appear as a consequence of the behavioral strategy based on the response-outcome association. In this strategy, after an error response in the previous trial, the rats press the opposite lever in the current trial despite a shift of the instruction cue, leading to another error response.

We have revised the explanations of the behavioral strategies in the section of the Results section (page 9, lines 192–201).

Comment 7: Related to WSW / LSL needing a better introduction, on lines 192/193 authors describe a result where they saw the WSW and LSL strategies increase and decrease, respectively, in saline-injected mice. Is the change in performance expected or an undesired effect of the saline injection? This is not clear now and it should be clarified.

The explanations of the win-shift-win and lose-shift-lose strategies have been revised in the Results section on excitotoxic lesion experiment (page 9, lines 192–201) as described in our response to Comment 6. Win-shift-win is an indicator of correct responses, while lose-shift-lose indicates errors. Therefore, win-shift-win is predicted to increase, and lose-shift-lose decrease, as discrimination learning progresses. Indeed, in the results of the behavioral experiments, shown in Figure 1, both indicators change in a similar pattern to those in the results of the lesion experiments (Figure 3).

We have added the explanation of the proportions of both strategies in intact rats (page 9, lines 203–204) with a supplementary figure (Figure S2) and accompanying legend (page 56, lines 1173–1177).

Comment 8: Muscimol experiments - two questions/comments. How often do rats receive muscimol?

In this section, muscimol is given on day 2 and on days after the animals hit a 60% or 80% success rate. Can the authors provide a mean and SEM for when are those injections?

The first injection was conducted on Day 2 to target the early stage. The second and third injections were conducted on the days after the success rate had reached 60% and 80% for the first time through the training, respectively, to target the middle and late stage. respectively. These conditions are described in the Results (page 10, lines 234– 237) and Methods (page 26, lines 633–636). The mean and s.e.m. of the injection day at the middle and late stages were not significantly different between the saline and muscimol-injected groups into the aDLS (see Author response image 2A) and pVLS (see Author response image 2B).

Author response image 2.

Injection days during auditory discrimination learning.

Injections with saline (SAL) and muscimol (MUS) into the aDLS (A) or pVLS (B) were performed after the success rate had reached 60% (middle stage) and 80% (late stage) for the first time through the training, respectively (A, Wilcoxon signed rank test, middle, Z = 65, p = 0.772, late, Z = 56.5, p = 0.242 for the aDLS; B, Wilcoxon signed rank test, middle, Z = 39, p = 1.000, late, Z = 43, p = 0.587). Data are indicated as the median and quartiles with the maximal and minimal values.

Comment 9: Muscimol experiments. Can the authors comment on the effects on performance vs learning? What happens on the days after Muscimol? Does performance bounce back or is it still impaired?

We conducted a transient inhibition experiment with muscimol to examine whether the neuronal activity in the striatal subregions is linked with the processes at different stages. In this experiment, to lower the possibility that compensation of learning may occur during a session after the muscimol injection (Day N), we limited the session time to 15 min (45 trials) and evaluated the impact of the injection on the success rate at specific stages. The success rate in the muscimol-injected groups into the aDLS significantly decreased at the middle stage compared to the corresponding salineinjected groups, but not at the early and late stages (Figure 4C), and the rate in the muscimol groups into the pVLS significantly decreased at the late stage compared with the respective saline groups, but not at the early and middle stages (Figure 4D). Our results demonstrated that the aDLS and pVLS mainly function at the middle and late stages of the auditory discrimination task, respectively.

In addition, we here reply to comment 10 as for the comparison of success rates before (Day N-1) and after (Day N+1) the injections (see Author response image 3). We focused on two injections into the aDLS at the middle stage and into the pVLS at the late stage, in which the rate was reduced soon after the muscimol injection on Day N. The success rate for the two injections showed no significant main effect regarding group (saline/muscimol) or day (Days N-1/N+1) and no significant interactions for group × day. Moreover, the success rate was not significantly increased on Day N+1 as compared to Day N-1, even in the saline-injected control group, probably because of the limited session time soon after the injection. Therefore, we consider that it was difficult to define the effects of drug injection on the learning of auditory discrimination in our behavioral protocol for the transient inhibition experiment, and that the reduced rates observed in the muscimol-injected group on Day N mostly reflect the impacts of muscimol at least partly on the performance of discriminative behavior.

Comparison of success rate between days before (Day N1) and after (Day N+1) the injections into striatal subregions.

Success rate in the saline (SAL)- and muscimol (MUS)-injected groups into the aDLS (A) or pVLS (B) at the early, middle, and late stages of auditory discrimination learning (two-way repeated ANOVA; early, day, F[1,14] = 5.266, p = 0.038, group, F[1,14] = 0.276, p = 0.608, day × group, F[1,14] = 0.118, p = 0.736; middle, day, F[1,14] = 4.110, p = 0.062, group, F[1,14] = 0.056, p = 0.816, day × group, F[1,14] = 1.150, p = 0.302; late, day, F[1,14] = 6.408, p = 0.024, group, F[1,14] = 0.229, p = 0.640, day × group, F[1,14] = 1.277, p = 0.278 for the aDLS; and early, day, F[1,10] = 0.115, p = 0.746, group, F[1,10] = 2.414, p = 0.151, day × group, F[1,10] = 0.157, p = 0.700; middle, day, F[1,10] = 0.278, p = 0.610, group, F[1,10] = 0.511, p = 0.491, day × group, F[1,10] = 4.144, p = 0.069; late, day, F[1,10] = 0.151, p = 0.705, group, F[1,10] = 0.719, p = 0.416, day × group, F[1,10] = 0.717, p = 0.417 for the pVLS). Data are indicated as the mean ± s.e.m.

Comment 10: Muscimol data has a pair before and after, can the authors show this comparison at early, middle, and late training? Not just the subtraction.

The comparison of success rates before and after drug injection is shown in Author response image 3.

Comment 11: Ephys recordings. These are complex figures and include a large number of acronyms. It would help to define them again and help the reader through these figures so the reader can focus on understanding the finding more than the figure presentation.

We replaced the abbreviations related to electrophysiological events (CO, CR, RS, and FL) with the original terms, and improved the explanation in the text and figures.

Comment 12: Figure 7B/E - on correct trials, they see a difference in the cue response to high tone / low tone but no difference in the choice. This is the one that seemed like a topography issue.

The transient activity of cue onset-related neurons in the pVLS did not appear at the early stage of learning, but was observed in a learning-dependent manner (Figures 7A and S8E). In addition, the cue onset-HR activity showed a slight but notable difference between the HR and LL trials at the middle and late stages (Figure 7B), whereas there was no difference between activities in the HL and LR incorrect trials at the corresponding stages (Wilcoxon signed rank test; early, p = 0.375, middle, p = 0.931, and late, p = 0.668). These results suggest that the cue onset-related neurons in the pVLS represents the stimulus and response association (task contingency) rather than the topography of tone frequency.

Comment 13: Animals were normally trained for 60 minutes but on muscimol days only trained for 15 mins. On PET days only trained for 30 minutes. Ephys sessions were 60 mins. Is this correct? Why?

We determined the session time for each experiment by considering both technical and behavioral aspects. In the initial behavioral experiment, the session time was set to 60 min per day. Under this condition, the rats acquired the discrimination learning within 13 days. In the imaging experiment, the session without a PET scan was conducted for 60 min, while the session with a PET scan was carried out for 30 min as described previously (Cui et al, Neuroimage, 2015). This time schedule produced a learning curve similar to that of the initial behavioral experiment. In the transient inhibition experiment, the sessions without drug injections lasted for 60 min. As described in our response to the comment 2, the time of the session soon after the injection was limited to 15 min to lower the possibility of compensation of learning during the session. In the chronic lesion and electrophysiological experiments, all sessions were conducted for 60 min, corresponding to the initial experiment.

References

Mizuma, H., Shukuri, M., Hayashi, T., Watanabe, Y. & Onoe, H. Establishment of in vivo brain imaging method in conscious mice. Journal of Nuclear Medicine 51, 10681075 (2010).

Cui, Y., et al. A voxel-based analysis of brain activity in high-order trigeminal pathway in the rat induced by cortical spreading depression. Neuroimage 108, 17-22 (2015).

Zimmer, E.R., et al. [18 F] FDG PET signal is driven by astroglial glutamate transport. Nat Neurosci 20, 393-395 (2017).

Valtcheva, S. & Venance, L. Astrocytes gate Hebbian synaptic plasticity in the striatum. Nature communications 7, 13845 (2016).

Gimenez T.L., Lorenc M., Jaramillo S. Adaptive categorization of sound frequency does not require the auditory cortex in rats. J Neurophysiol 114:1137-1145 (2015).

Version 2

Abstract

Abstract

The striatum, the central hub of cortico-basal ganglia loops, contains functionally heterogeneous subregions distinguished by the topographic patterns of structural connectivity. These subregions mediate various processes of procedural learning. However, it remains unclear when and how striatal subregions engage in the acquisition of sensory stimulus-based decision-making. A neuroimaging of regional brain activity shows that the anterior dorsolateral striatum (aDLS) and posterior ventrolateral striatum (pVLS) in rats are activated in a different temporal pattern during the acquisition phase of auditory discrimination. Chronic and transient pharmacologic manipulations show that the aDLS promotes the behavioral strategy driven by the stimulus-response association while suppressing that by the response-outcome association, and that the pVLS contributes to forming and maintaining the stimulus-response strategy. Electrophysiological recording indicates that subpopulations of aDLS neurons predominantly represent the outcome of specific behaviors at the initial period of discrimination learning, and that pVLS subpopulations encode the beginning and ending of each behavior according to the progress of learning. In addition, other subpopulations of striatal neurons indicate sustained activation after obtaining reward with distinct patterns reflecting the stimulus-response associations. Our findings demonstrate that aDLS and pVLS neurons integrate the new learning of auditory discrimination in spatiotemporally and functionally different manners.

Introduction

The neural network connecting the cerebral cortex, basal ganglia, and thalamus forms functionally heterogeneous loops distinguished by the topographic patterns of connectivity1,2, and is implicated in various learning processes by the integration of cognitive, sensorimotor, and reward information3,4. The striatum is a key node in the cortico-basal ganglia loops, and its segregated subregions are considered to mediate procedural learning, such as motor skills and instrumental behavior, through parallel processing via different loops5–7. Previous studies in humans and non-human primates have reported that the neural activity within the striatum shifts from the associative subregions (caudate and anterior putamen) to the sensorimotor subregions (posterior putamen) as motor learning progresses8–11. In rodents, an instrumental behavior initiates as goal-directed actions and then transits to stimulus-response habits12,13, requiring the dorsomedial striatum (DMS; a homologue to the caudate in primates) and the dorsolateral striatum (DLS; a homologue to the putamen in primates), respectively14,15. Based on these results, a model of procedural learning has been proposed, in which the functional dominance changes from the associative to sensorimotor subregions in the striatum during the learning phases16.

In contrast, deviations from the prior model are evident in other behavioral tasks on decision-making based on external sensory cues. For example, the neural activity in the caudate in humans is associated with learning of a visuomotor association task and the execution of learned behavior17,18. The tail of the caudate in non-human primates contributes to the automatic, habitual behavior in decision-making task using complex visual stimuli19. In rodents, the DLS is involved in the acquisiton and execution of the discrimination task with visual and auditory cues, wheras the DMS madiates the performance of learned behavior20,21. Moreover, other striatal subregions, such as the posterior striatum and ventrolateral striatum in rodents are also engaged in the auditory discrimination22–24. Based on these results, a hypothesis has arisen that the acquisition of sensory cue-based decision-making requires wide-range spatiotemporal processing within the striatum. However, the neural mechanism by which the functional circuits linked to striatal subregions change throughout learning processes is not yet fully understood.

To investigate a large-scale striatal reorganization in accordance with the acquisition of external cue-dependent decisions, we conducted a two-alternative auditory discrimination task with rats25. Using a small-animal neuroimaging technique, we found that the anterior dorsolateral striatum (aDLS) and posterior ventrolateral striatum (pVLS) are activated to the highest level at the middle and late stages during the acquisition phase of auditory discrimination, respectively. Then, chronic pharmacologic manipulations confirmed that the aDLS and pVLS are necessary for discrimination learning, and the transient manipulations also indicated the function of these subregions in the learning at the corresponding stages. The resultant effects of the manipulations on the behavioral strategy showed that the aDLS promotes the strategy based on the stimulus-response association, while suppressing that based on response-outcome associations. Additionally, the pVLS engages in the formation and maintenance of the stimulus-response association strategy.

Electrophysiological data of striatal neurons indicated the presence of aDLS subpopulations, mainly representing the outcome of specific behaviors at the initial period of auditory discrimination and pVLS subpopulations encoding the beginning and ending of behavior in association with the development of the discrimination. There were other subpopulations in the aDLS and pVLS showing long-term increased activity following reward with different patterns of the combination between the stimulus and response. These findings revealed that the two striatal subregions are involved in the acquisition of auditory discrimination through distinct spatiotemporal and functional fashions, challenging the prior model which proposed a transition from the associative to sensorimotor striatal subregions during learning.

Results

Leaning processes of auditory discrimination

We employed a two-alternative auditory discrimination task, in which rats are required to make new associations between a tone instruction cue (2 or 10 kHz) and a response (pressing the left or right lever)25, setting the interval between the stimulus presentation and lever insertion to 3 s (Figure 1A). The success rate gradually increased, reaching a plateau around Day 13, which persisted through Day 24 (Figure 1B). Based on changes observed in the success rate, we categorized the learning processes into two phases: the acquisition phase, which lasted until Day 13; and the learned phase, which began after Day 13 and then continued at least until Day 24. In addition, we divided the acquisition phase into three stages depending on the success rate that included the early (< 60%), middle (60–80%), and late (80–100%) stages. We chose Days 2, 6, and 10 at the early, middle, and late stages in the acquisition phase, respectively, as well as Days 13 and 24 in the learned phase for statistical comparison of the success rate. There were significant differences between Days 2 and 6, Days 6 and 10, as well as Days 10 and 13, showing no significant difference between Days 13 and 24 (Figure 1C). In contrast, the response time and omission rate did not vary among the training days (Figures 1D and 1E), and the variation of response bias decreased along with the progress of auditory discrimination (Figure 1F).

Behavioral performance of two-alternative auditory discrimination learning.

(A) Schematic diagram of the auditory discrimination task. Each trial started by presentation of a tone instruction cue with the high (10 kHz) or low (2 kHz) frequency. Three seconds later, the room light was illuminated, and two retractable levers were inserted at the same time. The rats were required to press the right and left levers in response to the high and low tones, respectively. (B) Learning curve of the auditory discrimination in intact rats (n = 14). Inset indicates the cumulative curve of number achieved the success rate of more than 90%. (C) Success rate at Days 2, 6, 10, 13, and 24 (one-way repeated ANOVA, F[1.897,24.659] = 106.706, p = 1.1 × 10-12, post hoc Bonferroni test, Day 2 vs. Day 6, p = 0.001, Day 6 vs. Day 10, p = 3.6 × 10-4, Day 10 vs. Day 13, p = 0.040, and Day 13 vs. Day 24, p = 0.080). (D) Response time (one-way repeated ANOVA, F[2.203,28.642] = 2.386, p = 0.105). (E) Omission rate (F[4,52] = 0.699, p = 0.596). (F) Response bias (one-way repeated ANOVA, F[1.129,14.674] = 0.250, p = 0.653). Data are indicated as the mean ± s.e.m., and individual data are overlaid. *p < 0.05, **p < 0.01, and ***p < 0.001.

Distinct brain activity patterns in striatal subregions during learning processes

To investigate dynamic changes in regional brain activity in the entire striatum during auditory discrimination in the same animals, we conducted the positron emission tomography (PET) for small animals with 2-deoxy-2-[18F]fluoro-D-glucose (18F-FDG), which measures cerebral glucose metabolism correlated to brain activity26–28 (Figure 2A). Before the auditory discrimination task, we performed the single lever press task, in which a lever pseudorandomly presented on either the left or right side, and the rats learned to press a lever associated with receiving a reward. This task was used as a control that does not require the decision process based on the auditory stimulus. 18F-FDG-PET scanning was carried out on Day 4 of the single lever press task, and then during a series of training sessions including the acquisition (Days 2, 6, and 10) and learned (Day 24) phases (Figure 2B). The average number of lever presses in the single lever press task exceeded 80 times (81.6 ± 1.3, mean ± s.e.m.) on Day 2 and maintained that level until Day 4 (Figure 2C). In the following auditory discrimination task, the success rate gradually increased along with the training days (Figure 2D), showing a significant difference among sessions in which we performed the 18F-FDG-PET scanning (Figure 2E). The response time and omission rate were consistent among the days, and the variation of response bias became smaller as the task progresses (Figure 2E). These performances were similar to those in the intact rats (compared to Figures 1C–1F), suggesting that the procedure for 18F-FDG-PET scans does not affect the acquisition of discrimination.

Dissociable brain activity patterns during auditory discrimination learning among three striatal subregions.

(A) Schematic illustration of an awake rat received intravenously 18F-FDG injection through an indwelling catheter attached to the tail. The rat was conducted for the behavioral experiment and then used for microPET imaging. (B) Schedule for the behavioral training and 18F-FDG-PET scan. After 18F-FDG injection, rats (n = 14 rats) were subjected to the behavioral experiment (30 min) and returned to the home cage (15 min). Ten min before the PET scan started, the rats were anesthetized and then the scan was started (30 min). (C and D) Learning curves of the single lever press task (C) and auditory discrimination task (D). Arrowheads indicate the PET scan days. (E) Behavioral performance on the scan days during the auditory discrimination. Success rate (one-way repeated ANOVA, F[3,39] = 125.012, p = 4.8 × 10-20, post hoc Bonferroni test; Day 2 vs. Day 6, p = 5.0 × 10-7; Day 6 vs. Day 10, p = 0.001; Day 10 vs. Day 24, p = 0.003), response time (one-way repeated ANOVA, F[3,39] = 0.156, p = 0.926), omission rate (one-way repeated ANOVA, F[3,39] = 1.559, p = 0.215), and response bias (one-way repeated ANOVA, F[1.779,23.130] = 17.734, p = 3.6 × 10-5) are shown. (F) Representative images of coronal sections that compared the brain activity in the single lever press task with the activity on Day 2, 6, 10, or 24 in the discrimination task. Left panels show schematic illustrations of striatal subregions. (G) Horizontal (top) and coronal (bottom) images of striatal activation areas shown on Day 10 vs. single lever in (F). (H) Representative images of coronal section that compared the brain activity on Day 2 with that on Day 6, 10, or 24 in the discrimination task. Color bars indicate the T-values, and a value of 3.8 was used as the threshold corresponding to the uncorrected threshold (p < 0.001). (I) Schematic pictures showing voxel of interests for the aDLS, pVLS, and DMS (green, left hemisphere; and purple, right hemisphere). (J-L) Regional 18F-FDG uptakes in the aDLS (J, one-way repeated ANOVA, left aDLS, F[4,52] = 10.322, p = 3.0 × 10-6, post hoc Bonferroni test; single lever vs. Day 6, p = 0.016; Day 2 vs. Day 6, p = 6.5 × 10-5; Day 2 vs. Day 10, p = 0.019; Day 2 vs. Day 24, p = 0.017; Day 6 vs. Day 24, p = 0.041; right aDLS, F[4,52] = 7.462, p = 7.9 × 10-5, post hoc Bonferroni test, Day 2 vs. Day 6, p = 2.9 × 10-4; Day 2 vs. Day 10, p = 0.036; Day 6 vs. Day 24, p = 0.005), pVLS (K, one-way repeated ANOVA, left pVLS, F[2.368,30.784] = 4.152, p = 0.020, post hoc Bonferroni test; single lever vs. Day 10, p = 1.8 × 10-5; right pVLS, F[4,52] = 5.995, p = 4.8 × 10-4, post hoc Bonferroni test, single lever vs. Day 10, p = 0.001; Day 2 vs. Day 10, p = 0.011; Day 6 vs. Day 10, p = 0.032), and DMS (L, one-way repeated ANOVA, left DMS, F[4,52] = 12.836, p = 2.4 × 10-7, post hoc Bonferroni test; Day 2 vs. Day 24, p = 1.8 × 10-5; Day 6 vs. Day 24, p = 4.8 × 10-4; Day 10 vs. Day 24, p = 5.5 × 10-5; right DMS, F[4,52] = 10.717, p = 2.0 × 10-6, post hoc Bonferroni test; single lever vs. Day 24, p = 0.036; Day 2 vs. Day 24, p = 0.011; Day 6 vs. Day 24, p = 2.6 × 10-4; Day 10 vs. Day 24, p = 0.011). Arrowheads indicate the day with the most activations throughout the learning process. Data are indicated as the mean ± s.e.m., and individual data are overlaid. The anteroposterior coordinates from bregma (mm) are shown (F–I). Scale bar; 2 mm (I). *p < 0.05, **p < 0.01, and ***p < 0.001.

We investigated the brain regions, in which 18F-FDG uptake significantly increased or decreased during the learning processes, applying a statistical significance threshold (p < 0.001, uncorrected) and an extent threshold, by using a voxel-based statistical parametric analysis26. We first analyzed task-related brain activity by comparing 18F-FDG uptake on either day of the discrimination task with that on Day 4 of the single lever press task (summarized in Table S1). In the striatal subregions, there were significant increases of the activity in the unilateral aDLS and the bilateral pVLS on Days 6 and 10, respectively (Figures 2F and 2G). However, no significant change was observed in the DMS during the acquisition phase except a unilateral decrease on Day 24 (Figure 2F). Next, to evaluate the activity related to the progress of learning, 18F-FDG uptake on Days 6, 10, and 24 was compared with that on Day 2. The learning-dependent brain activity was significantly increased in the bilateral aDLS and the unilateral pVLS on Days 6 and 10, respectively (Figure 2H), whereas the activity decreased in the bilateral DMS on Day 24 (Figure 2H). This decrease was also detected when 18F-FDG uptakes were compared between Days 10 and 24 (Figure 2H), suggesting the down-regulation of the DMS activity in the learned phase.

To quantitatively validate changes in the brain activity in striatal subregions, we analyzed the amount of 18F-FDG uptake in the voxel of interest of these subregions (Figure 2I). The 18F-FDG uptake in the bilateral aDLS reached a peak on Day 6, and it decreased during the subsequent days (Figure 2J), whereas the uptake in the bilateral pVLS gradually increased along with the progress of learning, showing the maximal value on Day 10 (Figure 2K). The uptake in the DMS did not alter during the acquisition phase, and it decreased on Day 24 (Figure 2L). These results suggest that the aDLS and pVLS contribute to learning processes in a different temporal pattern during the acquisition phase, whereas the DMS may be engaged to the behavior in the learned phase.

Excitotoxic lesion of the aDLS and pVLS, but not the DMS, disrupts the acquisition of auditory discrimination

To study whether the changed activity of striatal subregions is involved in learning processes, rats received a bilateral injection of solution containing ibotenic acid (IBO; 8 mg/mL, 0.3 μL/site) or phosphate buffered saline (PBS) into the subregions, and we conducted for the behavioral study. The range of IBO lesion in the subregions was analyzed by immunostaining with a neuronal marker NeuN after the behavioral tests (Figures 3A–3C). The number of single lever presses increased similarly along with the training in both the PBS- and IBO-injected groups into the aDLS, pVLS, or DMS along with the training (Figures 3D–3F). For the aDLS injections, increases in success rate were impaired during the acquisition phase in the IBO group, compared to the PBS group, and the impairments were mainly observed at the middle stage (Figures 3G). For the pVLS injection, success rate impairments were observed in the IBO group showed compared to the PBS group, although the period showing a significant difference was unclear (Figure 3H). The response time in the aDLS-lesioned group did not alter during the discrimination training, whereas that in the pVLS-lesioned group was continuously prolonged (Figure. S1A and S1B), and the omission rate and response bias were similar between the lesioned and the corresponding control groups (Figures S1D/S1G for aDLS lesion and Figures S1E/S1H for pVLS lesion). In contrast, for the DMS injection the success rate in the discrimination task was comparable between the PBS and IBO groups (Figure 3I). The DMS lesion lengthened the response time through the learning (Figure S1C), whereas it did not alter the omission rate and response bias (Figures S1F and S1I). These results indicate that the aDLS and pVLS are necessary for the acquisition of auditory discrimination, and that the aDLS appears to be mainly involved in the learning process at the middle stage. The results also reveal that the DMS contributes to the execution, but not learning, of discriminative behavior.

Impacts of excitotoxic lesion of striatal subregions on the acquisition of auditory discrimination.

Rats were given intracranial injection of PBS or IBO solution into the aDLS (n = 8 for each injection), pVLS (n = 8 for each injection), or DMS (n = 7 for PBS injection, and n = 6 for IBO injection) before start of the single lever press task. (A-C) Representative images of NeuN immunostaining and individual schematic illustrations showing lesioned area in the aDLS (A), pVLS (B), or DMS (C). Dotted lines in the images indicate the range of the lesioned area. ac, anterior commissure; cc, corpus callosum; and LV, lateral ventricle. (D-F) Learning curves during the single lever press task in the groups injected into the aDLS (D, two-way repeated ANOVA, group, F[1,14] = 1.275, p = 0.278, day, F[1.512,21.171] = 26.281, p = 7.0 × 10-6, group × day, F[1.512,21.171] = 0.834, p = 0.418), pVLS (E, two-way repeated ANOVA, group, F[1,14] = 2.542, p = 0.133, day, F[1.547,21.658] = 40.183, p = 2.2 × 10-7, group × day, F[1.547,21.658] = 0.953, p = 0.380), or DMS (F, two-way repeated ANOVA, group, F[1,11] = 0.025, p = 0.876, day, F[1.205,13.252] = 14.818, p = 0.001, group × day, F[1.205,13.252] = 0.610, p = 0.478). (G-I) Learning curves during the auditory discrimination task in the groups injected into the aDLS (G, two-way repeated ANOVA, group, F[1,14] = 11.578, p = 0.004, day, F[2.805,39.274] = 60.733, p = 1.8 × 10-14, group × day, F[2.805,39.274] = 3.863, p = 0.018, simple main effect, PBS vs. IBO, Day 1, p = 0.140, Day 2, p = 0.085, Day 3, p = 0.284, Day 4, p = 0.068, Day 5, p = 0.011, Day 6, p = 9.5 × 10-5, Day 7, p = 0.001, Day 8, p = 0.002, Day 9, p = 0.012, Day 10, p = 0.059, Day 11, p = 0.034, Day 12, p = 0.046, Day 13, p = 0.136, Day 14, p = 0.096), pVLS (H, two-way repeated ANOVA, group, F[1,14] = 7.221, p = 0.018, day, F[3.646,51.041] = 51.794, p = 9.2 × 10-17, group × day, F[3.646,51.041] = 1.612, p = 0.190), or DMS (I, two-way repeated ANOVA, group, F[1,11] = 0.469, p = 0.507, day, F[2.677,29.452] = 48.860, p = 3.9 × 10-11, group × day, F[2.677,29.452] = 1.094, p = 0.362). (J-L) Changes in the proportions of the win-shift-win or lose-shift-lose strategies through the discrimination learning in the groups treated into the aDLS (J, two-way repeated ANOVA, group, F[1,14] = 13.451, p = 0.003, day, F[3.048,42.673] = 54.777, p = 9.1 × 10-15, group × day, F[3.048,42.673] = 3.288, p = 0.029, simple main effect, PBS vs. IBO, Day 1, p = 0.728, Day 2, p = 0.225, Day 3, p = 0.165, Day 4, p = 0.044, Day 5, p = 0.015, Day 6, p = 1.1 × 10-4, Day 7, p = 0.002, Day 8, p = 5.1 × 10-4, Day 9, p = 0.004, Day 10, p = 0.060, Day 11, p = 0.053, Day 12, p = 0.081, Day 13, p = 0.198, Day 14, p = 0.075 for win-shift-win; group, F[1,14] = 5.039, p = 0.041, day, F[3.866,54.119] = 14.578, p = 5.2 × 10-8, group × day, F[3.866,54.119] = 2.471, p = 0.057 for lose-shift-lose), pVLS (K, two-way repeated ANOVA, group, F[1,14] = 9.251, p = 0.009, day, F[3.206,44.881] = 47.168, p = 2.7 × 10-14, group × day, F[3.206,44.881] = 1.862, p = 0.146 for win-shift-win; group, F[1,14] = 0.762, p = 0.397, day, F[4.773,66.815] = 24.018, p = 1.9 × 10-13, group × day, F[4.773,66.815] = 1.338, p = 0.260 for lose-shift-lose), or DMS (L, two-way repeated ANOVA, group, F[1,11] = 0.703, p = 0.420, day, F[3.052,33.568] = 48.424, p = 2.3 × 10-12, group × day, F[3.052,33.568] = 0.846, p = 0.480 for win-shift-win; group, F[1,11] = 0.002, p = 0.965, day, F[3.442,37.866] = 13.787, p = 1.0 × 10-6, group × day, F[3.442, 37.866] = 0.581, p = 0.654 for lose-shift-lose). Data are indicated as the mean ± s.e.m., and individual data are overlaid (except for panels J–L). The anteroposterior coordinates from bregma (mm) are shown (A–C). Scale bars; 2 mm (A–C). *p < 0.05, **p < 0.01 and ***p < 0.001. Abbreviations: WSW. win-shift-win; and LSL, lose-shift-lose.

Previous studies have reported that the dorsal striatum is involved in both behavioral strategies based on the stimulus-response association and the response-outcome association and during instrumental learning12,13. To assess the impact of striatal lesions on the behavioral strategies, we analyzed the proportions of responses attributed to two strategies in all responses of each session. One is the “win-shift-win” strategy, which is considered to reflect the behavioral strategy based on the stimulus-response association29,30. In this strategy, after a correct response in the previous trial, the rats press the opposite lever in the current trial in response to a shift of the instruction cue, resulting in the correct response. Another strategy is the “lose-shift-lose” strategy, which is considered to appear as a consequence of the behavioral strategy based on the response-outcome association29. In this strategy, after an error response in the previous trial, the rats press the opposite lever in the current trial, despite a shift of the instruction cue, leading to another error response. In the control group into either of three striatal subregions, the proportions of win-shift-win and lose-shift-lose strategies increased and decreased, respectively, during the acquisition phase (Figures 3J–3L). The changes in the proportion of the two strategies were similar to those in the intact rats (Figure S2). For the aDLS injection, the increases in the win-shift-win proportion were markedly impaired in the IBO group compared to the PBS group with significant differences mainly at the middle stage, and the decrease in the lose-shift-lose proportion was delayed compared to the same group (Figure 3J). For the pVLS-injection, the increasing win-shift-win proportion was also impaired in the IBO group compared to the PBS group, although the change in the lose-shift-lose proportion was comparable between the two groups (Figure 3K). In contrast, for the DMS lesion both win-shift-win and lose-shift-lose proportions in the IBO group did not show significant differences from the corresponding values in their control group (Figure 3L). These results indicate that the aDLS promotes the win-shift-win strategy through stimulus-response association and suppresses the lose-shift-lose strategy rooted in response-outcome association, and that the pVLS selectively facilitates the win-shift-win strategy acquisition. Thus, these two subregions not only play roles in the acquisition of auditory discrimination but also control distinct behavioral strategies during the learning processes.

After the completion of discrimination training, we tested locomotion and feeding behavior of rats with aDLS and pVLS lesions. There was no difference in locomotor activity in the open field and the amount of food intake in both lesioned groups (Figures S1J-S1M), suggesting that learning deficits following excitotoxic lesions of the striatal subregions cannot be attributable to abnormalities in these general motor behaviors.

Inhibiting aDLS and pVLS activity impairs discriminative behavior at different stages of the acquisition phase

Our 18F-FDG-PET imaging study indicated that the brain activity in the aDLS and pVLS was increased in a different temporal pattern during the acquisition phase of auditory discrimination. To examine whether neuronal activity in striatal subregions is linked with the processes at different stages, we temporarily inhibited the activity in each subregion by the treatment with a GABAA receptor agonist muscimol (MUS). In a pilot study, we tested the effect of the bilateral injection of several doses of MUS (0.1, 0.2, and 0.5 μg/μL) into the aDLS or pVLS on the single lever press behavior and confirmed that the administration of the lowest dose of MUS did not affect the frequency of single lever presses and the quantity of food intake (Figures S3A–S3D). In addition, to achieve the stage-specific inactivation of the aDLS or pVLS neurons, rats were given bilateral injections of MUS or saline (SAL) at three different timings during a series of the training. The first injection was conducted on Day 2 to target the early stage. The second and third injections were conducted on the day after the success rate had reached 60% and 80% for the first time through the training, respectively, to target the middle or late stage. After the injections, the rats were used for the behavioral test. Placement locations of the injection cannula into the striatal subregions were assessed by cresyl violet staining after the behavioral tests (Figures 4A and 4B). The success rates on the day before the injection (day N-1) at the early, middle, and late stages were similar between the two groups that received the bilateral injections of MUS or SAL (Figures S3E and S3F). Rats were then administered injections into the aDLS on the next day (Day N), and the success rate in the MUS-injected groups significantly decreased at the middle stage compared to the corresponding SAL-injected groups, but not at the early and late stages (Figure 4C, left). In contrast, the success rate in the MUS groups into the pVLS significantly decreased at the late stage, but not at the early and middle stages, compared with the respective SAL groups (Figure 4D, left). To normalize the variation of learning speed in the individual animals, we calculated the value by subtracting the success rate on Day N-1 from that on Day N in the individual animals. There were significant reductions in the subtracted rate at the middle stage in the MUS-injected group into the aDLS (Figure 4C, right) and at the late stage in the MUS-injected group into the pVLS (Figure 4D, right). These results indicate that the aDLS and pVLS mainly function at the middle and late stages, respectively, in the acquisition phase of auditory discrimination, supporting the results obtained from our imaging study.

Influence of transient inhibition of striatal subregions at different timings on the performance of auditory discrimination.

Rats received intracranial injection of SAL or MUS solution into the aDLS (n = 8 for each injection) or pVLS (n = 6 for each injection). (A and B) Representative images of cresyl violet staining and schematic illustrations showing placement sites of the tip of guide cannula in the aDLS (A) or pVLS (B). Dotted lines in the images indicate the position of the cannula placement. Ac, anterior commissure; cc, corpus callosum; and LV, lateral ventricle. (C and B) Effects of transient striatal inhibition on the performance. For aDLS inhibition (C), success rate (left, two-way repeated ANOVA, stage, F[2,28] = 28.708, p = 1.7 × 10-7, group, F[1,14] = 8.840, p = 0.010, stage × group, F[2,28] = 0.757, p = 0.478; unpaired Student’s t-test, early, t[14] = 0.558, p = 0.586; middle, t[14] = 2.764, p = 0.015; Welch’s t-test, late, t[12.010] = 1.617, p = 0.132, p = 0.016 after a Bonferroni correction method) and subtracted success rate on day N-1 from that on day N (right, two-way repeated ANOVA, stage, F[2,28] = 1.231, p = 0.307, group, F[1,14] = 10.797, p = 0.005, stage × group, F[2,28] = 2.360, p = 0.113; unpaired Student’s t-test, early, t[14] = 0.285, p = 0.780, late, t[14] = 1.271, p = 0.225, Welch’s t-test, middle, t[9.603] = 3.331, p = 0.008, p = 0.016 after a Bonferroni correction method). For pVLS inhibition (D), success rate (left, two-way repeated ANOVA, stage, F[2,20] = 17.642, p = 3.8 × 10-5, group, F[1,10] = 43.942, p = 5.9 × 10-5, stage × group, F[2,20] = 4.729, p = 0.012; simple main effect, early, p = 0.055, middle, p = 0.109, late, p = 4.2 × 10-5) and subtracted success rate on day N-1 from that on day N (right, two-way repeated ANOVA, stage, F[2,20] = 12.105, p = 3.6 × 10-4, group, F[1,10] = 16.310, p = 0.002, stage × group, F[2,20] = 8.500, p = 0.002; simple main effect, early, p = 0.595, middle, p = 0.085, late, p = 3.0 × 10-6). (E and F) Subtracted proportion of the win-shift-win or lose-shift-lose strategies in the groups injected into the aDLS at the middle stage (E, unpaired Student’s t-test, t[14] = 2.038, p = 0.061 for win-shift-win, and t[14] = 2.714, p = 0.017 for lose-shift-lose) and the late stage (F, unpaired Student’s t-test, t[14] = 0.898, p = 0.384 for win-shift-win, and t[14] = 0.226, p = 0.824 for lose-shift-lose). (G and H) Subtracted proportion of the win-shift-win or lose-shift-lose strategies in the groups injected into the pVLS at the middle stage (G, unpaired Student’s t-test, t(10) = 1.924, p = 0.083 for win-shift-win, and Welch’s t-test, t[6.095] = 1.364, p = 0.221 for lose-shift-lose) and the late stage (H, unpaired Student’s t-test, t[10] = 3.629, p = 0.005 for win-shift-win, and t[10] = 1.577, p = 0.146 for lose-shift-lose). Data are indicated as the mean ± s.e.m., and individual data are overlaid. The anteroposterior coordinates from bregma (mm) are shown (A and B). Scale bars; 2 mm (A and B). *p < 0.05, **p < 0.01, and ***p < 0.001. Abbreviations: WSW, win-shift-win; and LSL, lose-shift-lose.

Next, we analyzed the influence of temporal striatal inactivation on the behavioral strategy at the middle and late stages of the acquisition phase. We calculated the proportions of the win-shift-win and lose-shift-lose strategies, and subtracted the values of these proportions on Day N-1 from those on Day N. The proportion of both strategies on Day N-1 were comparable between the MUS-and SAL-injected groups (Figures S3G and S3H). In the aDLS, the subtracted proportion of the lose-shift-lose strategy at the middle stage significantly increased in the MUS group as compared to the SAL group, although the subtracted proportion of the win-shift-win strategy tended to decrease in the MUS group as compared to the control group (Figure 4E), and the subtracted proportions of the win-shift-win or lose-shift-lose strategy at the late stage was similar between both groups (Figure 4F). In the pVLS, the subtracted proportions of the win-shift-win and lose-shift-lose strategies at the middle stage were similar between the MUS and SAL groups (Figure 4G), and the subtracted win-shift-win proportion at the late stage significantly decreased in the MUS group as compared to the SAL group, although the subtracted lose-shift-lose proportion was comparable between the two groups (Figure 4H). Our results show that transient inhibition of aDLS activity at the middle stage affects the lose-shift-lose strategy and has a subtle effect on the win-shift-win strategy, supporting that the aDLS mediates the acquisition process of auditory discrimination at the middle stage at least partly through the suppression of the behavior based on the response-outcome association. Although the impact of inhibition on the win-shift-win strategy was subtle, excitotoxic lesion of the aDLS influenced the win-shift-win strategy in addition to the lose-shift-lose strategy (see Figure 3J), suggesting that the aDLS acts to enhance the auditory discrimination by promoting the stimulus-response association, together with the suppression of the response-outcome association. The discrepancy in the extent of changes in the win-shift-win strategy between the two treatments may be generated from the difference in the timing or sustainability of drug inhibition of the striatal subregion. In addition, our results show that pVLS inactivation at the late stage selectively impairs the win-shift-win strategy, supporting contribution of the pVLS to the process via progression of the stimulus-response association.

Neuronal activity in the aDLS and pVLS during the acquisition phase

To examine the firing activity in striatal subregions during the acquisition phase of auditory discrimination, we simultaneously recorded neuronal activity in the aDLS and pVLS at the unilateral (left) side by using a multi-unit recording procedure (Figure 5A). Using immunostaining for tyrosine hydroxylase (TH) after the recordings, we verified the location of electrodes within the striatal subregions (Figure 5B). After the single lever press task, we conducted the auditory discrimination task for the recordings. In correct trials, the delay period (0.5 ± 0.2 s) was inserted between a lever press and a sound indicating the reward to distinguish neuronal activity related to the two events (Figure 5C). The behavioral data in individual rats were divided into three grouped sessions corresponding to the early, middle, and late stages, based on a sigmoid function analysis31 of the success rate in the task. The success rates in the grouped sessions were 52.7 ± 0.8 %, 61.6 ± 0.7 %, and 85.3 ± 1.7 % (mean ± s.e.m.) for the early, middle, and late stages, respectively, and the rate was significantly increased throughout the three stages (Figure S4A). However, the response time, omission rate, and response bias were similar among the stages (Figures S4B-S4D). Thus, these stages corresponded to the timing window defined in the acquisition phase of the transient MUS inhibition experiment of neuronal activity.

Multi-unit recording of neurons in striatal subregions and firing activity related to the behavioral outcome of reward sound-related HR type neurons in the aDLS.

(A) Schematic illustration of a freely moving rat used for simultaneous multi-unit recordings in the aDLS and pVLS. Six rats were used for the following analysis. (B) Representative images showing positions of electrode tips (arrows) in the aDLS and pVLS (left rows). The recording sites estimated by electrode tracks and electrical marks in individual rats are shown (center and right rows). ac, anterior commissure; LV, lateral ventricle. (C) Sequence of some events in correct and error trials. The delay periods were pseudorandomly added between a correct lever press and the reward sound (0.5 ± 0.2 s). The room light turned off the extended time (4 s) after the correct response or immediately after the error response. (D) Mean firing rate (top rows) and auROC values (bottom rows) of reward sound-related HR type neurons in the aDLS at the early (n = 43 neurons), middle (n = 37 neurons), and late (n = 44 neurons) stages. These data are aligned to the choice response. Green shadows show the period when the reward sound is presented (500 ± 200 ms after the lever press). Time bins with significant differences between the mean firing rate in the HR or LR trial and any of the rates in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. (E) Averaged firing rate during the reward sound period in the HR and LR trials at the early, middle, and late stages (Kruskal-Wallis test: HR trial, χ2 = 3.028, p = 0.220; and LR trial, χ2 = 1.249, p = 0.536). (F) Cumulative probability of the proportion of reward sound-related HR type neuron number at the three stages (Kruskal-Wallis test, χ2 = 45.378, p = 1.4 × 10-10; post hoc Tukey-Kramer test, p = 7.1 × 10-5 for early vs. middle, p = 1.0 × 10-9 for early vs. late, and p = 0.040 for middle vs. late). Data are indicated as the mean ± s.e.m. (D) or the median and quartiles with the maximal and minimal values (E). The anteroposterior coordinates from bregma (mm) are shown (B). Scale bars; 2 mm (B). *p < 0.05 and ***p < 0.001.

We identified 1,062 and 423 well-isolated neurons from the aDLS and pVLS, respectively, across the three stages (Figure S4E) and focused on the major population of striatal neurons, putative medium spiny neurons (MSNs; Figures S4F and S4G). The total numbers of MSNs at the early, middle, and late stages were 295, 364, and 360 neurons for the aDLS, and 87, 110, and 190 neurons for pVLS, respectively. Based on the combinatorial pattern of the tone instruction cue and lever press in our discrimination task, we categorized the electrophysiological data into four trial types: (1) high-frequency tone/right lever press (HR) and (2) low-frequency tone/left lever press (LL) as correct responses; and (3) low-frequency tone/right lever press (LR) and (4) high-frequency tone/left lever press (HL) as error responses (see Figure 1A). We identified HR or LL type neurons showing significant changes in firing rate related to the cue onset, choice response, reward sound, or first licking as compared to the baseline firing rate in each of two trials (Mann-Whitney U test, p < 0.01). These neurons were further divided into two groups based on increased or decreased activity (Figures S5A and S5B). In the following analyses, we focused on the HR and LL type neurons with increased event-related activity to explore the firing patterns of neurons underlying behavioral processes of auditory discrimination.

Subpopulations of aDLS neurons, but not of pVLS neurons, show firing activity related to the behavioral outcome

Previous studies have reported that neurons in the dorsal striatum show the outcome-related activity in the instrumental conditioning tasks24,32. We thus investigated whether aDLS and pVLS neurons exhibit a response related to the behavioral outcome during the acquisition of auditory discrimination. In the aDLS, the firing activity of reward sound-related HR type neurons showed a higher extent in the increase during a period, when the reward sound is presented in the correct responses, in the HR trial as compared with the LR, LL, or HL trial throughout the three stages (Figure 5D, top rows). The rate of these neurons also moderately increased during the same period in the LR trial compared to the LL or HL trial across the three stages, although the extent of the increase was lower than that in the HR trial (Figure 5D, top rows). The area under the receiver operating characteristic (auROC) values revealed a significant difference of the HR trial from the LR trial during the reward sound period, whereas the values of HL and LL trials indicated a significant difference in the opposite direction from the LR trial during the same period (Figure 5D, bottom rows). The averaged firing rate during the reward sound period in the HR or LR trial appeared to modestly increase at the middle stage, but no significant difference existed among the stages (Figure 5E). To directly compare the number of neurons showing event-related activity among different stages, we calculated the number of reward sound-HR neurons by matching the trial number at each stage (Figure 5F). The proportion decreased progressively from the early to middle stages, and to the late stage. In addition, the firing rate of reward sound-related LL type neurons exhibited a higher increase in the LL trial and a modest increase in the HL trial while the averaged firing rate during the reward sound period was similar among the three stages. Furthermore, the proportion of reward sound-LL neurons reduced progressively through the stages (Figures S6A–S6C). These results indicate that the aDLS contained neurons related to the outcome of choice, showing a greater increase in the firing activity for the correct responses (HR and LL trials) than the error responses (LR and HL trials), and that the proportion of these subpopulations was gradually reduced through the progress of learning. In contrast, during the reward sound period at all stages, the firing rate and auROC values of reward sound-HR and reward sound-LL neurons in the pVLS revealed no significant differences across the four kinds of trial (Figures S6D and S6E), indicating that the firing activity of pVLS neurons did not appear to be associated with the behavioral outcome.

Subpopulations of aDLS and pVLS neurons show sustained firing activity after the reward in different manners

The dorsal striatum is known to demonstrate a response of neuronal activity after the positive outcome in the instrumental behaviors33,34. We thus examined whether aDLS and pVLS neurons show the response after obtaining the reward during the acquisition phase of auditory discrimination. In the aDLS, the firing rate of the first licking-related HR type neurons indicated a sustained elevation after the first licking in the HR trial as compared with any of the other three trials across the early, middle, and late stages (Figure 6A, top rows). The auROC values revealed a significant difference of the LR, LL, or HL trial from the HR trial across the three stages (Figure 6A, middle rows). In addition, we calculated the averaged firing rate and increased firing period, defined as the total number of time bins above 3 s.d. of the baseline firing rate for 5 s after the first licking in the HR trial. The averaged firing rate was comparable between the early and middle stages, but the rate was significantly lower at the late stage compared to either of the other two stages (Figure 6B). The increased firing period was also comparable between the early and middle stages, showing a significant reduction at the late stage (Figure 6C). The proportions of first licking-HR neurons were similar between the early and late stages, although they showed a slight decrease at the middle stage (Figure 6D). There were few neurons with significant correlations between the firing rate and licking performance (Figure 6E). In addition, the firing rate of the first licking-related LL type neurons exhibited the largest increase in the LL trial at the three stages, showing a tendency to reduce the average firing rate and increased firing period after the first licking across the stages (Figures S7A-S7C). The proportion of first licking-LL neurons during the learning processes displayed similar changes as observed in the first licking-HR neurons (Figure S7D). These results indicate that the aDLS contained neurons showing long-lasting, increased activity after the reward in the specific correct combination of the stimulus and response (HR or LL trial), and that the extent of firing activity and length of firing period in the response appeared to be gradually reduced with no correlation between the proportion of neurons and the progress of the learning stages.

Sustained activity after the reward of first licking-related HR type neurons in the aDLS and pVLS.

(A) Firing rate (top rows), auROC (middle rows), and licking rate (bottom rows) after the reward of first licking-related HR type neurons in the aDLS at the early (n = 27 neurons), middle (n = 31 neurons), and late (n = 49 neurons) stages. These data are aligned to the choice response. Time bins with significant differences between the mean firing rate in the HR trial and any of the rate in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. Timings of the reward sound and first licking are shown. (B) Averaged firing rate for 5 s after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 16.396, p = 2.8 × 10-4, post hoc Tukey-Kramer test, early vs. middle, p = 0.726, early vs. late, p = 7.2 × 10-4, middle vs. late, p = 0.009). (C) Averaged total number of time bins above 3 S.D. of the baseline firing of aDLS neurons for 5 s after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 12.803, p = 0.002, post hoc Tukey-Kramer test, early vs. middle, p = 0.173, early vs. late, p = 0.001, middle vs. late, p = 0.219). (D) Cumulative probability of the proportion of first licking-related HR type neuron number (Kruskal-Wallis test, χ2 = 39.280, p = 3.0 × 10-9, post hoc Tukey-Kramer test, early vs. middle, p = 1.3 × 10-7, early vs. late, p = 0.995, middle vs. late, p = 2.3 × 10-7). (E) Distribution of the correlation coefficient between the numbers of spikes of first licking-related HR type neurons and licking. Closed column indicates the number of neurons showing significant correlations between the two parameters (5 out of 107 neurons). (F) Firing rate (top rows), auROC (middle rows), and licking rate (bottom rows) after the reward of FL-HR neurons in the pVLS at the early (n = 17 neurons), middle (n = 11 neurons), and late (n = 64 neurons) stages. These data are aligned to the choice response. Time bins with significant differences between the rate in the HR trial and any of the rate in other trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. The timings of the reward sound and first licking are shown. (G) Averaged firing rate of pVLS neurons after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 7.652, p = 0.022; post hoc Tukey-Kramer test, early vs. middle, p = 0.975, early vs. late, p = 0.076, and middle vs. late, p = 0.096). (H) Averaged total number of time bins above 3 S.D. of the baseline firing after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 0.023, p = 0.989). (I) Cumulative probability of the proportion of first licking-related HR type neuron number (Kruskal-Wallis test, χ2 = 47.922, p = 3.9 × 10-11; post hoc Tukey-Kramer test, early vs. middle, p = 0.011, early vs. late, p = 1.7 × 10-4, and middle vs. late, p = 9.7 × 10-10). (J) Distribution of the correlation coefficient between the number of spikes in the pVLS and the licking. Closed column indicates the number of neurons showing significant correlations between the two parameters (5 out of 92 neurons). Data are indicated as the mean ± s.e.m. (A and F) or the median and quartiles with the maximal and minimal values (B, C, G, and H). **p < 0.01 and ***p < 0.001. Abbreviations: RS, reward sound; and FL, first licking.

In the pVLS, the first licking-HR neurons showed a sustained elevation in the firing rate after receiving a reward in the HR trial compared to that in the LR or HL trial throughout the three stages (Figure 6F, top rows). A similar elevation in the firing rate was also observed in the LL trial throughout the stages (Figure 6F, top rows). The auROC values revealed a significant difference of the LR or HL trial, but not the LL trial, from the HR trial throughout the stages (Figure 6F, middle rows). The average firing rate did not show significant changes among the three stages (Figure 6G), and the increased firing period was also similar across the stages (Figure 6H). The proportion of first licking-HR neurons increased at the late stage compared to the early or middle stage (Figure 6I). The firing rate showed almost no significant correlation to the licking performance (Figure 6J). In addition, the first licking-LL neurons exhibited similar firing patterns and changes in the proportion of neurons across the stages (Figures S7F–S7J). Therefore, the pVLS included neurons representing the long-lasting increased activity after the reward in the two kinds of correct responses (HR and LL trials) with the increased number of these neurons at the late stage of learning.

Neurons in the pVLS, but not in the aDLS, exhibit firing patterns related to the beginning and ending of the behavior

The dorsal striatum plays a key role in the acquisition and execution of behaviors based on the stimulus-response association, and the association is accompanied by a transient activation of neuronal firing immediately after the stimulus presentation or action in trials, which is considered to represent the beginning and ending of a learned behavior35–37. We first examined whether the firing activity of cue onset-related HR type or cue onset-related LL type neurons corresponds to the beginning of the behavior in the auditory discrimination task. In the aDLS, the firing rate of the cue onset-HR and cue onset-LL neurons showed a significant difference at some time points in each trial compared to those in any of the other three trials, with a significant difference of auROC values between the trials (Figures S8A and S8C). However, the transiently increased activity did not emerge at the time points corresponding to the beginning of the behavior. In contrast, the cue onset-HR neurons in the pVLS showed a transient increase in the firing rate immediately after the cue onset, and the activity appeared at the middle and late stages (Figure 7A, top). The transient activation was observed not only in the HR trial but also in the other three trials, being consistent with the results of a previous report38. The increased firing rate showed a slight but notable difference between the HR and LL trials with a significant difference in the auROC values between the two trials (Figure 7A, low), suggesting that different combinations between the stimulus and response may affect the level of firing activity. Differences in the averaged firing rate between the two kinds of trial was observed at the middle and late stages (Figure 7B). The proportion of cue onset-HR neurons increased at the middle and late stages compared to the early stage (Figure 7C). In addition, the firing rate of the cue onset-LL neurons exhibited patterns similar to those observed in the cue onset-HR neurons (Figure S8E). The data indicate that subpopulations of pVLS neurons show firing activity related to the beginning of the behavior, with a rising number of neurons throughout discrimination learning.

Transient activity related to the beginning and ending of a behavior of cue onset-HR and choice response-related HR type neurons in the pVLS.

(A) Firing rate (top rows) and auROC values (bottom rows) of cue onset-related HR type neurons in the pVLS at the early (n = 4 neurons), middle (n = 21 neurons), and late (n = 24 neurons) stages. These data are aligned to the cue onset. Time bins with significant differences between the mean firing rate in the HR trial and either of the rate in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. (B) Averaged firing rate of the cue onset-related HR type neurons in HR and LL trials (0-600 ms after the cue onset) (Wilcoxon signed rank test; early, p = 0.875, middle, p = 0.006, and late, p = 0.004). (C) Cumulative probability of the proportion of the cue onset-related HR type neuron number (Kruskal-Wallis test, χ2 = 51.114, p = 8.0 × 10-12; post hoc Tukey-Kramer test, early vs. middle, p = 9.6 × 10- 10, early vs. late, p = 0.002, and middle vs. late, p = 6.4 × 10-4). (D) Firing rate (top rows) and auROC values (bottom rows) of the choice response-related HR type neurons in the pVLS at the early (n = 11 neurons), middle (n = 22 neurons), and late (n = 41 neurons) stages. These data are aligned to the choice response. Time bins with significant differences between the mean rate in the HR trial and any of the rate in other trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. The timing of the reward sound is presented. (E) Averaged firing rate of the choice response-related HR type neurons in HR and LL trials (0-600 ms after the choice response) (Wilcoxon signed rank test; early, p = 0.278, middle, p = 0.758, and late, p = 0.279). (F) Cumulative probability of the proportion of the choice response-related HR type neuron number (Kruskal-Wallis test, χ2 = 13.786, p = 0.001; post hoc Tukey-Kramer test, early vs. middle, p = 0.003, early vs. late, p = 0.005, and middle vs. late, p = 0.976). Data are indicated as the mean ± s.e.m. (A and B) or the median and quartiles with the maximal and minimal values (B and E). **p < 0.01 and ***p < 0.001. Abbreviation: RS, reward sound.

Next, we examined the firing activity of choice response-related HR type or choice response-related LL type neurons, corresponding to the ending of a behavior in auditory discrimination. In the aDLS, the firing rate of these neurons did not exhibit the transient increase related to the ending of the behavior, instead, there was a modest increase before the lever press (Figures S8B and S8D). In contrast, the choice response-HR neurons in the pVLS were transiently activated immediately after the lever press in the four kinds of trial at the middle and late stages, and the rate appeared to gradually decline toward the reward sound presentation (Figure 7D). The averaged firing rate was similar between the HR and LL trials though the stages (Figure 7E). The proportion of the choice response-HR neurons increased at the middle and late stages compared to the early stage (Figure 7F). In addition, the rate of choice response-LL neurons displayed similar firing patterns to those observed in the choice response-HR neurons (Figure S8F). Thus, subpopulations of pVLS neurons also represent the activity related to the ending of the behavior in auditory discrimination, increasing the neuron number throughout the learning stages in a similar pattern to neurons relevant to the beginning of the behavior in the task.

Discussion

Our findings demonstrate that the aDLS and pVLS are required for the acquisition of auditory discrimination at different stages of learning, and that the DMS is not necessary for the acquisition of discrimination, although it appears to contribute to the execution of actions, in particular to the regulation of response time. These results not only support some recent studies reporting no requirement of the DMS in the procedural learning20,39,40, but also reveal important roles of two different subregions in the lateral striatum in the external cue-dependent decision-making. We propose that aDLS and pVLS neurons act to integrate the new learning of auditory discrimination in spatiotemporally and functionally distinct manners, showing a remarkable difference from the prior learning model that proposes the functional dominance changes from the DMS to DLS subregions during procedural learning12–16.

Our 18F-FDG-PET imaging study demonstrated that the brain activity in the aDLS reaches a peak at the middle stage of the acquisition phase of auditory discrimination (Figure 2I). The results of excitotoxic lesion experiment of striatal subregions suggest that the aDLS functions to enhance the acquisition of discrimination through the promotion of behavior based on the stimulus-response association and the suppression of behavior based on the response-outcome association (Figure 3J). Transient inhibition experiment of neuronal activity confirmed this aDLS function at the middle stage of the learning, although the drug inhibition tended to reduce the stimulus-response behavior (Figures 4E and 4F). In addition, we found subpopulations of aDLS neurons (reward sound-related neurons) which represent the behavioral outcome and differentiate the correct (HR or LL) and incorrect (LR or HL) choices (Figures 5D and S6A). The distinct responses of these neurons may explain the mechanism by which the aDLS regulates the stimulus-response and response-outcome associations through distinct neural circuits leading to a differential control of the two associations. Alternatively, the responses mainly promote the behavior based on the stimulus-response association, thereby secondarily inhibiting the response-outcome association. Indeed, some previous studies have suggested that an action based on either the stimulus-response association or response-outcome association is generated as a result of competition between different striatal subregions41. We also found other subpopulations of aDLS neurons (first licking-related neurons) showing long-lasting increased activity after obtaining reward in response to a specific combination between the stimulus and response (HR or LL trial) (Figures 6A and S7A). This activity may function as the feedback signal to promote the stimulus and response association, resulting in the enhancement of discrimination learning. The proportion of neurons related to the behavioral outcome showed a gradual reduction through the progress of auditory discrimination (Figures 5E and S6C), and the extent of the firing activity and the length of the firing period of neurons showing sustained activation after reward were also decreased at the late stage of learning (Figures 6B and 6C). The learning-dependent changes in the properties of these subpopulations in the aDLS may explain the mechanism underlying the earlier acquisition process of the auditory discrimination.

The 18F-FDG-PET imaging indicated that the pVLS activity was gradually elevated through the acquisition phase of auditory discrimination, showing the maximal value at the late stage (Figure 2J). The lesion experiment suggested that the pVLS drives the formation of discrimination learning via the selective progression of the stimulus-response association (Figure 3K), and the transient inhibition experiment ascertained the pVLS function at the late stage of learning (Figures 4G and 4H). Previous studies have shown that the stimulus-response association is encoded in the DLS or VLS as the transient neuronal activity representing the beginning and ending of a learned behavior35–37,42. A recent study reported that the VLS is causally related to learned performance based on the auditory stimulus, while the DLS shows a weaker relationship to the behavior22. In the present study, we found subpopulations of pVLS neurons that represent the beginning and ending of a behavior (cue onset- and choice response-related neurons, respectively) (Figures 7A, 7D, S8E, and S8F), showing an increased number of neurons along with the progress of discrimination learning (Figures 7C and 7F). These data suggest that pVLS neurons contribute to the formation of the stimulus-response association through the learning processes. In addition, we found another subpopulation of pVLS neurons (first licking-related neurons) that are continuously activated after the reward outcome in response to two kinds of correct choices (HR and LL trials) (Figures 6F and S7F), and the proportion of neurons was increased at the late stage (Figures 6I and S7I). This activity may act as a feedback signal to consolidate the stimulus and response association, leading to the maintenance of associative memory at the late stage during the acquisition phase of learning.

In the present study, we identified neurons with four kinds of event-related activity in the aDLS and pVLS, and classified them into the HR and LL type neurons in which their firing activity showed a significant increase in the HR and LL trials, respectively, as compared to the baseline (Figures 5–7 and S6-S8). Parts of these two types of neurons with the respective event-related activity overlapped, and the numbers of overlapping neurons varied between striatal subregions (Figures S9A and S9B). In the pVLS, the overlapping neurons for the same event-related activities were 35.1%, 45.2%, and 67.0% in the cue onset, choice response, and first licking, respectively. In contrast, the aDLS had lower overlaps of 12.6% and 22.7% in the reward sound and first licking, respectively. Thus, distinct neurons in the aDLS tend to have properties of a single trial type while the same neurons in the pVLS seem to possess properties of both types. One possible explanation for the diversity in the proportion of overlapping neurons between the two subregions may be the difference of input patterns into striatal neurons or plastic changes in functional connectivity during learning processes. The detailed mechanism by which aDLS and pVLS neurons react to the events in a different manner should be investigated in the future.

Anatomically, striatal subregions are known to form heterogeneous cortico-basal ganglia loops in mice2,43,44. We examined the neural circuits linked to the aDLS and pVLS in rats by using bidirectional tracers. The results showed that two striatal subregions receive innervations from different subregions in the cerebral cortex, intralaminar thalamus, and substantia nigra pars compacta, and send projections to different subregions in the basal ganglia nuclei, including the globus pallidus, entopeduncular nucleus, and substantia nigra pars reticulata (Figure S10), confirming that the aDLS and pVLS form the heterogenous cortico-basal ganglia loops in the rat brain. These data suggest that the aDLS and pVLS mediate the acquisition of auditory discrimination via the parallel loops, and cooperate to achieve neural processing related to the dual regulation of the stimulus-response and response-outcome associations. Further studies are needed to elucidate the mechanism by which the two kinds of loops are processed separately and integrate the new learning of discrimination.

Limitations

In the 18F-FDG PET neuroimaging study, the brain activity in the aDLS reached a peak at the middle stage of the acquisition phase of auditory discrimination (Figure 2I). In the multi-unit electrophysiological recording experiment, the firing activity of aDLS neuron subpopulations related to the behavioral outcome showed no significant differences among the three stages (Figure 5E), while the proportions of these subpopulations gradually reduced through the progress of learning stages (Figure 5F). The extent of firing activity and length of firing period of other subpopulations showing sustained activation after the reward appeared to show learning-dependent decrease (Figure 6B and 6C), although the proportions of these subpopulations indicated no correlation to the progress of the learning (Figure 6D). The patterns of the temporal changes in the brain activity of the striatal subregions during the learning stages did not match completely the time variation in the property and proportion of specific event-related neurons. In our electrophysiological analysis, we identified well-isolated neurons from striatal subregions across the behavioral task, focusing on putative MSNs (Figures S4E–S4G). Based on the combinatorial pattern of the tone instruction cue and lever press, we categorized the electrophysiological data into the four trials, including the HR, LL. LR, and HL. We identified HR or LL type neurons showing significant changes in the firing rate related to specific events compared to the baseline firing rate. These neurons were further divided into two groups with increased or decreased activity relative to the baseline firing (Figures S5A and S5B). In the present study, we focused on event-related neurons with the increased activity. Because the analysis was limited to neuronal subpopulations related to specific events with the increased activity, we consider that it is difficult to explain fully dynamic shifts in the brain activity of the striatal subregions dependent on the progress of learning by the time variation of firing activity of individual event-related neurons. The activity of other subpopulations in the striatum may be involved in shifts in brain activity during the learning processes. In addition, recent studies have reported that the activity of glial cells influences the uptake of 18F-FDG28 and that these cells regulate spike timing-dependent plasticity45. Changes in glial cellular activity, through the control of synaptic plasticity, may partly contribute to the pattern formation of learning-dependent shift of brain activity.

Materials and Methods

Animals

Animal care and handling procedures were carried out in accordance with the guidelines established by RIKEN Center for Biosystems Dynamics Research, Fukushima Medical University, and Osaka City University. All procedures were approved by their Institutional Animal Care and Use Committees. All efforts were made to minimize the number of animals used and their suffering throughout the course of the experiments. Male Long-Evans rats (8–13 weeks old, Institute for Animal Reproduction (Ibaraki, Japan)) were used for the present study. The rats were maintained at 22 ± 2 °C and 60% humidity in a 12-h light/12-h dark cycle, and food and water were continuously available unless stated otherwise.

Surgery

The rats underwent a stereotaxic surgery under isoflurane anesthesia (1–5%). For excitotoxic lesion of striatal neurons, IBO was dissolved in PBS (0.8 mg/mL, Wako Pure Chemical), and IBO solution (0.3 μL per site) was injected bilaterally with the coordinates (mm) from bregma and dura for the aDLS (AP +1.6, ML ±3.3, DV -3.9), pVLS (AP -1.6, ML ±4.6, and DV -5.6), or DMS (AP -1.6, ML ±1.4, and DV -4.8) according to the atlas of rat brain46. Injection was performed through a glass pipette (diameter: 60 μm) at a constant velocity of 0.1 mL/min with the microinfusion pump (EICOM). For transient inactivation of striatal neurons, two guide cannulae (length: 10 mm, outer diameter: 26 gauge) were bilaterally placed on the skull and fixed using four screws and dental cement. The chip of the cannula was placed at 1 mm above the target coordinates. A solution containing MUS (Sigma-Aldrich) in saline (0.1 μg/μL, 0.25 μL per site) was injected into the target coordinates through the internal cannula (outer diameter: 30 gauge) at a constant velocity of 0.25 μL/min. For anterograde and retrograde tracing, a solution containing cholera toxin subunit B (CTb; 1.0 mg/mL) conjugated to Alexa 488 or Alexa 555 (Invitrogen) as a bidirectional tracer was injected into the aDLS and pVLS (0.3 μL/site), respectively. The coordinates for the injection into striatal subregions were the same as described above. For the electrophysiological experiments, surgery was performed as described previously47. A 64-channel silicon probe (Buzsaki64spl; NeuroNexus) attaching to a 3D-printed microdrive with a movable screw was chronically implanted with the coordinates (mm) from the bregma and dura into the left aDLS (AP +1.6, ML -3.3, and DV -2.5) and pVLS (AP -1.6, ML -4.6, and DV -3.6). A faraday cage composed of copper mesh was secured on the skull with dental cement to reduce electrical noise and avoid damage to the implants. The rats were allowed to recover for 1 week after the surgeries.

Behavioral analysis

Behavioral tasks were conducted in operant conditioning chambers (30.5 × 24.1 × 29.2 cm; ENV-007CT; MED Associates) equipped with two retractable levers on both sides of a reward port in the front panel (ENV-203M-45, ENV-652AM, and ENV-201A; MED Associates). A multi-tone generator (ENV-223; MED Associates) and room light were mounted on the top of the center and corner panels on the rear side, respectively. For the electrophysical experiment, a 3D printed reward port was used to detect the licking and to prevent contact with the head attachment and front panel. The behavioral experiment in intact rats was performed using grain pellets (F0021, Bioserv) as a food reward. To prevent spillover of radioactivity into the temporal cortex as a consequence of excessive 18F-FDG accumulation into the muscles by their activity to eat the pellets48,49, the PET imaging experiments were carried out using 0.02% saccharine solution (B0131, Tokyo Chemical Industry) as a liquid reward. The electrophysiological experiment was also carried out using the saccharine solution. The pharmacological experiment was performed using the pellets.

The rats were subjected to dietary restrictions of 12 g/day (CE-2, CLEA Japan) and water access was limited to 6 to 9 h in accordance with the reward used in the experiments for 3 days before the beginning of the training schedule, and the body weights of the rats were maintained at ∼85% of their normal weights. The training schedule consisted of four continuous steps. In the first step (magazine training), the rats were habituated to the operant conditioning chambers, and the reward was presented with the click every 20 s. Each daily session was started by illumination of the room light, and continued for 30 min. The sessions were conducted for 3 consecutive days. In the second step (shaping), only one lever on the left or right side was inserted. When the rats pressed the lever, they earned a reward accompanied by a click. Each daily session was started by illumination of the light, and lasted 30 min or until the criteria, in which the rats received 90 rewards. When the number of lever presses on either side reached the criteria, the lever was inserted in the opposite side on the next day. When the rats achieved the criteria on both sides for 2 or 3 days, they moved to the following step. In the third step (single lever press task), each trial was started by illumination of the light and insertion of one lever on either the left or right side in a pseudorandom manner with equal frequencies on both sides for every set of four trials. When the rats pressed the lever, the reward was presented with the sound, and the lever was immediately retracted. The room light turned off 4 s after the lever press. When they did not press the lever within 10 s after the start, the trial was terminated when the light turned off. The trial was repeated every 20 s. Each daily session continued for 30 min, and the sessions were conducted for 4 consecutive days. The number of lever presses in each session was counted. In the final step (two-alternative auditory discrimination task), each trial was started by the presentation of a high (10 kHz) or low (2 kHz) frequency tone as an instruction cue in a pseudorandom manner. After three seconds, the room light was illuminated, and two retractable levers were inserted at the same time. The rats were required to press the left or right levers in response to the low or high tone cues, respectively. When the rats pressed the correct lever, the reward was presented with the sound, and the levers were immediately retracted. Correct response allowed the rats to access the reward: the retractable spout delivering water was inserted for 1 s for the PET imaging experiment; a pellet was delivered by a food dispenser for the pharmacological experiment; and three drops of water were delivered through a solenoid valve for the electrophysiological experiment. The room light turned off 4 s after the correct response. On the other hand, error response resulted in the retraction of levers and the turning off of the room light. No response within 10 s after the lever insertion was recorded as an omission response. The success rate in each session was calculated by dividing the number of correct response trials with the number of lever press trials minus the number of omission trials. The trial was repeated every 20 s. Each daily session continued for 60 min, and the sessions were conducted for 14 or 24 days according to the experiments.

To assess the behavioral strategies in the auditory discrimination task, a trial-by-trial analysis was conducted; the win-shift-win strategy was defined as a correct response, in which, after a correct response during the previous trial, the rats press the opposite lever during the current trial in response to the shift of the tone instruction cue; the lose-shift-lose strategy was defined as an error response, in which, after an error response during the previous trial, the rats press the opposite lever during the current trial despite the shift of the instruction cue. The proportion of both strategies was calculated by dividing the number of each strategy with the number of lever press trials minus the number of omission trials.

For the transient inactivation experiment, MUS injections were bilaterally performed 20 min before start of the session. For habituation, all rats were injected with saline bilaterally into the corresponding striatal subregions on the last day (Day 4) of the single lever press task. During the auditory discrimination task, the first injection was conducted on Day 2 of the task. The second and third injections were conducted the next day, after the success rate had exceeded 60% and 80%, respectively. Each session continued for 15 min.

For the electrophysiological experiments, a delay period (0.5 ± 0.2 s) was added in the auditory discrimination task between the lever press and reward sound to distinguish the motor- and reward-related responses of neurons. When the rats pressed the lever, the lever was immediately retracted. After the delay period, the reward was presented with a sound. The room light turned off 4 s after the lever press. The trial was repeated every 20 s, and each daily session continued for 60 min.

PET imaging analysis

An 18F-FDG-PET scan was performed with a microPET Focus220 (Siemens Medical Solutions, Knoxville) designed for high-resolution imaging of small animals (spatial resolution of 1.4 mm in full width at half maximum at the center of the field of view) as described previously26. Serial 18F-FDG-PET scans were performed on the last day of the single lever press task and on Days 2, 6, 10, and 24 of auditory discrimination task. To administer the 18F-FDG under the awake rats, an indwelling catheter was attached to the tail under anesthesia with 5% induction and 2.0–2.5% maintenance of isoflurane (median [interquartile range]: 6 [5–8] min). The recovery period was set to at least 2 hours until the beginning of the behavioral session, with the aim of minimizing the impact of anesthesia as much as possible. A solution of 18F-FDG (ca. 74 MBq/0.4 mL) was intravenously injected just before the beginning of each behavioral session through the indwelling catheter. After the 30-min session, the rats were returned to their home cage, and then a 30-min static PET scan was performed under anesthesia with 2.0–2.5% isoflurane 55-min after the 18F-FDG injection. During the PET scan, the body temperature of the anesthetized rats was maintained at 37°C using a small animal warmer (BWT-100A, Bio Research Center). The acquired emission data were sorted into a single sinogram and reconstructed by a standard 2D filtered back projection (FBP) with a ramp filter and cutoff frequency at 0.5 cycles per pixel or by a statistical maximum a posteriori probability algorithm (MAP), 12 iterations with point spread function effect.

Voxel-based analysis of PET images was conducted according to the method described previously26. Briefly, individual MAP-reconstructed FDG (MAP-FDG) images were aligned to a FDG template image of Long Evans rats using PMOD software package (version 3.2, PMOD Technologies), and then the aligned images were transformed into the space of a T2 weighted MRI templates for the Long-Evans rats (https://www.nitrc.org/projects/tpm_rat/). The transformation parameters obtained from individual MAP-FDG images were used to match the individual FBP-reconstructed FDG (FBP-FDG) images with the MRI templates. The voxel size of the template was magnified by 10 to a size similar to that of a human brain. All FBP-FDG images were resampled with a voxel size of 1.2 × 1.2 × 1.2 mm and spatially smoothed through an isotropic Gaussian kernel (6-mm FWHM). Voxel-based statistical analysis was performed by using the Statistical Parametric Mapping (SPM) software (version 8; https://www.fil.ion.ucl.ac.uk/spm/). For visualization of the results, SPM t-statistic values were overlayed on MRI templates to define the voxels showing significantly increased or decreased brain activity, using a high threshold (p = 0.001, uncorrected) and an extent threshold (with the output values from the voxel-based analysis). The voxel of interest (VOI) was determined ass voxels with more than 85% of the maximum T-value within activated areas obtained by voxel-based statistical comparison of FBP-FDG images. The average regional 18F-FDG uptake was calculated on the basis of the values of the VOI in FBP-FDG images.

Electrophysiological recording

Extracellular multi-unit recording was conducted at a sampling rate of 20 kHz from the striatum of freely moving rats by using a 512-ch acquisition board (Open Ephys) via 64-ch recording headstages (C3325; Intan Technologies). The tips of the probe were lowered by turning the screw (> 140 µm/day) until it reached the target region (the aDLS and pVLS) during the single lever press task. The tips were sometimes lowered by the same procedure (140 to 280 µm/day) after the daily recording session to improve cell yield.

Spike sorting was performed automatically by the Kilosort software (https://github.com/cortex-lab/KiloSort). Clear noise clusters were removed in the first manual sorting using the phy GUI (https://github.com/cortex-lab/phy). The remaining clusters were then re-clustered by using the Klusta Kwik (https://klusta.readthedocs.io/en/latest/), and the clusters were adjusted in the second manual sorting by using the phy GUI again. The sorted clusters satisfying all the following criteria were used for further analysis: L-ratio < 0.05, isolation distance > 15, and ISI index < 0.250,51. Well-isolated single neurons (n = 1,485; aDLS, n = 1,062; pVLS, n = 423) were classified as putative medium spiny neurons (MSNs), fast spiking interneurons (FSIs), tonically active neurons (TANs), and unclassified interneurons (UIs). Putative MSNs and TANs were separated from FSIs and UIs by trough-to-peak duration of mean spike waveform (> 0.6 ms) according to previous studies, with slight modifications52,53. Putative FSIs were then separated from UIs by utilizing the proportion of time associated with long inter-spike intervals. This was achieved by summing the inter-spike intervals longer than 2 s and subsequently dividing the resulting sum by the total recording time (PropISIs > 2 s)54. Neurons with values of PropISIs > 2 s less than 0.4 were classified as putative FSIs, while neurons with PropISIs > 2 s greater than 0.4 were classified as putative UIs. The putative MSNs and TANs were separated by measuring the post-spike suppression54. We measured the length of suppressed time during which the firing rate of a neuron was suppressed following a spike by counting the number of 1-ms bins in its autocorrelation function until the rate of the neuron was equal to or greater than its averaged rate over the 600-ms to 900-ms autocorrelation bins. Putative TANs were separated by post-spike suppression of less than 40 ms, and the remaining neurons were classified as putative MSNs. Because the numbers of FSIs, TANs, and UIs were low (79 out of 1,485, 5.3%), they were excluded from further analysis.

To divide the three stages during the acquisition phase, the sigmoidal function was fitted using the following four-parameter logistic equation: Where α is the inflection point, β is a variable for slope, and λ and γ are parameters related to the lower asymptote (γ) and upper asymptote (λ) of the learning curve. The function was used to evaluate the day when the success rate reached 20% (t20%) or 90% (t90%) of the maximum rate change defined by the difference between the γ and λ. Three stages were defined according to the following criteria: early stage (up to 3 days before the middle stage); middle stage (3 days including before and after the day of t20%); and late stage (3 days including the day of t90% and before). One rat learned the discrimination faster, and the success rate reached t90% 2 days after the day of t20%. One day before the day of t90% was assigned to the late stage. For the following analysis, spike data in each striatal subregion were separated into the early, middle, and late stages (n = 295, 364 and 360, respectively, for the aDLS; and n = 87, 110 and 190, respectively, for the pVLS).

To determine the neurons showing significant changes in firing rate related to the task events, electrophysiological data were grouped into four trial types based on the instruction cue (high or low tone) and the choice (right or left lever); HR, high tone and right lever press; HL, high tone and left lever press; LR, low tone and right lever press; and LL, low tone and left lever press. Neurons with three or more trial data were analyzed. We identified the HR or LL type neurons, detecting significant increases or decreases of the firing rate by applying the Wilcoxon rank-sum test (p < 0.01) to trial-by-trial spike number between the baseline period (BL, -1,500 to -500 ms or -1,100 to -500 ms before presentation of the instruction cue) and any of four peri-event periods (cue onset, 0 to +1,000 ms after the onset of the instruction cue; choice response, -700 to +300 ms around the time of lever press; reward sound, -300 to 300 ms around the time of sound onset instructing the reward delivery; and first licking, 0 to +1,000 ms after the time of first licking after the choice response). Any spikes around the events were averaged in 50-ms bins. The event-related activity was shown as averaged firing rate aligned based on the timing of an event for each trial. The auROC values were calculated by using the spike counts between the HR or LR trials and any of the other three trials. Bins showing significant differences in the auROC data were determined by applying the Wilcoxon rank-sum test (p < 0.05) to the distribution of 200 ms bin-by-bin auROC values and 0.5.

To compare the number of neurons showing event-related activity among different stages, the number of HR or LL type neurons were calculated by matching the trial number at each stage. Twenty trials were randomly selected for a neuron, and neurons with less than 20 trials were removed from the analysis. The neurons with event-related activity were identified as described above. The proportion (percentage) of neurons relative to the total number of neurons was calculated at each individual stage. The same analysis was repeated 30 times and the distribution of the proportion of neurons with event-related activity at the stage was obtained. The cumulative probability against the proportion was calculated using cumulative distribution function (cdfplot, MATLAB code from Mathworks).

Licking responses were detected with an infrared photobeam sensor placed in the 3D printed reward port, and generated timestamps for the onset of each spout contact according to the analog voltage signal. The licking responses corresponding to four kinds of trial in the first licking-related neurons were shown as averaged rates (50-ms bins) aligned based on the timing of the choice response. To calculate the correlation coefficient between neuronal activity and licking responses, the spikes and licking for 5 s after the first licking were counted every trial for each first licking-related neuron. Trials in which neurons never responded were removed from both the spike and licking data. The correlation coefficients between the two remaining data were calculated for each neuron (corr, MATLAB code from Mathworks), and neurons showing significant differences were determined by applying the Pearson correlation coefficient (p < 0.01).

Histology

After the behavioral experiments, the rats were anesthetized with pentobarbital (50 mg/kg body weight) and isoflurane (3%), then transcardially perfused 4% paraformaldehyde in phosphate-buffer (PB). The brains were post-fixed overnight at 4 ℃ and cryoprotect with 10, 20, and 30% sucrose in PB. Brain tissues were cut into 30-µm sections by using a cryostat (Leica). For the lesion experiment, cresyl violet staining or immunostaining were performed to define the lesioned areas in the striatum. The sections were stained with a mouse monoclonal anti-NeuN IgG (1:1000 dilution; Millipore, Cat# MAB377, RRID:AB 2298772), and then with a goat anti-mouse IgG conjugated with Alexa Fluor 488 (1:400 dilution; Thermo Fisher Scientific, Cat# A-11029, RRID:AB 2534088). The sections were counterstained with 0.3 μM 4’,6-diamidino-2-phenylindole (DAPI; Sigma-Aldrich). For transient inactivation experiment, sections were stained by cresyl violet to visualize the positions of guide cannulae. For the electrophysiological experiments, small lesions were made by applying an electrolytic current (3 μA, 10 s) through top and bottom channels of each shank by using the stimulus isolator (A365, World Precision Instrument). Sections through the striatum were stained with a mouse monoclonal anti-TH IgG (1:1,000 dilution; Millipore, Cat# MAB318, RRID:AB 2201528), and then with a goat anti-mouse IgG conjugated with Alexa Fluor 488 (1:400 dilution; Thermo Fisher Scientific, Cat# A-11029, RRID:AB 2534088). The recording sites were verified by the staining with DAPI and red fluorescent Nissl stain solution (1:200 dilution; Thermo Fisher Scientific, Cat# N21482, RRID:AB 2620170). Fluorescence and immunostained signals were visualized under fluorescence microscopes (BZ-X800; Keyence) and confocal microscope (A1R; Nikon).

Statistical analysis

Statistical analysis was performed using MATLAB (MathWorks) and SPSS Statistics (Ver. 27, IBM). Comparisons of data between two groups were performed using the Student’s t-test (paired), the Wilcoxon signed rank test (paired), and the Wilcoxon rank-sum test (unpaired). Comparisons of data among three groups were constructed using a one-way repeated measures ANOVA, two-way repeated measures ANOVA, or the Kruskal-Wallis test (without assuming normal distribution and homogeneity of variances). All statistical tests were two-sided. The Bonferroni post-hoc test and Mann–Whitney U-test with Bonferroni adjustment were carried out as needed to compare means. For boxplots, the data points below the lower (25th percentile) -1.5 interquartile range (IQR) and above the upper (75th percentile) quartiles +1.5 IQR were regarded as outliers, and they are not included in the boxplots.

eLife Assessment

This study provides an important understanding of the contribution of different striatal subregions, the anterior Dorsal Lateral Striatum (aDLS) and the posterior Ventrolateral Striatum (pVLS), to auditory discrimination learning. The authors have included robust behavior combined with multiple observational and perturbation techniques. The data provided are convincing of the relevance of task-related activity in these two subregions during learning.

Reviewer #1 (Public review):

Summary:

In this study, Setogawa et al. employ an auditory discrimination task in freely moving rats, coupled with small animal imaging, electrophysiological recordings, and pharmacological inhibition/lesioning experiments to better understand the role of two striatal subregions: the anterior Dorsal Lateral Striatum (aDLS) and the posterior Ventrolateral Striatum (pVLS), during auditory discrimination learning. Attempting to better understand the contribution of different striatal subregions to sensory discrimination learning strikes me as a highly relevant and timely question, and the data presented in this study are certainly of major interest to the field. The authors have set up a robust behavioral task, systematically tackled the question about a striatal role in learning with multiple observational and manipulative techniques. Additionally, the structured approach the authors take by using neuroimaging to inform their pharmacological manipulation experiments and electrophysiological recordings is a strength.

Comments on revisions:

The authors have addressed some concerns raised in the initial review but some remain. In particular it is still unclear what conclusions can be drawn about task-related activity from scans that are performed 30 minutes after the behavioral task. I continue to think that a reorganization/analysis data according to event type would be useful and easier to interpret across the two brain areas, but the authors did not choose to do this. Finally, switching the cue-response association, I am convinced, would help to strengthen this study.

Reviewer #2 (Public review):

The study by Setogawa et al. aims to understand the role that different striatal subregions belonging to parallel brain circuits have in associative learning and discrimination learning (S-O-R and S-R tasks). Strengths of the study are the use of multiple methodologies to measure and manipulate brain activity in rats, from microPET imaging to excitotoxic lesions and multielectrode recordings across anterior dorsolateral (aDLS), posterior ventral lateral (pVLS)and dorsomedial (DMS) striatum.

The main conclusions are that the aDLS promotes stimulus-response association and suppresses response-outcome associations. The pVLS is engaged in the formation and maintenance of the stimulus-response association. There is a lot of work done and some interesting findings however, the manuscript can be improved by clarifying the presentation and reasoning. The inclusion of important controls will enhance the rigor of the data interpretation and conclusions.

Comments on revisions:

The authors have made important revisions to the manuscript and it has improved in clarity. They also added several figures in the rebuttal letter to answer questions by the reviewers. I would ask that these figures are also made public as part of the authors' response or if not, included in the manuscript.

Author response:

The following is the authors’ response to the previous reviews.

Although we have no further revisions on the manuscript, we would like to respond to the remaining comments from the reviewers as follows.

Reviewer 1:

The authors have addressed some concerns raised in the initial review but some remain. In particular it is still unclear what conclusions can be drawn about taskrelated activity from scans that are performed 30 minutes after the behavioral task. I continue to think that a reorganization/analysis data according to event type would be useful and easier to interpret across the two brain areas, but the authors did not choose to do this. Finally, switching the cue-response association, I am convinced, would help to strengthen this study.

As for the task-related activity, the strategy for PET scan was explained in our response to the comment 2 from Reviewer 2. Briefly, rats receive intravenous administration of 18F-FDG solution before the start of the behavioral session. The 18FFDG uptake into the cells starts immediately and reaches the maximum level until 30 min, being kept at least for 1 h. A 30-min PET scan is executed 25 min after the session. Therefore, the brain activity reflects the metabolic state during task performance in rats.

Regarding data presentation of the electrophysiological experiments, we described the subpopulations of event-related neurons showing notable neuronal activity patterns in the order of aDLS and pVLS, according to the procedure of explanations for the behavioral study

For switching the cue-response association, we mentioned the difference in firing activity between HR and LL trials, suggesting that different combinations between the stimulus and response may affect the level of firing activity. As suggested by the reviewer, an examination of switching the cue-response association is useful to confirm our interpretation. We will address this issue in our future studies.

Reviewer 2:

The authors have made important revisions to the manuscript and it has improved in clarity. They also added several figures in the rebuttal letter to answer questions by the reviewers. I would ask that these figures are also made public as part of the authors' response or if not, included in the manuscript.

We will present the figures publicly available as part of our response.

Version 3

Abstract

Abstract

The striatum, the central hub of cortico-basal ganglia loops, contains functionally heterogeneous subregions distinguished by the topographic patterns of structural connectivity. These subregions mediate various processes of procedural learning. However, it remains unclear when and how striatal subregions engage in the acquisition of sensory stimulus-based decision-making. A neuroimaging of regional brain activity shows that the anterior dorsolateral striatum (aDLS) and posterior ventrolateral striatum (pVLS) in rats are activated in a different temporal pattern during the acquisition phase of auditory discrimination. Chronic and transient pharmacologic manipulations show that the aDLS promotes the behavioral strategy driven by the stimulus- response association while suppressing that by the response-outcome association, and that the pVLS contributes to forming and maintaining the stimulus-response strategy. Electrophysiological recording indicates that subpopulations of aDLS neurons predominantly represent the outcome of specific behaviors at the initial period of discrimination learning, and that pVLS subpopulations encode the beginning and ending of each behavior according to the progress of learning. In addition, other subpopulations of striatal neurons indicate sustained activation after obtaining reward with distinct patterns reflecting the stimulus-response associations. Our findings demonstrate that aDLS and pVLS neurons integrate the new learning of auditory discrimination in spatiotemporally and functionally different manners.

Introduction

The neural network connecting the cerebral cortex, basal ganglia, and thalamus forms functionally heterogeneous loops distinguished by the topographic patterns of connectivity (Alexander and Crutcher, 1990; Foster et al., 2021), and is implicated in various learning processes by the integration of cognitive, sensorimotor, and reward information (Kim and Hikosaka, 2015; Yin and Knowlton, 2006). The striatum is a key node in the cortico-basal ganglia loops, and its segregated subregions are considered to mediate procedural learning, such as motor skills and instrumental behavior, through parallel processing via different loops (Kupferschmidt et al., 2017; Thorn et al., 2010; Yin et al., 2009). Previous studies in humans and non-human primates have reported that the neural activity within the striatum shifts from the associative subregions (caudate and anterior putamen) to the sensorimotor subregions (posterior putamen) as motor learning progresses (Doyon et al., 2009; Lehéricy et al., 2005; Miyachi et al., 2002; Miyachi et al., 1997). In rodents, an instrumental behavior initiates as goal-directed actions and then transits to stimulus-response habits (Dickinson, 1995; O’Hare et al., 2016), requiring the dorsomedial striatum (DMS; a homologue to the caudate in primates) and the dorsolateral striatum (DLS; a homologue to the putamen in primates), respectively (Yin et al., 2004; Yin et al., 2005). Based on these results, a model of procedural learning has been proposed, in which the functional dominance changes from the associative to sensorimotor subregions in the striatum during the learning phases (Redgrave et al., 2010).

In contrast, deviations from the prior model are evident in other behavioral tasks on decision-making based on external sensory cues. For example, the neural activity in the caudate in humans is associated with learning of a visuomotor association task and the execution of learned behavior (Seger and Cincotta, 2005; Toni and Passingham, 1999). The tail of the caudate in non- human primates contributes to the automatic, habitual behavior in decision-making task using complex visual stimuli (Kim and Hikosaka, 2013). In rodents, the DLS is involved in the acquisiton and execution of the discrimination task with visual and auditory cues, wheras the DMS madiates the performance of learned behavior (Featherstone and McDonald, 2004; Featherstone and McDonald, 2005). Moreover, other striatal subregions, such as the posterior striatum and ventrolateral striatum in rodents are also engaged in the auditory discrimination (Guo et al., 2019; Lee et al., 2020; Xiong et al., 2015). Based on these results, a hypothesis has arisen that the acquisition of sensory cue-based decision-making requires wide-range spatiotemporal processing within the striatum. However, the neural mechanism by which the functional circuits linked to striatal subregions change throughout learning processes is not yet fully understood.

To investigate a large-scale striatal reorganization in accordance with the acquisition of external cue-dependent decisions, we conducted a two-alternative auditory discrimination task with rats (Nishizawa et al., 2012). Using a small-animal neuroimaging technique, we found that the anterior dorsolateral striatum (aDLS) and posterior ventrolateral striatum (pVLS) are activated to the highest level at the middle and late stages during the acquisition phase of auditory discrimination, respectively. Then, chronic pharmacologic manipulations confirmed that the aDLS and pVLS are necessary for discrimination learning, and the transient manipulations also indicated the function of these subregions in the learning at the corresponding stages. The resultant effects of the manipulations on the behavioral strategy showed that the aDLS promotes the strategy based on the stimulus-response association, while suppressing that based on response-outcome associations. Additionally, the pVLS engages in the formation and maintenance of the stimulus-response association strategy. Electrophysiological data of striatal neurons indicated the presence of aDLS subpopulations, mainly representing the outcome of specific behaviors at the initial period of auditory discrimination and pVLS subpopulations encoding the beginning and ending of behavior in association with the development of the discrimination. There were other subpopulations in the aDLS and pVLS showing long-term increased activity following reward with different patterns of the combination between the stimulus and response. These findings revealed that the two striatal subregions are involved in the acquisition of auditory discrimination through distinct spatiotemporal and functional fashions, challenging the prior model which proposed a transition from the associative to sensorimotor striatal subregions during learning.

Results

Leaning processes of auditory discrimination

We employed a two-alternative auditory discrimination task, in which rats are required to make new associations between a tone instruction cue (high frequency tone of 10 kHz or low frequency tone of 2 kHz) and a response (pressing the right or left lever) (Nishizawa et al., 2012), setting the interval between the stimulus presentation and lever insertion to 3 s (Figure 1A). The success rate gradually increased, reaching a plateau around Day 13, which persisted through Day 24 (Figure 1B). Based on changes observed in the success rate, we categorized the learning processes into two phases: the acquisition phase, which lasted until Day 13; and the learned phase, which began after Day 13 and then continued at least until Day 24. In addition, we divided the acquisition phase into three stages depending on the success rate that included the early (< 60%), middle (60–80%), and late (80–100%) stages. We chose Days 2, 6, and 10 at the early, middle, and late stages in the acquisition phase, respectively, as well as Days 13 and 24 in the learned phase for statistical comparison of the success rate. There were significant differences between Days 2 and 6, Days 6 and 10, as well as Days 10 and 13, showing no significant difference between Days 13 and 24 (Figure 1C). In contrast, the response time and omission rate did not vary among the training days (Figure 1D and E), and the variation of response bias decreased along with the progress of auditory discrimination (Figure 1F).

Behavioral performance of two-alternative auditory discrimination learning.

(A) Schematic diagram of the auditory discrimination task. Each trial started by presentation of a tone instruction cue with the high (10 kHz) or low (2 kHz) frequency. Three seconds later, the room light was illuminated, and two retractable levers were inserted at the same time. The rats were required to press the right and left levers in response to the high and low tones, respectively. (B) Learning curve of the auditory discrimination in intact rats (n = 14). Inset indicates the cumulative curve of number achieved the success rate of more than 90%. (C) Success rate at Days 2, 6, 10, 13, and 24 (one-way repeated ANOVA, F[1.897,24.659] = 106.706, p = 1.1 × 10-12, post hoc Bonferroni test, Day 2 vs. Day 6, p = 0.001, Day 6 vs. Day 10, p = 3.6 × 10-4, Day 10 vs. Day 13, p = 0.040, and Day 13 vs. Day 24, p = 0.080). (D) Response time (one-way repeated ANOVA, F[2.203,28.642] = 2.386, p = 0.105). (E) Omission rate (F[4,52] = 0.699, p = 0.596). (F) Response bias (one-way repeated ANOVA, F[1.129,14.674] = 0.250, p = 0.653). Data are indicated as the mean ± s.e.m., and individual data are overlaid. *p < 0.05, **p < 0.01, and ***p < 0.001.

Distinct brain activity patterns in striatal subregions during learning processes

To investigate dynamic changes in regional brain activity in the entire striatum during auditory discrimination in the same animals, we conducted the positron emission tomography (PET) for small animals with 2-deoxy-2-[18F]fluoro-D-glucose (18F-FDG), which measures cerebral glucose metabolism correlated to brain activity (Cui et al., 2015; Kornblum et al., 2000; Zimmer et al., 2017) (Figure 2A). Before the auditory discrimination task, we performed the single lever press task, in which a lever pseudorandomly presented on either the left or right side, and the rats learned to press a lever associated with receiving a reward. This task was used as a control that does not require the decision process based on the auditory stimulus. 18F-FDG-PET scanning was carried out on Day 4 of the single lever press task, and then during a series of training sessions including the acquisition (Days 2, 6, and 10) and learned (Day 24) phases (Figure 2B). The average number of lever presses in the single lever press task exceeded 80 times (81.6 ± 1.3, mean ± s.e.m.) on Day 2 and maintained that level until Day 4 (Figure 2C). In the following auditory discrimination task, the success rate gradually increased along with the training days (Figure 2D), showing a significant difference among sessions in which we performed the 18F-FDG-PET scanning (Figure 2E). The response time and omission rate were consistent among the days, and the variation of response bias became smaller as the task progresses (Figure 2E). These performances were similar to those in the intact rats (compared to Figure 1C–F), suggesting that the procedure for 18F-FDG-PET scans does not affect the acquisition of discrimination.

Dissociable brain activity patterns during auditory discrimination learning among three striatal subregions.

(A) Schematic illustration of an awake rat received intravenously 18F- FDG injection through an indwelling catheter attached to the tail. The rat was conducted for the behavioral experiment and then used for microPET imaging. (B) Schedule for the behavioral training and 18F-FDG-PET scan. After 18F-FDG injection, rats (n = 14 rats) were subjected to the behavioral experiment (30 min) and returned to the home cage (15 min). Ten min before the PET scan started, the rats were anesthetized and then the scan was started (30 min). (C and D) Learning curves of the single lever press task (C) and auditory discrimination task (D). Arrowheads indicate the PET scan days. (E) Behavioral performance on the scan days during the auditory discrimination. Success rate (one-way repeated ANOVA, F[3,39] = 125.012, p = 4.8 × 10-20, post hoc Bonferroni test; Day 2 vs. Day 6, p = 5.0 × 10-7; Day 6 vs. Day 10, p = 0.001; Day 10 vs. Day 24, p = 0.003), response time (one-way repeated ANOVA, F[3,39] = 0.156, p = 0.926), omission rate (one-way repeated ANOVA, F[3,39] = 1.559, p = 0.215), and response bias (one-way repeated ANOVA, F[1.779,23.130] = 17.734, p = 3.6 × 10-5) are shown. (F) Representative images of coronal sections that compared the brain activity in the single lever press task with the activity on Day 2, 6, 10, or 24 in the discrimination task. Left panels show schematic illustrations of striatal subregions. (G) Horizontal (top) and coronal (bottom) images of striatal activation areas shown on Day 10 vs. single lever in (F). (H) Representative images of coronal section that compared the brain activity on Day 2 with that on Day 6, 10, or 24 in the discrimination task. Color bars indicate the T-values, and a value of 3.8 was used as the threshold corresponding to the uncorrected threshold (p < 0.001). (I) Schematic pictures showing voxel of interests for the aDLS, pVLS, and DMS. (J-L) Regional 18F- FDG uptakes in the aDLS (J, one-way repeated ANOVA, left aDLS, F[4,52] = 10.322, p = 3.0 × 10- 6, post hoc Bonferroni test; single lever vs. Day 6, p = 0.016; Day 2 vs. Day 6, p = 6.5 × 10-5; Day 2 vs. Day 10, p = 0.019; Day 2 vs. Day 24, p = 0.017; Day 6 vs. Day 24, p = 0.041; right aDLS, F[4,52] = 7.462, p = 7.9 × 10-5, post hoc Bonferroni test, Day 2 vs. Day 6, p = 2.9 × 10-4; Day 2 vs. Day 10, p = 0.036; Day 6 vs. Day 24, p = 0.005), pVLS (K, one-way repeated ANOVA, left pVLS, F[2.368,30.784] = 4.152, p = 0.020, post hoc Bonferroni test; single lever vs. Day 10, p = 1.8 × 10- 5; right pVLS, F[4,52] = 5.995, p = 4.8 × 10-4, post hoc Bonferroni test, single lever vs. Day 10, p = 0.001; Day 2 vs. Day 10, p = 0.011; Day 6 vs. Day 10, p = 0.032), and DMS (L, one-way repeated ANOVA, left DMS, F[4,52] = 12.836, p = 2.4 × 10-7, post hoc Bonferroni test; Day 2 vs. Day 24, p = 1.8 × 10-5; Day 6 vs. Day 24, p = 4.8 × 10-4; Day 10 vs. Day 24, p = 5.5 × 10-5; right DMS, F[4,52] = 10.717, p = 2.0 × 10-6, post hoc Bonferroni test; single lever vs. Day 24, p = 0.036; Day 2 vs. Day 24, p = 0.011; Day 6 vs. Day 24, p = 2.6 × 10-4; Day 10 vs. Day 24, p = 0.011). Arrowheads indicate the day with the most activations throughout the learning process. Data are indicated as the mean ± s.e.m., and individual data are overlaid. The anteroposterior coordinates from bregma (mm) are shown (F–I). Scale bar; 2 mm (I). *p < 0.05, **p < 0.01, and ***p < 0.001.

We investigated the brain regions, in which 18F-FDG uptake significantly increased or decreased during the learning processes, applying a statistical significance threshold (p < 0.001, uncorrected) and an extent threshold, by using a voxel-based statistical parametric analysis (Cui et al., 2015). We first analyzed task-related brain activity by comparing 18F-FDG uptake on either day of the discrimination task with that on Day 4 of the single lever press task (summarized in Figure 2–table supplement 1). In the striatal subregions, there were significant increases of the activity in the unilateral aDLS and the bilateral pVLS on Days 6 and 10, respectively (Figure 2F and G).

However, no significant change was observed in the DMS during the acquisition phase except a unilateral decrease on Day 24 (Figure 2F). Next, to evaluate the activity related to the progress of learning, 18F-FDG uptake on Days 6, 10, and 24 was compared with that on Day 2. The learning- dependent brain activity was significantly increased in the bilateral aDLS and the unilateral pVLS on Days 6 and 10, respectively (Figure 2H), whereas the activity decreased in the bilateral DMS on Day 24 (Figure 2H). This decrease was also detected when 18F-FDG uptakes were compared between Days 10 and 24 (Figure 2H), suggesting the down-regulation of the DMS activity in the learned phase.

To quantitatively validate changes in the brain activity in striatal subregions, we analyzed the amount of 18F-FDG uptake in the voxel of interest of these subregions (Figure 2I). The 18F- FDG uptake in the bilateral aDLS reached a peak on Day 6, and it decreased during the subsequent days (Figure 2J), whereas the uptake in the bilateral pVLS gradually increased along with the progress of learning, showing the maximal value on Day 10 (Figure 2K). The uptake in the DMS did not alter during the acquisition phase, and it decreased on Day 24 (Figure 2L). These results suggest that the aDLS and pVLS contribute to learning processes in a different temporal pattern during the acquisition phase, whereas the DMS may be engaged to the behavior in the learned phase.

Excitotoxic lesion of the aDLS and pVLS, but not the DMS, disrupts the acquisition of auditory discrimination

To study whether the changed activity of striatal subregions is involved in learning processes, rats received a bilateral injection of solution containing ibotenic acid (IBO; 8 mg/mL, 0.3 μL/site) or phosphate buffered saline (PBS) into the subregions, and we conducted for the behavioral study. The range of IBO lesion in the subregions was analyzed by immunostaining with a neuronal marker NeuN after the behavioral tests (Figure 3A–C). The number of single lever presses increased similarly along with the training in both the PBS- and IBO-injected groups into the aDLS, pVLS, or DMS along with the training (Figure 3D–F). For the aDLS injections, increases in success rate were impaired during the acquisition phase in the IBO group, compared to the PBS group, and the impairments were mainly observed at the middle stage (Figure 3G). For the pVLS injection, success rate impairments were observed in the IBO group showed compared to the PBS group, although the period showing a significant difference was unclear (Figure 3H). The response time in the aDLS-lesioned group did not alter during the discrimination training, whereas that in the pVLS- lesioned group was continuously prolonged (Figure 3–figure supplement 1A and B), and the omission rate and response bias were similar between the lesioned and the corresponding control groups (Figure 3–figure supplement 1D and G for aDLS lesion, and Figure 3 –figure supplement 1E and H for pVLS lesion). In contrast, for the DMS injection the success rate in the discrimination task was comparable between the PBS and IBO groups (Figure 3I). The DMS lesion lengthened the response time through the learning (Figure 3–figure supplement 1C), whereas it did not alter the omission rate and response bias (Figure 3–figure supplement 1F and I). These results indicate that the aDLS and pVLS are necessary for the acquisition of auditory discrimination, and that the aDLS appears to be mainly involved in the learning process at the middle stage. The results also reveal that the DMS contributes to the execution, but not learning, of discriminative behavior.

Impacts of excitotoxic lesion of striatal subregions on the acquisition of auditory discrimination.

Rats were given intracranial injection of PBS or IBO solution into the aDLS (n = 8 for each injection), pVLS (n = 8 for each injection), or DMS (n = 7 for PBS injection, and n = 6 for IBO injection) before start of the single lever press task. (A-C) Representative images of NeuN immunostaining and individual schematic illustrations showing lesioned area in the aDLS (A), pVLS (B), or DMS (C). Dotted lines in the images indicate the range of the lesioned area. ac, anterior commissure; cc, corpus callosum; and LV, lateral ventricle. (D-F) Learning curves during the single lever press task in the groups injected into the aDLS (D, two-way repeated ANOVA, group, F[1,14] = 1.275, p = 0.278, day, F[1.512,21.171] = 26.281, p = 7.0 × 10-6, group × day, F[1.512,21.171] = 0.834, p = 0.418), pVLS (E, two-way repeated ANOVA, group, F[1,14] = 2.542, p = 0.133, day, F[1.547,21.658] = 40.183, p = 2.2 × 10-7, group × day, F[1.547,21.658] = 0.953, p = 0.380), or DMS (F, two-way repeated ANOVA, group, F[1,11] = 0.025, p = 0.876, day, F[1.205,13.252] = 14.818, p = 0.001, group × day, F[1.205,13.252] = 0.610, p = 0.478). (G-I) Learning curves during the auditory discrimination task in the groups injected into the aDLS (G, two-way repeated ANOVA, group, F[1,14] = 11.578, p = 0.004, day, F[2.805,39.274] = 60.733, p = 1.8 × 10-14, group × day, F[2.805,39.274] = 3.863, p = 0.018, simple main effect, PBS vs. IBO, Day 1, p = 0.140, Day 2, p = 0.085, Day 3, p = 0.284, Day 4, p = 0.068, Day 5, p = 0.011, Day 6, p = 9.5 × 10-5, Day 7, p = 0.001, Day 8, p = 0.002, Day 9, p = 0.012, Day 10, p = 0.059, Day 11, p = 0.034, Day 12, p = 0.046, Day 13, p = 0.136, Day 14, p = 0.096), pVLS (H, two-way repeated ANOVA, group, F[1,14] = 7.221, p = 0.018, day, F[3.646,51.041] = 51.794, p = 9.2 × 10-17, group × day, F[3.646,51.041] = 1.612, p = 0.190), or DMS (I, two-way repeated ANOVA, group, F[1,11] = 0.469, p = 0.507, day, F[2.677,29.452] = 48.860, p = 3.9 × 10-11, group × day, F[2.677,29.452] = 1.094, p = 0.362). (J-L) Changes in the proportions of the win-shift-win or lose-shift-lose strategies through the discrimination learning in the groups treated into the aDLS (J, two-way repeated ANOVA, group, F[1,14] = 13.451, p = 0.003, day, F[3.048,42.673] = 54.777, p = 9.1 × 10-15, group × day, F[3.048,42.673] = 3.288, p = 0.029, simple main effect, PBS vs. IBO, Day 1, p = 0.728, Day 2, p = 0.225, Day 3, p = 0.165, Day 4, p = 0.044, Day 5, p = 0.015, Day 6, p = 1.1 × 10-4, Day 7, p = 0.002, Day 8, p = 5.1 × 10-4, Day 9, p = 0.004, Day 10, p = 0.060, Day 11, p = 0.053, Day 12, p = 0.081, Day 13, p = 0.198, Day 14, p = 0.075 for win-shift-win; group, F[1,14] = 5.039, p = 0.041, day, F[3.866,54.119] = 14.578, p = 5.2 × 10-8, group × day, F[3.866,54.119] = 2.471, p = 0.057 for lose-shift-lose), pVLS (K, two-way repeated ANOVA, group, F[1,14] = 9.251, p = 0.009, day, F[3.206,44.881] = 47.168, p = 2.7 × 10-14, group × day, F[3.206,44.881] = 1.862, p = 0.146 for win- shift-win; group, F[1,14] = 0.762, p = 0.397, day, F[4.773,66.815] = 24.018, p = 1.9 × 10-13, group × day, F[4.773,66.815] = 1.338, p = 0.260 for lose-shift-lose), or DMS (L, two-way repeated ANOVA, group, F[1,11] = 0.703, p = 0.420, day, F[3.052,33.568] = 48.424, p = 2.3 × 10-12, group × day, F[3.052,33.568] = 0.846, p = 0.480 for win-shift-win; group, F[1,11] = 0.002, p = 0.965, day, F[3.442,37.866] = 13.787, p = 1.0 × 10-6, group × day, F[3.442, 37.866] = 0.581, p = 0.654 for lose-shift-lose). Data are indicated as the mean ± s.e.m., and individual data are overlaid (except for panels J–L). The anteroposterior coordinates from bregma (mm) are shown (A–C). Scale bars; 2 mm (A–C). *p < 0.05, **p < 0.01 and ***p < 0.001. Abbreviations: WSW, win-shift-win; and LSL, lose-shift-lose.

Previous studies have reported that the dorsal striatum is involved in both behavioral strategies based on the stimulus-response association and the response-outcome association and during instrumental learning (Dickinson, 1985; O’Hare et al., 2016). To assess the impact of striatal lesions on the behavioral strategies, we analyzed the proportions of responses attributed to two strategies in all responses of each session. One is the “win-shift-win” strategy, which is considered to reflect the behavioral strategy based on the stimulus-response association (Bergstrom et al., 2018; Packard et al., 1989). In this strategy, after a correct response in the previous trial, the rats press the opposite lever in the current trial in response to a shift of the instruction cue, resulting in the correct response. Another strategy is the “lose-shift-lose” strategy, which is considered to appear as a consequence of the behavioral strategy based on the response-outcome association (Bergstrom et al., 2018). In this strategy, after an error response in the previous trial, the rats press the opposite lever in the current trial, despite a shift of the instruction cue, leading to another error response. In the control group into either of three striatal subregions, the proportions of win-shift- win and lose-shift-lose strategies increased and decreased, respectively, during the acquisition phase (Figure 3J–L). The changes in the proportion of the two strategies were similar to those in the intact rats (Figure 1–figure supplement 1). For the aDLS injection, the increases in the win-shift- win proportion were markedly impaired in the IBO group compared to the PBS group with significant differences mainly at the middle stage, and the decrease in the lose-shift-lose proportion was delayed compared to the same group (Figure 3J). For the pVLS-injection, the increasing win- shift-win proportion was also impaired in the IBO group compared to the PBS group, although the change in the lose-shift-lose proportion was comparable between the two groups (Figure 3K). In contrast, for the DMS lesion both win-shift-win and lose-shift-lose proportions in the IBO group did not show significant differences from the corresponding values in their control group (Figure 3L). These results indicate that the aDLS promotes the win-shift-win strategy through stimulus- response association and suppresses the lose-shift-lose strategy rooted in response-outcome association, and that the pVLS selectively facilitates the win-shift-win strategy acquisition. Thus, these two subregions not only play roles in the acquisition of auditory discrimination but also control distinct behavioral strategies during the learning processes.

After the completion of discrimination training, we tested feeding behavior and locomotion of rats with aDLS and pVLS lesions. There was no difference in the amount of food intake and locomotor activity in the open field in both lesioned groups (Figure 3–figure supplement 2A-D), suggesting that learning deficits following excitotoxic lesions of the striatal subregions cannot be attributable to abnormalities in these general motor behaviors.

Inhibiting aDLS and pVLS activity impairs discriminative behavior at different stages of the acquisition phase

Our 18F-FDG-PET imaging study indicated that the brain activity in the aDLS and pVLS was increased in a different temporal pattern during the acquisition phase of auditory discrimination. To examine whether neuronal activity in striatal subregions is linked with the processes at different stages, we temporarily inhibited the activity in each subregion by the treatment with a GABAA receptor agonist muscimol (MUS). In a pilot study, we tested the effect of the bilateral injection of several doses of MUS (0.1, 0.2, and 0.5 μg/μL) into the aDLS or pVLS on the single lever press behavior and confirmed that the administration of the lowest dose of MUS did not change the frequency of single lever presses as compared to the saline (SAL) injection (Figure 4–figure supplement 1A and B). The lowest dose of MUS into the two striatal subregions unaffected the amount of food intake (Figure 4–figure supplement 1C and D). In addition, to achieve the stage- specific inactivation of the aDLS or pVLS neurons, rats were given bilateral injections of MUS (0.1 μg/μL, 0.25 μL per site) or SAL at three different timings during a series of the training. The first injection was conducted on Day 2 to target the early stage. The second and third injections were conducted on the day after the success rate had reached 60% and 80% for the first time through the training, respectively, to target the middle or late stage. After the injections, the rats were used for the behavioral test. Placement locations of the injection cannula into the striatal subregions were assessed by cresyl violet staining after the behavioral tests (Figure 4A and B). The success rates on the day before the injection (Day N-1) at the early, middle, and late stages were similar between the two groups that received the bilateral injections of MUS or SAL (Figure 4–figure supplement 2A and B). Rats were then administered injections into the aDLS on the next day (Day N), and the success rate in the MUS-injected groups significantly decreased at the middle stage compared to the corresponding SAL-injected groups, but not at the early and late stages (Figure 4C, left). In contrast, the success rate in the MUS groups into the pVLS significantly decreased at the late stage, but not at the early and middle stages, compared with the respective SAL groups (Figure 4D, left). To normalize the variation of learning speed in the individual animals, we calculated the value by subtracting the success rate on Day N-1 from that on Day N in the individual animals. There were significant reductions in the subtracted rate at the middle stage in the MUS-injected group into the aDLS (Figure 4C, right) and at the late stage in the MUS-injected group into the pVLS (Figure 4D, right). These results indicate that the aDLS and pVLS mainly function at the middle and late stages, respectively, in the acquisition phase of auditory discrimination, supporting the results obtained from our imaging study.

Influence of transient inhibition of striatal subregions at different timings on the performance of auditory discrimination.

Rats received intracranial injection of SAL or MUS solution into the aDLS (n = 8 for each injection) or pVLS (n = 6 for each injection). (A and B) Representative images of cresyl violet staining and schematic illustrations showing placement sites of the tip of guide cannula in the aDLS (A) or pVLS (B). Dotted lines in the images indicate the position of the cannula placement. ac, anterior commissure; cc, corpus callosum; and LV, lateral ventricle. (C and D) Effects of transient striatal inhibition on the performance. For aDLS inhibition (C), success rate (left, two-way repeated ANOVA, stage, F[2,28] = 28.708, p = 1.7 × 10-7, group, F[1,14] = 8.840, p = 0.010, stage × group, F[2,28] = 0.757, p = 0.478; unpaired Student’s t-test, early, t[14] = 0.558, p = 0.586; middle, t[14] = 2.764, p = 0.015; Welch’s t-test, late, t[12.010] = 1.617, p = 0.132, p = 0.016 after a Bonferroni correction method) and subtracted success rate on day N-1 from that on day N (right, two-way repeated ANOVA, stage, F[2,28] = 1.231, p = 0.307, group, F[1,14] = 10.797, p = 0.005, stage × group, F[2,28] = 2.360, p = 0.113; unpaired Student’s t- test, early, t[14] = 0.285, p = 0.780, late, t[14] = 1.271, p = 0.225, Welch’s t-test, middle, t[9.603] = 3.331, p = 0.008, p = 0.016 after a Bonferroni correction method). For pVLS inhibition (D), success rate (left, two-way repeated ANOVA, stage, F[2,20] = 17.642, p = 3.8 × 10-5, group, F[1,10] = 43.942, p = 5.9 × 10-5, stage × group, F[2,20] = 4.729, p = 0.012; simple main effect, early, p = 0.055, middle, p = 0.109, late, p = 4.2 × 10-5) and subtracted success rate on day N-1 from that on day N (right, two-way repeated ANOVA, stage, F[2,20] = 12.105, p = 3.6 × 10-4, group, F[1,10] = 16.310, p = 0.002, stage × group, F[2,20] = 8.500, p = 0.002; simple main effect, early, p = 0.595, middle, p = 0.085, late, p = 3.0 × 10-6). (E and F) Subtracted proportion of the win-shift-win or lose- shift-lose strategies in the groups injected into the aDLS at the middle stage (E, unpaired Student’s t-test, t[14] = 2.038, p = 0.061 for win-shift-win, and t[14] = 2.714, p = 0.017 for lose-shift-lose) and the late stage (F, unpaired Student’s t-test, t[14] = 0.898, p = 0.384 for win-shift-win, and t[14] = 0.226, p = 0.824 for lose-shift-lose). (G and H) Subtracted proportion of the win-shift-win or lose- shift-lose strategies in the groups injected into the pVLS at the middle stage (G, unpaired Student’s t-test, t(10) = 1.924, p = 0.083 for win-shift-win, and Welch’s t-test, t[6.095] = 1.364, p = 0.221 for lose-shift-lose) and the late stage (H, unpaired Student’s t-test, t[10] = 3.629, p = 0.005 for win- shift-win, and t[10] = 1.577, p = 0.146 for lose-shift-lose). Data are indicated as the mean ± s.e.m., and individual data are overlaid. The anteroposterior coordinates from bregma (mm) are shown (A and B). Scale bars; 2 mm (A and B). *p < 0.05, **p < 0.01, and ***p < 0.001. Abbreviations: WSW, win-shift-win; and LSL, lose-shift-lose.

Next, we analyzed the influence of temporal striatal inactivation on the behavioral strategy at the middle and late stages of the acquisition phase. We calculated the proportions of the win- shift-win and lose-shift-lose strategies, and subtracted the values of these proportions on Day N-1 from those on Day N. The proportion of both strategies on Day N-1 were comparable between the MUS- and SAL-injected groups (Figure 4–figure supplement 3A and B). In the aDLS, the subtracted proportion of the lose-shift-lose strategy at the middle stage significantly increased in the MUS group as compared to the SAL group, although the subtracted proportion of the win-shift-win strategy tended to decrease in the MUS group as compared to the control group (Figure 4E), and the subtracted proportions of the win-shift-win or lose-shift-lose strategy at the late stage was similar between both groups (Figure 4F). In the pVLS, the subtracted proportions of the win-shift- win and lose-shift-lose strategies at the middle stage were similar between the MUS and SAL groups (Figure 4G), and the subtracted win-shift-win proportion at the late stage significantly decreased in the MUS group as compared to the SAL group, although the subtracted lose-shift-lose proportion was comparable between the two groups (Figure 4H). Our results show that transient inhibition of aDLS activity at the middle stage affects the lose-shift-lose strategy and has a subtle effect on the win-shift-win strategy, supporting that the aDLS mediates the acquisition process of auditory discrimination at the middle stage at least partly through the suppression of the behavior based on the response-outcome association. Although the impact of inhibition on the win-shift-win strategy was subtle, excitotoxic lesion of the aDLS influenced the win-shift-win strategy in addition to the lose-shift-lose strategy (see Figure 3J), suggesting that the aDLS acts to enhance the auditory discrimination by promoting the stimulus-response association, together with the suppression of the response-outcome association. The discrepancy in the extent of changes in the win-shift-win strategy between the two treatments may be generated from the difference in the timing or sustainability of drug inhibition of the striatal subregion. In addition, our results show that pVLS inactivation at the late stage selectively impairs the win-shift-win strategy, supporting contribution of the pVLS to the process via progression of the stimulus-response association.

Neuronal activity in the aDLS and pVLS during the acquisition phase

To examine the firing activity in striatal subregions during the acquisition phase of auditory discrimination, we simultaneously recorded neuronal activity in the aDLS and pVLS at the unilateral (left) side by using a multi-unit recording procedure (Figure 5A). Using immunostaining for tyrosine hydroxylase (TH) after the recordings, we verified the location of electrodes within the striatal subregions (Figure 5B). After the single lever press task, we conducted the auditory discrimination task for the recordings. In correct trials, the delay period (0.5 ± 0.2 s) was inserted between a lever press and a sound indicating the reward to distinguish neuronal activity related to the two events (Figure 5C). The behavioral data in individual rats were divided into three grouped sessions corresponding to the early, middle, and late stages, based on a sigmoid function analysis (Bathellier et al., 2013) of the success rate in the task. The success rates in the grouped sessions were 52.7 ± 0.8 %, 61.6 ± 0.7 %, and 85.3 ± 1.7 % (mean ± s.e.m.) for the early, middle, and late stages, respectively, and the rate was significantly increased throughout the three stages (Figure 5– figure supplement 1A). However, the response time, omission rate, and response bias were similar among the stages (Figure 5–figure supplement 1B-D). Thus, these stages corresponded to the timing window defined in the acquisition phase of the transient MUS inhibition experiment of neuronal activity.

Multi-unit recording of neurons in striatal subregions and firing activity related to the behavioral outcome of reward sound-related HR type neurons in the aDLS.

(A) Schematic illustration of a freely moving rat used for simultaneous multi-unit recordings in the aDLS and pVLS. Six rats were used for the following analysis. (B) Representative images of TH immunostaining, showing positions of electrode tips (arrows) in the aDLS and pVLS (left rows). The recording sites estimated by electrode tracks and electrical marks in individual rats are shown (center and right rows). ac, anterior commissure; LV, lateral ventricle. (C) Sequence of some events in correct and error trials. The delay periods were pseudorandomly added between a correct lever press and the reward sound (0.5 ± 0.2 s). The room light turned off the extended time (4 s) after the correct response or immediately after the error response. (D) Mean firing rate (top rows) and auROC values (bottom rows) of reward sound-related HR type neurons in the aDLS at the early (n = 43 neurons), middle (n = 37 neurons), and late (n = 44 neurons) stages. These data are aligned to the choice response. Green shadows show the period when the reward sound is presented (500 ± 200 ms after the lever press). Time bins with significant differences between the mean firing rate in the HR or LR trial and any of the rates in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. (E) Averaged firing rate during the reward sound period in the HR and LR trials at the early, middle, and late stages (Kruskal-Wallis test: HR trial, χ2 = 3.028, p = 0.220; and LR trial, χ2 = 1.249, p = 0.536). (F) Cumulative probability of the proportion of reward sound-related HR type neuron number at the three stages (Kruskal-Wallis test, χ2 = 45.378, p = 1.4 × 10-10; post hoc Tukey- Kramer test, p = 7.1 × 10-5 for early vs. middle, p = 1.0 × 10-9 for early vs. late, and p = 0.040 for middle vs. late). Data are indicated as the mean ± s.e.m. (D) or the median and quartiles with the maximal and minimal values (E). The anteroposterior coordinates from bregma (mm) are shown (B). Scale bars; 2 mm (B). *p < 0.05 and ***p < 0.001.

We identified 1,062 and 423 well-isolated neurons from the aDLS and pVLS, respectively, across the three stages (Figure 5–figure supplement 2A) and focused on the major population of striatal neurons, putative medium spiny neurons (MSNs; Figure 5–figure supplement 2B). The numbers of MSNs were 96% and 91% of the total numbers of identified neurons in the aDLS and pVLS, respectively (Figure 5-figure supplement 2C). The neuron numbers at the early, middle, and late stages were 295, 364, and 360 neurons in the aDLS, and 87, 110, and 190 neurons in the pVLS, respectively. Based on the combinatorial pattern of the tone instruction cue and lever press in our discrimination task, we categorized the electrophysiological data into four trial types: (1) high- frequency tone/right lever press (HR) and (2) low-frequency tone/left lever press (LL) as correct responses; and (3) high-frequency tone/left lever press (HL) and (4) low-frequency tone/right lever press (LR) as error responses (see Figure 1A). We identified HR or LL type neurons showing significant changes in firing rate related to the cue onset, choice response, reward sound, or first licking as compared to the baseline firing rate in each of two trials (Mann-Whitney U test, p < 0.01). These neurons were further divided into two groups based on increased or decreased activity (Figure 5–figure supplement 3A and B). In the following analyses, we focused on the HR and LL type neurons with increased event-related activity to explore the firing patterns of neurons underlying behavioral processes of auditory discrimination.

Subpopulations of aDLS neurons, but not of pVLS neurons, show firing activity related to the behavioral outcome

Previous studies have reported that neurons in the dorsal striatum show the outcome-related activity in the instrumental conditioning tasks (Guo et al., 2019; Nonomura et al., 2018). We thus investigated whether aDLS and pVLS neurons exhibit a response related to the behavioral outcome during the acquisition of auditory discrimination. In the aDLS, the firing activity of reward sound- related HR type neurons showed a higher extent in the increase during a period, when the reward sound is presented in the correct responses, in the HR trial as compared with the LL, HL, or LR trial throughout the three stages (Figure 5D, top rows). The rate of these neurons also moderately increased during the same period in the LR trial compared to the LL or HL trial across the three stages, although the extent of the increase was lower than that in the HR trial (Figure 5D, top rows). The area under the receiver operating characteristic (auROC) values revealed a significant difference of the HR trial from the LR trial during the reward sound period, whereas the values of LL and HL trials indicated a significant difference from the LR trial in the opposite direction during the same period (Figure 5D, bottom rows). The averaged firing rate during the reward sound period in the HR or LR trial appeared to modestly increase at the middle stage, but no significant difference existed among the stages (Figure 5E). To directly compare the number of neurons showing event-related activity among different stages, we calculated the number of reward sound- HR type neurons by matching the trial number at each stage (Figure 5F). The proportion decreased progressively from the early to middle stages, and to the late stage. In addition, the firing rate of reward sound-related LL type neurons exhibited a higher increase in the LL trial and a modest increase in the HL trial while the averaged firing rate during the reward sound period was similar among the three stages (Figure 5–figure supplement 4A and B). Furthermore, the proportion of reward sound-LL neurons reduced progressively through the stages (Figure 5–figure supplement 4C). These results indicate that the aDLS contained neurons related to the outcome of choice, showing a greater increase in the firing activity for the correct responses (HR and LL trials) than the error responses (HL and LR trials), and that the proportion of these subpopulations was gradually reduced through the progress of learning. In contrast, during the reward sound period at all stages, the firing rate and auROC values of reward sound-HR and reward sound-LL type neurons in the pVLS revealed no significant differences across the four kinds of trial (Figure 5–figure supplement 5A and B), indicating that the firing activity of pVLS neurons did not appear to be associated with the behavioral outcome.

Subpopulations of aDLS and pVLS neurons show sustained firing activity after the reward in different manners

The dorsal striatum is known to demonstrate a response of neuronal activity after the positive outcome in the instrumental behaviors (Smith and Graybiel, 2016; Williams and Eskandar, 2006). We thus examined whether aDLS and pVLS neurons show the response after obtaining the reward during the acquisition phase of auditory discrimination. In the aDLS, the firing rate of the first licking-related HR type neurons indicated a sustained elevation after the first licking in the HR trial as compared with any of the other three trials across the early, middle, and late stages (Figure 6A, top rows). The auROC values revealed a significant difference of the LL, HL, or LR trial from the HR trial across the three stages (Figure 6A, middle rows). In addition, we calculated the averaged firing rate and increased firing period, defined as the total number of time bins above 3 s.d. of the baseline firing rate for 5 s after the first licking in the HR trial. The averaged firing rate was comparable between the early and middle stages, but the rate was significantly lower at the late stage compared to either of the other two stages (Figure 6B). The increased firing period was also comparable between the early and middle stages, showing a significant reduction at the late stage (Figure 6C). The proportions of first licking-HR type neurons were similar between the early and late stages, although they showed a slight decrease at the middle stage (Figure 6D). There were only a few neurons with significant correlations between the firing rate and licking performance (Figure 6E). In addition, the firing rate of the first licking-related LL type neurons exhibited the largest increase in the LL trial at the three stages, showing a tendency to reduce the average firing rate and increased firing period after the first licking across the stages (Figure 6–figure supplement 1A-C). The proportion of first licking-LL type neurons during the learning processes displayed similar changes as observed in the first licking-HR neurons, and there were few neurons with significant correlations between the firing rate and licking performance (Figure 6–figure supplement 1D and E). These results indicate that the aDLS contained neurons showing long- lasting, increased activity after the reward in the specific correct combination of the stimulus and response (HR or LL trial), and that the extent of firing activity and length of firing period in the response appeared to be gradually reduced with no correlation between the proportion of neurons and the progress of the learning stages.

Sustained activity after the reward of first licking-related HR type neurons in the aDLS and pVLS.

(A) Firing rate (top rows), auROC (middle rows), and licking rate (bottom rows) after the reward of first licking-related HR type neurons in the aDLS at the early (n = 27 neurons), middle (n = 31 neurons), and late (n = 49 neurons) stages. These data are aligned to the choice response. Timings of the reward sound and first licking are shown. Time bins with significant differences between the mean firing rate in the HR trial and any of the rate in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. (B) Averaged firing rate for 5 s after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 16.396, p = 2.8 × 10-4, post hoc Tukey-Kramer test, early vs. middle, p = 0.726, early vs. late, p = 7.2 × 10-4, middle vs. late, p = 0.009). (C) Averaged total number of time bins above 3 s.d. of the baseline firing for 5 s after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 12.803, p = 0.002, post hoc Tukey-Kramer test, early vs. middle, p = 0.173, early vs. late, p = 0.001, middle vs. late, p = 0.219). (D) Cumulative probability of the proportion of first licking-related HR type neuron number (Kruskal-Wallis test, χ2 = 39.280, p = 3.0 × 10-9, post hoc Tukey-Kramer test, early vs. middle, p = 1.3 × 10-7, early vs. late, p = 0.995, middle vs. late, p = 2.3 × 10-7). (E) Distribution of the correlation coefficient between the numbers of spikes and licking. Closed column indicates the number of neurons showing significant correlations between the two parameters (5 out of 107 neurons). (F) Firing rate (top rows), auROC (middle rows), and licking rate (bottom rows) after the reward of first licking-related HR neurons in the pVLS at the early (n = 17 neurons), middle (n = 11 neurons), and late (n = 64 neurons) stages. The data are aligned to the choice response, and the timings of the reward sound and first licking are shown. Time bins with significant differences between the rate in the HR trial and any of the rate in other trials or the distribution of auROC values and 0.5 are represented as above. (G) Averaged firing rate of pVLS neurons after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 7.652, p = 0.022; post hoc Tukey-Kramer test, early vs. middle, p = 0.975, early vs. late, p = 0.076, and middle vs. late, p = 0.096). (H) Averaged total number of time bins above 3 s.d. of the baseline firing of pVLS neurons after the first licking in the HR trial (Kruskal-Wallis test, χ2 = 0.023, p = 0.989). (I) Cumulative probability of the proportion of first licking-related HR type neuron number (Kruskal-Wallis test, χ2 = 47.922, p = 3.9 × 10-11; post hoc Tukey-Kramer test, early vs. middle, p = 0.011, early vs. late, p = 1.7 × 10-4, and middle vs. late, p = 9.7 × 10-10). (J) Distribution of the correlation coefficient between the number of spikes and the licking. Closed column indicates the number of neurons showing significant correlations between the two parameters (5 out of 92 neurons). Data are indicated as the mean ± s.e.m. (A and F) or the median and quartiles with the maximal and minimal values (B, C, G, and H). **p < 0.01 and ***p < 0.001. Abbreviations: RS, reward sound; and FL, first licking.

In the pVLS, the first licking-HR type neurons showed a sustained elevation in the firing rate after receiving a reward in the HR trial compared to that in the HL or LR trial throughout the three stages (Figure 6F, top rows). A similar elevation in the firing rate was also observed in the LL trial throughout the stages (Figure 6F, top rows). The auROC values revealed a significant difference of the HL or LR trial, but not the LL trial, from the HR trial throughout the stages (Figure 6F, middle rows). The average firing rate did not show significant changes among the three stages (Figure 6G), and the increased firing period was also similar across the stages (Figure 6H).

The proportion of first licking-HR neurons increased at the late stage compared to the early or middle stage (Figure 6I). The firing rate showed almost no significant correlation to the licking performance (Figure 6J). In addition, the first licking-LL neurons exhibited similar firing patterns and changes in the proportion of neurons across the stages (Figure 6–figure supplement 2A-E).

Therefore, the pVLS included neurons representing the long-lasting increased activity after the reward in the two kinds of correct responses (HR and LL trials) with the increased number of these neurons at the late stage of learning.

Neurons in the pVLS, but not in the aDLS, exhibit firing patterns related to the beginning and ending of the behavior

The dorsal striatum plays a key role in the acquisition and execution of behaviors based on the stimulus-response association, and the association is accompanied by a transient activation of neuronal firing immediately after the stimulus presentation or action in trials, which is considered to represent the beginning and ending of a learned behavior (Barnes et al., 2005; Jin and Costa, 2010; Jog et al., 1999). We first examined whether the firing activity of cue onset-related HR and LL type neurons corresponds to the beginning of the behavior in the auditory discrimination task. In the aDLS, the firing rate of both types of neurons showed a significant difference at some time points in each trial compared to those in any of the other three trials, with a significant difference of auROC values between the trials (Figure 7–figure supplement 1A and B). However, the transiently increased activity did not emerge at the time points corresponding to the beginning of the behavior. In contrast, the cue onset-related HR type neurons in the pVLS showed a transient increase in the firing rate immediately after the cue onset, and the activity appeared at the middle and late stages (Figure 7A, top rows). The transient activation was observed not only in the HR trial but also in the other three trials, being consistent with the results of a previous report (Barnes et al., 2011). The increased firing rate showed a slight but notable difference between the HR and LL trials with a significant difference in the auROC values between the two kinds of trials (Figure 7A, bottom rows), suggesting that different combinations between the stimulus and response may affect the level of firing activity. The rate differences between the two trials were observed at the middle and late stages (Figure 7B). The proportion of cue onset-HR neurons increased at the middle and late stages compared to the early stage (Figure 7C). In addition, the firing rate of the cue onset-related LL type neurons exhibited patterns similar to those observed in the cue onset-HR neurons (Figure 7–figure supplement 2A). The data indicate that subpopulations of pVLS neurons show firing activity related to the beginning of the behavior, with a rising number of neurons throughout discrimination learning.

Transient activity related to the beginning and ending of a behavior of cue onset- and choice response-related HR type neurons in the pVLS.

(A) Firing rate (top rows) and auROC values (bottom rows) of cue onset-related HR type neurons in the pVLS at the early (n = 4 neurons), middle (n = 21 neurons), and late (n = 24 neurons) stages. These data are aligned to the cue onset. Time bins with significant differences between the mean firing rate in the HR trial and either of the rate in other three trials (Wilcoxon signed rank test) or the distribution of auROC values and 0.5 (Wilcoxon rank-sum test) are represented by the circles at the top. (B) Averaged firing rate of the cue onset-related HR type neurons in HR and LL trials (0-600 ms after the cue onset) (Wilcoxon signed rank test; early, p = 0.875, middle, p = 0.006, and late, p = 0.004). (C) Cumulative probability of the proportion of the cue onset-related HR type neuron number (Kruskal- Wallis test, χ2 = 51.114, p = 8.0 × 10-12; post hoc Tukey-Kramer test, early vs. middle, p = 9.6 × 10- 10, early vs. late, p = 0.002, and middle vs. late, p = 6.4 × 10-4). (D) Firing rate (top rows) and auROC values (bottom rows) of the choice response-related HR type neurons in the pVLS at the early (n = 11 neurons), middle (n = 22 neurons), and late (n = 41 neurons) stages. These data are aligned to the choice response. The timing of the reward sound is presented. Time bins with significant differences between the mean rate in the HR trial and any of the rate in other trials or the distribution of auROC values and 0.5 are represented as above. (E) Averaged firing rate of the choice response-related HR type neurons in HR and LL trials (0-600 ms after the choice response) (Wilcoxon signed rank test; early, p = 0.278, middle, p = 0.758, and late, p = 0.279). (F) Cumulative probability of the proportion of the choice response-related HR type neuron number (Kruskal- Wallis test, χ2 = 13.786, p = 0.001; post hoc Tukey-Kramer test, early vs. middle, p = 0.003, early vs. late, p = 0.005, and middle vs. late, p = 0.976). Data are indicated as the mean ± s.e.m. (A and D) or the median and quartiles with the maximal and minimal values (B and E). **p < 0.01 and ***p < 0.001. Abbreviation: RS, reward sound.

Next, we examined the firing activity of choice response-related HR and LL type neurons, corresponding to the ending of a behavior in auditory discrimination. In the aDLS, the firing rate of these neurons did not exhibit the transient increase related to the ending of the behavior, instead, there was a modest increase before the lever press (Figure 7–figure supplement 3A and B). In contrast, the choice response-related HR type neurons in the pVLS were transiently activated immediately after the lever press in the four kinds of trial at the middle and late stages, and the rate appeared to gradually decline toward the reward sound presentation (Figure 7D). The averaged firing rate was similar between the HR and LL trials though the stages (Figure 7E). The proportion of the choice response-HR neurons increased at the middle and late stages compared to the early stage (Figure 7F). In addition, the rate of choice response-related LL type neurons displayed similar firing patterns to those observed in the choice response-HR neurons (Figure 7–figure supplement 2B). Thus, subpopulations of pVLS neurons also represent the activity related to the ending of the behavior in auditory discrimination, increasing the neuron number throughout the learning stages in a similar pattern to neurons relevant to the beginning of the behavior in the task.

Structure of neural circuits linked to the aDLS and pVLS

Anatomically, striatal subregions are known to form heterogeneous cortico-basal ganglia loops in mice (Foster et al., 2021; Hintiryan et al., 2016; Hunnicutt et al., 2016). To characterize the structure of neural circuits connected to the aDLS and pVLS in rats, a solution containing cholera toxin subunit B (CTb) labeled with Alexa 488 or Alexa 555 (1.0 mg/mL, 0.3 μL/site) as a bidirectional tracer (Amita et al., 2019) was injected into the aDLS and pVLS of the same animals, respectively (Figure 8A). Brain sections were prepared, and cell bodies and axon terminals were detected by CTb with florescence signals. Sections through the aDLS and pVLS indicated the florescence signals around the injection sites (Figure 8B). Alexa 488 signals were localized in cell bodies in the bilateral, rostrodorsal regions of some cortical areas, including the primary motor cortex, secondary motor cortex, primary somatosensory cortex, and secondary somatosensory cortex, and in the ipsilateral regions of various thalamic nuclei, such as the centrolateral nucleus, centromedian nucleus, parafascicular nucleus, and ventromedial nucleus, whereas Alexa 555 signals were observed in the bilateral, ventrolateral regions of some cortical areas, including the most rostral parts of the primary motor cortex, agranular/dysgranular/granular insular cortices, rhinal cortex, and amygdala nucleus, as well as in the ipsilateral regions of various thalamic nuclei, such as the centromedian nucleus, parafascicular nucleus, and medial geniculate nucleus (Figure 8C and D). In addition, Alexa 488 signals were distributed in axon terminals in the anterior part of the globus pallidus (GP) and entopeduncular nucleus (EPN), whereas Alexa 555 signals were detected in the posterior GP (Figure 8E). In the ventral midbrain. Alexa 488 and 555 signals in the nerve terminals were distributed distinctively in the ventromedial and dorsolateral parts of the substantia nigra pars reticulata (SNr), respectively, along with the rostrocaudal axis (Figure 8F). In the cell bodies, both signals were localized in the centromedial and lateral parts of the substantia nigra pars compacta (SNc), respectively (Figure 8F). The signals in the SNc were colocalized with immunoreactivities for TH, a marker of dopamine neurons in the ventral midbrain (Figure 8G).

Segregation of neural circuits linked to the aDLS and pVLS.

(A) Schematic illustration of the major inputs to and outputs from the striatum. A solution containing CTb labeled with Alexa 488 (green) or Alexa 555 (magenta) was injected into the aDLS and pVLS of the same animals (n = 2). (B) Coordinates for tracer injection into the aDLS and pVLS. Right, representative fluorescence images of sections through the injection sites. (C) Typical images of sections through the cerebral cortex and intralaminar thalamus. (D) Distribution of cell bodies innervating the aDLS and pVLS. Fluorescence signals in the cell bodies were marked by green or magenta dots using ImageJ (version 1.53) and overlaid on gray-scaled images of a brain section. (E) Axonal terminals arising from the aDLS and pVLS to the GP or EPN. Representative images of tracing are shown in the left panels. Camera lucida drawings of labeled signals are shown in the right panels. (F) Labeling of axonal terminals and cell bodies in the ventral midbrain. Representative images are shown in the left panels. Upper images are magnified views of the rectangles in the lower image. Camera lucida drawings are shown in the right panels. Arrowheads in the magnified images indicate labeled cell bodies. (G) Triple-fluorescence histochemistry with sections through the ventral midbrain. The sections were stained with an anti-TH antibody and then with a secondary IgG conjugated with Alexa 647. Representative images of the centromedial (left) and lateral (right) parts of the SNc are shown. Arrowheads indicate TH-positive cells with CTb-labeled signals in each image. The anteroposterior coordinates from bregma (mm) are shown. Scale bars; 1 mm (B-D), 500 μm (E), 100 μm (F and G). Abbreviations: AI, agranular insular cortex; Amy, amygdala nucleus; Au, auditory cortex; CL, centrolateral nucleus; CM, centromedian nucleus; DI, dysgranular insular cortex; GI, granular insular cortex; M1, primary motor cortex; M2, secondary motor cortex; MG, medial geniculate nucleus; PF, parafascicular nucleus; Rhi, rhinal cortex; S1, primary somatosensory cortex; S2, secondary somatosensory cortex; TeA, temporal association cortex; VM, ventromedial nucleus; and VTA, ventral tegmental area.

These results indicate that the aDLS and pVLS receive innervations from different subregions in the cerebral cortex, intralaminar thalamus, and SNc, and send projections to different subregions in the basal ganglia nuclei including the GP, EPN, and SNr, being consistent with previous anatomical studies of cortico-basal ganglia loops (Foster et al., 2021; Hintiryan et al., 2016; Hunnicutt et al., 2016).

Discussion

Our findings demonstrate that the aDLS and pVLS are required for the acquisition of auditory discrimination at different stages of learning, and that the DMS is not necessary for the acquisition of discrimination, although it appears to contribute to the execution of actions, in particular to the regulation of response time. These results not only support some recent studies reporting no requirement of the DMS in the procedural learning (Featherstone and McDonald, 2004; Smith et al., 2021; Wolff et al., 2022), but also reveal important roles of two different subregions in the lateral striatum in the external cue-dependent decision-making. We propose that aDLS and pVLS neurons act to integrate the new learning of auditory discrimination in spatiotemporally and functionally distinct manners, showing a remarkable difference from the prior learning model that proposes the functional dominance changes from the DMS to DLS subregions during procedural learning (Dickinson, 1985; O’Hare et al., 2016; Redgrave et al., 2010; Yin et al., 2004: Yin et al., 2005).

Our 18F-FDG-PET imaging study demonstrated that the brain activity in the aDLS reaches a peak at the middle stage of the acquisition phase of auditory discrimination (Figure 2J). The results of excitotoxic lesion experiment of striatal subregions suggest that the aDLS functions to enhance the acquisition of discrimination through the promotion of behavior based on the stimulus- response association and the suppression of behavior based on the response-outcome association (Figure 3G and J). Transient inhibition experiment of neuronal activity confirmed this aDLS function at the middle stage of the learning, although the drug inhibition tended to reduce the stimulus-response behavior (Figure 4C, E and F). In addition, we found subpopulations of aDLS neurons (reward sound-related neurons) which represent the behavioral outcome and differentiate the correct (HR or LL) and incorrect (HL or LR) choices (Figure 5D and Figure 5–figure supplement 4A). The distinct responses of these neurons may explain the mechanism by which the aDLS regulates the stimulus-response and response-outcome associations through distinct neural circuits leading to a differential control of the two associations. Alternatively, the responses mainly promote the behavior based on the stimulus-response association, thereby secondarily inhibiting the response-outcome association. Indeed, some previous studies have suggested that an action based on either the stimulus-response association or response-outcome association is generated as a result of competition between different striatal subregions (Gremel et al., 2016). We also found other subpopulations of aDLS neurons (first licking-related neurons) showing long-lasting increased activity after obtaining reward in response to a specific combination between the stimulus and response (HR or LL trial) (Figures 6A and Figure 6–figure supplement 1A). This activity may function as the feedback signal to promote the stimulus and response association, resulting in the enhancement of discrimination learning. The proportion of neurons related to the behavioral outcome showed a gradual reduction through the progress of auditory discrimination (Figures 5F and Figure 5–figure supplement 4C, and the extent of the firing activity and the length of the firing period of neurons showing sustained activation after reward were also decreased at the late stage of learning (Figure 6B and C). The learning-dependent changes in the properties of these subpopulations in the aDLS may explain the mechanism underlying the earlier acquisition process of the auditory discrimination.

The 18F-FDG-PET imaging indicated that the pVLS activity was gradually elevated through the acquisition phase of auditory discrimination, showing the maximal value at the late stage (Figure 2K). The lesion experiment suggested that the pVLS drives the formation of discrimination learning via the selective progression of the stimulus-response association (Figure 3H and K), and the transient inhibition experiment ascertained the pVLS function at the late stage of learning (Figure 4D, G and H). Previous studies have shown that the stimulus-response association is encoded in the DLS or VLS as the transient neuronal activity representing the beginning and ending of a learned behavior (Barnes et al., 2005; Chen et al., 2021; Jin and Costa, 2010; Jog et al., 1999). A recent study reported that the VLS is causally related to learned performance based on the auditory stimulus, while the DLS shows a weaker relationship to the behavior (Lee et al., 2020). In the present study, we found subpopulations of pVLS neurons that represent the beginning and ending of a behavior (cue onset- and choice response-related neurons, respectively) (Figure 7A and D, Figure 7–figure supplement 2A and B), showing an increased number of neurons along with the progress of discrimination learning (Figure 7C and F). These data suggest that pVLS neurons contribute to the formation of the stimulus-response association through the learning processes. In addition, we found another subpopulation of pVLS neurons (first licking-related neurons) that are continuously activated after the reward outcome in response to two kinds of correct choices (HR and LL trials) (Figure 6F and Figure 6–figure supplement 2A), and the proportion of neurons was increased at the late stage (Figure 6I and Figure 6–figure supplement 2D). This activity may act as a feedback signal to consolidate the stimulus and response association, leading to the maintenance of associative memory at the late stage during the acquisition phase of learning.

In the present study, we identified neurons with four kinds of event-related activity in the aDLS and pVLS, and classified them into the HR and LL type neurons (Figure 5-7, Figure 5- figure supplement 4 and 5, Figure 6-figure supplement 1 and 2, and Figure 7-figure supplement 1-3). Parts of these two types of neurons with the respective event-related activity overlapped, and the numbers of overlapping neurons varied between striatal subregions. In the pVLS, the overlapping neurons for the same event-related activities were 35.1%, 45.2%, and 67.0% in the cue onset, choice response, and first licking, respectively. In contrast, the aDLS had lower overlaps of 12.6% and 22.7% in the reward sound and first licking, respectively. Thus, distinct neurons in the aDLS tend to have properties of a single trial type while the same neurons in the pVLS seem to possess properties of both types. One possible explanation for the diversity in the proportion of overlapping neurons between the two subregions may be the difference of input patterns into striatal neurons or plastic changes in functional connectivity during learning processes. The detailed mechanism by which aDLS and pVLS neurons react to the events in a different manner should be investigated in the future.

We examined the neural circuits linked to the aDLS and pVLS in rats by using bidirectional tracers. The results showed that these two striatal subregions receive innervations from different subregions in the cerebral cortex, intralaminar thalamus, and ventral midbrain, and provide projections to different subregions in the basal ganglia nuclei (Figure 8), confirming that the two subregions form the heterogenous cortico-basal ganglia loops in the rodent brain (Foster et al., 2021; Hintiryan et al., 2016; Hunnicutt et al., 2016). These data suggest that the aDLS and pVLS mediate the acquisition of auditory discrimination via the parallel loops, and cooperate to achieve neural processing related to the dual regulation of the stimulus-response and response-outcome associations. Further studies are needed to elucidate the mechanism by which the two kinds of loops are processed separately and integrate the new learning of discrimination.

Limitations

In the 18F-FDG PET neuroimaging study, the brain activity in the aDLS reached a peak at the middle stage of the acquisition phase of auditory discrimination (Figure 2J). In the multi-unit electrophysiological recording experiment, the firing activity of aDLS neuron subpopulations related to the behavioral outcome showed no significant differences among the three stages (Figure 5E and Figure 5-figure supplement 4B), while the proportions of these subpopulations gradually reduced through the progress of learning stages (Figure 5F and Figure 5-figure supplement 4C). The extent of firing activity and length of firing period of other subpopulations showing sustained activation after the reward appeared to show learning-dependent decrease (Figure 6B and C, and Figure 6-figure supplement 1B and C), although the proportions of these subpopulations indicated no correlation to the progress of the learning (Figure 6D and Figure 6-figure supplements 1D).

The patterns of the temporal changes in the brain activity of the striatal subregions during the learning stages did not match completely the time variation in the property and proportion of specific event-related neurons. In our electrophysiological analysis, we identified well-isolated neurons from striatal subregions across the behavioral task, focusing on putative MSNs (Figure 5– figure supplement 2). Based on the combinatorial pattern of the tone instruction cue and lever press, we categorized the electrophysiological data into the four trials, including the HR, LL, HL, and LR. We identified HR or LL type neurons showing significant changes in the firing rate related to specific events compared to the baseline firing rate. These neurons were further divided into two groups with increased or decreased activity relative to the baseline firing (Figure 5–figure supplement 3). In the present study, we focused on event-related neurons with the increased activity. Because the analysis was limited to neuronal subpopulations related to specific events with the increased activity, we consider that it is difficult to explain fully dynamic shifts in the brain activity of the striatal subregions dependent on the progress of learning by the time variation of firing activity of individual event-related neurons. The activity of other subpopulations in the striatum may be involved in the shifts in brain activity during the learning processes. In addition, recent studies have reported that the activity of glial cells influences the uptake of 18F-FDG and that these cells regulate spike timing-dependent plasticity (Valtcheva and Venance, 2016). Changes in glial cellular activity, through the control of synaptic plasticity, may partly contribute to the pattern formation of learning-dependent shift of brain activity.

Materials and methods

Animals

Animal care and handling procedures were carried out in accordance with the guidelines established by RIKEN Center for Biosystems Dynamics Research, Fukushima Medical University, and Osaka City University. All procedures were approved by their Institutional Animal Care and Use Committees. All efforts were made to minimize the number of animals used and their suffering throughout the course of the experiments. Male Long-Evans rats (8–13 weeks old, Institute for Animal Reproduction, Ibaraki, Japan) were used for the present study. The rats were kept at 22 ± 2 °C and 60% humidity in a 12-h light/12-h dark cycle, and were continuously available for laboratory chow (CE-2, CLEA Japan, Fukuoka, Japan) and water. Three days before the behavioral training, the rats were subjected to dietary restriction to maintain their body weight at ∼85% of the free feeding weight or water restriction in which they received free access to water for 6 to 9 h each day.

Surgery

Rats underwent a stereotaxic surgery under isoflurane anesthesia (1–5%). For excitotoxic lesion of striatal neurons, a solution containing IBO (533-54971, FUJIFILM Wako Pure Chemical, Osaka, Japan) in PBS (0.8 mg/mL, 0.3 μL per site) was injected bilaterally with the coordinates (mm) from bregma and dura for the aDLS (AP +1.6, ML ±3.3, DV −3.9), pVLS (AP −1.6, ML ±4.6, and DV - 5.6), or DMS (AP −1.6, ML ±1.4, and DV -4.8) according to the atlas of rat brain (Paxinos and Watson, 1998). Injection was performed through a thin-wall borosilicate glass pipette (diameter: 60 μm, B100-75-10, Sutter Instrument, Novato, CA) at a constant velocity of 0.1 mL/min with the microinfusion pump (ESP-32, EICOM, Kyoto, Japan). For transient inactivation of striatal neurons, two guide cannulae (length: 10 mm, outer diameter: 26 gauge, C315G, P1 Technologies, Roanoke, VA) were bilaterally placed on the skull and fixed using four screws and dental cement. The chip of the cannula was placed at 1 mm above the target coordinates. A solution containing MUS (M1523, Sigma-Aldrich, St. Louis, MO) in saline (0.1 μg/μL, 0.25 μL per site) was injected into the target coordinates through the internal cannula (outer diameter: 30 gauge, C315I, P1 Technologies) at a constant velocity of 0.25 μL/min. For the electrophysiological experiments, surgery was performed as described previously (Kitanishi et al., 2021). A 64-channel silicon probe (Buzsaki64spl, NeuroNexus, Ann Arbor, MI) attaching to a 3D-printed microdrive with a movable screw was chronically implanted with the coordinates (mm) from the bregma and dura into the left aDLS (AP +1.6, ML −3.3, and DV −2.5) and pVLS (AP −1.6, ML -4.6, and DV −3.6). A faraday cage composed of copper mesh was secured on the skull with dental cement to reduce electrical noise and avoid damage to the implants. For anterograde and retrograde tracing, a solution containing CTb conjugated to Alexa 488 (1.0 mg/mL, C34775, Invitrogen, Waltham, MA) or CTb conjugated to Alexa 555 (1.0 mg/mL, C34776, Invitrogen) was injected into the aDLS and pVLS (0.3 μL/site), respectively. The coordinates for the injection into striatal subregions were the same as described above. The rats were allowed to recover for 1 week after the surgeries.

Behavioral analysis

The behavioral tasks were conducted in operant conditioning chambers (30.5 × 24.1 × 29.2 cm; ENV-007, MED Associates, Fairfax, VT) equipped with two retractable levers on both sides of a reward port in the front panel (ENV-203M-45, ENV-112CM, and ENV-201A, MED Associates). A multi-tone generator (ENV-223, MED Associates) and room light were mounted on the top of the center and corner panels on the rear side, respectively. For the electrophysiological experiment, a 3D printed reward port was used to detect the licking and to prevent contact with the head attachment and front panel. The behavioral experiment was performed using reinforcement pellets (grain-based precision pellets; F0021, Bio-serv, Flemington, NJ) as a food reward. To prevent spillover of radioactivity into the temporal cortex as a consequence of excessive 18F-FDG accumulation into the muscles by their activity to eat the pellets (Endepols et al., 2010; Marx et al., 2012). the PET imaging experiment was carried out using 0.02% saccharin solution (B0131, Tokyo Chemical Industry, Tokyo, Japan) as a liquid reward. The electrophysiological experiment was also carried out by using the saccharin solution.

The training schedule consisted of four continuous steps. In the first step (magazine training), rats were habituated to the operant conditioning chambers, and the reward was presented with the click every 20 s. Each daily session was started by illumination of the room light, and continued for 30 min. The sessions were conducted for 3 consecutive days. In the second step (shaping), only one lever on the left or right side was inserted. When the rats pressed the lever, they earned a reward accompanied by a click. Each daily session was started by illumination of the light, and lasted 30 min or until the criteria, in which the rats received 90 rewards. When the number of lever presses on either side reached the criteria, the lever was inserted in the opposite side on the next day. When the rats achieved the criteria on both sides for 2 or 3 days, they moved to the following step. In the third step (single lever press task), each trial was started by illumination of the light and insertion of one lever on either the left or right side in a pseudorandom manner with equal frequencies on both sides for every set of four trials. When the rats pressed the lever, the reward was presented with the sound, and the lever was immediately retracted. The room light turned off 4 s after the lever press. When they did not press the lever within 10 s after the start, the trial was terminated when the light turned off. The trial was repeated every 20 s. Each daily session continued for 30 min, and the sessions were conducted for 4 consecutive days. The number of lever presses in each session was counted. In the final step (two-alternative auditory discrimination task), each trial was started by the presentation of a high (10 kHz) or low (2 kHz) frequency tone as an instruction cue in a pseudorandom manner. After 3 s, the room light was illuminated, and two retractable levers were inserted at the same time. The rats were required to press the left or right levers in response to the low or high tone cues, respectively. When the rats pressed the correct lever, the reward was presented with the sound, and the levers were immediately retracted. Correct response allowed the rats to access the reward: the retractable spout delivering water was inserted for 1 s for the PET imaging experiment; a pellet was delivered by a food dispenser for the pharmacological experiment; and three drops of water were delivered through a solenoid valve for the electrophysiological experiment. The room light turned off 4 s after the correct response. On the other hand, error response resulted in the retraction of levers and the turning off of the room light.

No response within 10 s after the lever insertion was recorded as an omission response. The success rate in each session was calculated by dividing the number of correct response trials with the number of lever press trials minus the number of omission trials. The trial was repeated every 20 s. Each daily session continued for 60 min, and the sessions were conducted for 14 or 24 days according to the experiments.

To assess the behavioral strategies in the auditory discrimination task, a trial-by-trial analysis was conducted; the win-shift-win strategy was defined as a correct response, in which, after a correct response during the previous trial, the rats press the opposite lever during the current trial in response to the shift of the tone instruction cue; the lose-shift-lose strategy was defined as an error response, in which, after an error response during the previous trial, the rats press the opposite lever during the current trial despite the shift of the instruction cue. The proportion of both strategies was calculated by dividing the number of each strategy with the number of lever press trials minus the number of omission trials.

For the transient inactivation experiment, MUS injections were bilaterally performed 20 min before start of the session. For habituation, all rats were injected with saline bilaterally into the corresponding striatal subregions on the last day (Day 4) of the single lever press task. During the auditory discrimination task, the first injection was conducted on Day 2 of the task. The second and third injections were conducted the next day, after the success rate had exceeded 60% and 80%, respectively. Each session continued for 15 min.

For the electrophysiological experiments, a delay period (0.5 ± 0.2 s) was added in the auditory discrimination task between the lever press and reward sound to distinguish the motor- and reward-related responses of neurons. When the rats pressed the lever, it was immediately retracted. After the delay period, the reward was presented with a sound. The room light turned off 4 s after the lever press. The trial was repeated every 20 s, and each daily session continued for 60 min.

PET imaging analysis

An 18F-FDG-PET scan was performed with a microPET Focus220 (Siemens Medical Solutions, Knoxville, TN) designed for high-resolution imaging of small animals (spatial resolution of 1.4 mm in full width at half maximum at the center of the field of view) as described previously (Cui et al., 2015). Serial 18F-FDG-PET scans were performed on the last day of the single lever press task and on Days 2, 6, 10, and 24 of auditory discrimination task. To administer 18F-FDG under the awake condition, rats were anesthetized through an indwelling catheter attached to their tail with 5% induction and 2.0–2.5% maintenance of isoflurane (median [interquartile range]: 6 [5–8] min). The recovery period was set to at least 2 h until the beginning of the behavioral session, with the aim of minimizing the impact of anesthesia as much as possible. A solution of 18F-FDG (ca. 74 MBq/0.4 mL) was intravenously injected just before the beginning of each behavioral session through the indwelling catheter. After the 30-min session, the rats were returned to their home cage, and then a 30-min static PET scan was performed under anesthesia with 2.0–2.5% isoflurane 55 min after the 18F-FDG injection. During the PET scan, the body temperature of the anesthetized rats was maintained at 37°C using a small animal warmer (BWT-100A, Bio Research Center, Nagoya, Japan). The acquired emission data were sorted into a single sinogram and reconstructed by a standard 2D filtered back projection (FBP) with a ramp filter and cutoff frequency at 0.5 cycles per pixel or by a statistical maximum a posteriori probability algorithm (MAP), 12 iterations with point spread function effect.

Voxel-based analysis of PET images was conducted according to the method described previously (Cui et al., 2015). Briefly, individual MAP-reconstructed FDG images were aligned to a FDG template image of Long Evans rats using PMOD software package (version 3.2; PMOD Technologies, Fällanden, Switzerland; https://www.pmod.com/web/) and then the aligned images were transformed into the space of a T2 weighted MRI templates for the Long-Evans rat brain (Neuroimaging Tools and Resources Collaboratory; https://www.nitrc.org/projects/tpm_rat/). The transformation parameters obtained from individual MAP-reconstructed FDG images were used to match the individual FBP-reconstructed FDG images with the MRI templates. The voxel size of the template was magnified by 10 to a size similar to that of a human brain. All FBP-reconstructed FDG images were resampled with a voxel size of 1.2 × 1.2 × 1.2 mm and spatially smoothed through an isotropic Gaussian kernel (6-mm FWHM). Voxel-based statistical analysis was performed by using the Statistical Parametric Mapping (SPM) software (version 8; Wellcome Department of Imaging Neuroscience, London, UK; https://www.fil.ion.ucl.ac.uk/spm/). For visualization of the results, SPM t-statistic values were overlayed on MRI templates to define the voxels showing significantly increased or decreased brain activity, using a high threshold (p = 0.001, uncorrected) and an extent threshold (with the output values from the voxel-based analysis). The voxel of interest was determined ass voxels with more than 85% of the maximum T-value within activated areas obtained by voxel-based statistical comparison of the FBP-FDG images. The average regional 18F-FDG uptake was calculated on the basis of values of the voxel of interest in the images.

Electrophysiological recording

Extracellular multi-unit recording was conducted at a sampling rate of 20 kHz from the striatum of freely moving rats by using a 512-ch acquisition board (Open Ephys, Atlanta, GA) via 64-ch recording headstages (C3325, Intan Technologies, Los Angeles, CA). The tips of the probe were lowered by turning the screw (> 140 µm/day) until it reached the target regions (the aDLS and pVLS) during the single lever press task. The tips were sometimes lowered by the same procedure (140 to 280 µm/day) after the daily recording session to improve cell yield. Spike sorting was performed automatically by the Kilosort2 software (Stringer et al., 2015; Marius Pachitariu, Ashburn, VA; https://www.github.com/MouseLand/Kilosort2). Clear noise clusters were removed in the first manual sorting using the phy software (cortex-lab, London, UK; https://github.com/cortex-lab/phy). The remaining clusters were then re-clustered by using the KlustaKwik2 software (Kadir et al., 2014; https://github.com/kwikteam/klustakwik2), and the clusters were adjusted in the second manual sorting by using the phy again. The sorted clusters satisfying all the following criteria were used for further analysis: L-ratio < 0.05, isolation distance > 15, and inter-spike interval (ISI) index < 0.2 (Fee et al., 1996; Schmitzer-Torbert et al., 2005). Well-isolated single neurons (n = 1,485; aDLS, n = 1,062; pVLS, n = 423) were classified as putative MSNs, fast spiking interneurons (FSIs), tonically active neurons (TANs), and unclassified interneurons (UIs). Putative MSNs and TANs were separated from FSIs and UIs by trough-to-peak duration of mean spike waveform (> 0.6 ms) according to previous studies with slight modifications (Bakhurin et al., 2016; Kim et al., 2019). Putative FSIs were then separated from UIs by utilizing the proportion of time associated with long ISIs. This was achieved by summing the ISIs longer than 2 s and subsequently dividing the resulting sum by the total recording time (PropISIs > 2 s) (Peters et al., 2021). Neurons with values of PropISIs > 2 s less than 0.4 were classified as putative FSIs, while neurons with PropISIs > 2 s greater than 0.4 were classified as putative UIs. The putative MSNs and TANs were separated by measuring the post-spike suppression (Peters et al., 2021). We measured the length of suppressed time during which the firing rate of a neuron was suppressed following a spike by counting the number of 1-ms bins in its autocorrelation function until the rate of the neuron was equal to or greater than its averaged rate over the 600-ms to 900-ms autocorrelation bins. Putative TANs were separated by post-spike suppression of less than 40 ms, and the remaining neurons were classified as putative MSNs.

To divide the three stages during the acquisition phase, the sigmoidal function was fitted using the following four-parameter logistic equation: Where α is the inflection point, β is a variable for slope, and λ and γ are parameters related to the lower asymptote and upper asymptote of the learning curve, respectively. The function was used to evaluate the day when the success rate reached 20% (t20%) or 90% (t90%) of the maximum rate change defined by the difference between the γ and λ. Three stages were defined according to the following criteria: early stage (up to 3 days before the middle stage); middle stage (3 days including before and after the day of t20%); and late stage (3 days including the day of t90% and before). One rat learned the discrimination faster, and the success rate reached t90% 2 days after the day of t20%. One day before the day of t90% was assigned to the late stage. For the following analysis, spike data in each striatal subregion were separated into the early, middle, and late stages (n = 295, 364 and 360, respectively, for the aDLS; and n = 87, 110 and 190, respectively, for the pVLS).

To determine the neurons showing significant changes in firing rate related to the task events, electrophysiological data were grouped into four trial types based on the instruction cue (high or low tone) and the choice (right or left lever); HR, high tone and right lever press; LL, low tone and left lever press; HL, high tone and left lever press; and LR, low tone and right lever press. Neurons with three or more trial data were analyzed. We identified the HR or LL type neurons, detecting significant increases or decreases of the firing rate by applying the Wilcoxon rank-sum test (p < 0.01) to trial-by-trial spike number between the baseline period (BL, −1,500 to −500 ms or −1,100 to −500 ms before presentation of the instruction cue) and any of four peri-event periods (cue onset, 0 to +1,000 ms after the onset of the instruction cue; choice response, -700 to +300 ms around the time of lever press; reward sound, −300 to 300 ms around the time of sound onset instructing the reward delivery; and first licking, 0 to +1,000 ms after the time of first licking after the choice response). Any spikes around the events were averaged in 50-ms bins. The event-related activity was shown as averaged firing rate aligned based on the timing of an event for each trial.

The auROC values were calculated by using the spike counts between one trial and either of the other three trials. Bins showing significant differences in the auROC data were determined by applying the Wilcoxon rank-sum test (p < 0.05) to the distribution of 200 ms bin-by-bin auROC values and 0.5.

To compare the number of neurons showing event-related activity among different stages, the number of HR or LL type neurons were calculated by matching the trial number at each stage. Twenty trials were randomly selected for a neuron, and neurons with less than 20 trials were removed from the analysis. The neurons with event-related activity were identified as described above. The proportion (percentage) of neurons relative to the total number of neurons was calculated at each individual stage. The same analysis was repeated 30 times and the distribution of the proportion of neurons with event-related activity at the stage was obtained. The cumulative probability against the proportion was calculated using cumulative distribution function (cdfplot; MATLAB, MathWorks, Natick, MA).

Licking responses were detected with an infrared photobeam sensor placed in the 3D printed reward port, and generated timestamps for the onset of each spout contact according to the analog voltage signal. The licking responses corresponding to four kinds of trial in the first licking- related neurons were shown as averaged rates (50-ms bins) aligned based on the timing of the choice response. To calculate the correlation coefficient between neuronal activity and licking responses, the spikes and licking for 5 s after the first licking were counted every trial for each first licking-related neuron. Trials in which neurons never responded were removed from both the spike and licking data. The correlation coefficients between the two kinds of data were calculated for each neuron (corr; MATLAB, MathWorks), and neurons showing significant differences were determined by applying the Pearson correlation coefficient (p < 0.01).

Histology

After the behavioral experiments, rats were anesthetized with pentobarbital (50 mg/kg body weight) and isoflurane (3%), then transcardially perfused 4% paraformaldehyde in phosphate-buffer (PB).

The brains were post-fixed overnight at 4 ℃ and cryoprotect with 10, 20, and 30% sucrose in PB. Brain tissues were cut into 30-µm sections by using a cryostat (CM3050 S, Leica Biosystems, Wetzlar, Germany). To define the lesioned areas in the striatum, sections were immunostained with a mouse anti-NeuN antibody (1:1,000 dilution; MAB377, Millipore, Burlington, MA), and then with a goat anti-mouse IgG antibody conjugated with Alexa Fluor 488 (1:1,000 dilution; A-11029, Thermo Fisher Scientific, Waltham, MA). The sections were counterstained with 0.3 μM 4’,6- diamidino-2-phenylindole (DAPI; 0.1 μg/mL, D1306, Thermo Fisher Scientific). For the transient inactivation experiment, the positions of guide cannulae were confirmed by staining with 0.1% cresyl violet solution (41021, Muto Pure Chemicals, Tokyo, Japan). For the electrophysiological experiments, small lesions were made by applying an electrolytic current (3 μA, 10 s) through top and bottom channels of each shank by using the stimulus isolator (A365, World Precision Instruments, Sarasota, FL). Sections through the striatum were stained with a mouse anti-TH antibody (1:1,000 dilution; MAB318-AF488, Millipore), and then with a goat anti-mouse IgG antibody conjugated with Alexa Fluor 488 (1:1,000 dilution; Jackson Immunoresearch Laboratories, West Grove, PA). The recording sites were verified by staining with DAPI and red fluorescent Nissl stain solution (1:10,000/1:200 dilution; N21482, Thermo Fisher Scientific). For the tracing, sections through the ventral midbrain were stained with a mouse anti-TH antibody (1:1,000 dilution), and then with a donkey anti-mouse IgG antibody conjugated with Alexa Fluor 647 (1:1,000 dilution; A- 31571, Thermo Fisher Scientific). Fluorescence and immunostained signals were visualized under fluorescence microscopes (BZ-X800, Keyence, Osaka, Japan) and confocal microscope (A1R, Nikon, Osaka, Japan).

Statistical analysis

Statistical analysis was performed using MATLAB (MathWorks) and SPSS Statistics (version 27; IBM, Armonk, NY). Comparisons of data between two groups were performed using the Student’s t-test (paired), the Wilcoxon signed rank test (paired), and the Wilcoxon rank-sum test (unpaired).

Comparisons of data among three groups were constructed using a one-way repeated measures ANOVA, two-way repeated measures ANOVA, or the Kruskal-Wallis test (without assuming normal distribution and homogeneity of variances). All statistical tests were two-sided. The Bonferroni post-hoc test and Mann–Whitney U-test with Bonferroni adjustment were carried out as needed to compare means. For boxplots, the data points below the lower (25th percentile) −1.5 interquartile range and above the upper (75th percentile) quartiles +1.5 interquartile range were regarded as outliers, and they are not included in the boxplots.